---
title: "Analyzing microbiome data"
---

The human microbiome is a crowded ecosystem of bacteria, viruses, fungi, and other microorganisms living in and on the human body. These microbes shape human health and disease, influencing everything from digestion and metabolism to immune function and mood. Working out *which* microbes are present, in *what* proportions, and how those proportions shift between healthy and sick people is one of the most active areas in biology today.

In this chapter you'll take a curated gut-microbiome dataset from a colorectal cancer study, load it into R, and ask the kinds of questions a microbiome researcher asks: How diverse is each sample? Do cancer patients and controls separate when you compare whole communities? Which microbes dominate? You'll do all of this without leaving a single, self-describing data object.

We'll assume the raw sequencing data has already been processed into a tidy abundance table, so you can focus on the analysis rather than the bioinformatics that came before it.

## What you'll learn

By the end of this chapter you will be able to:

- Load a curated microbiome dataset as a `TreeSummarizedExperiment` object.
- Reach into its parts with `assays()`, `colData()`, `rowData()`, and `rowTree()`.
- Subset samples and agglomerate features to a coarser taxonomic rank.
- Compute alpha diversity (within a sample) and beta diversity (between samples).
- Visualize community composition with abundance plots, an ordination, and a heatmap.

This chapter builds directly on the [SummarizedExperiment chapter](bioc-summarizedexperiment.qmd). A `TreeSummarizedExperiment` *is* a `SummarizedExperiment` with a couple of extra parts, so if the three-linked-sheets idea (assays, `rowData`, `colData`) isn't fresh, skim that chapter first.

## Getting started

We'll lean on the `mia` package (short for *microbiome analysis*) throughout this chapter. It provides a suite of tools for loading, transforming, and visualizing microbiome data, all built around the `SummarizedExperiment` and `TreeSummarizedExperiment` classes you'll meet below.

`mia` is a released Bioconductor package, so you install it the usual way with `BiocManager`. If you don't already have `BiocManager` itself, install that first (see the Bioconductor installation chapter), then run:

```{r eval=FALSE}
# Install the mia package (a normal Bioconductor install)
BiocManager::install("mia")
```

::: {.callout-note}
`mia` is a standard, released Bioconductor package. The `BiocManager::install("mia")` call above pulls the stable version that matches your Bioconductor release — the same way you install every other Bioconductor package in this book.
:::

Once `mia` is installed, load it into your R session:

```{r message=FALSE, warning=FALSE}
# Load the mia package
library(mia)
```

Now that the `mia` package is loaded, we can load the microbiome dataset that we will be working with. 

## Accessing microbiome datasets

There are many publicly available microbiome datasets you can use for analysis. Bioconductor provides several packages that bundle curated microbiome datasets, making it easy to load real data into R without first wrangling raw sequencing files.

[curatedMetagenomicData](https://bioconductor.org/packages/curatedMetagenomicsData) is a large collection of curated human microbiome datasets, provided as TreeSE objects (Pasolli et al. 2017). The resource provides curated human microbiome data including gene families, marker abundance, marker presence, pathway abundance, pathway coverage, and relative abundance for samples from different body sites. See the package homepage for more details on data availability and access.

[The microbiomeDataSets package](https://bioconductor.org/packages/microbiomeDataSets) provides a collection of curated microbiome datasets for teaching and research purposes. The package contains several example datasets that can be used to explore different aspects of microbiome analysis, such as alpha and beta diversity, differential abundance analysis, and visualization.

[The mia package](https://bioconductor.org/packages/mia) and several other packages provide example datasets for learning and testing the functions in the package. These datasets are typically small and easy to work with, making them ideal for learning how to analyze microbiome data in R.

## Data containers

Bioconductor provides several specialized data structures for storing and working with microbiome data in R. These data structures are designed to handle the complex and high-dimensional nature of microbiome data and provide a convenient and efficient way to store, manipulate, and analyze microbiome data in R.

[SummarizedExperiment (SE)](https://bioconductor.org/packages/SummarizedExperiment) (Morgan et al. 2020) is a generic, highly optimized container for complex experimental data. It has become a common choice for many kinds of biomedical profiling data, such as RNA-seq, ChIP-seq, microarrays, flow cytometry, proteomics, and single-cell sequencing.

[TreeSummarizedExperiment](https://bioconductor.org/packages/TreeSummarizedExperiment) (Huang 2020) extends that container to carry hierarchical information — phylogenetic trees and sample hierarchies — alongside the usual assays and metadata.

::: {.callout-note}
## A mental model: three linked sheets, plus a tree

In the [SummarizedExperiment chapter](bioc-summarizedexperiment.qmd) you met the *three linked sheets* picture: an **assay** sheet of numbers (features × samples), a **row** sheet (`rowData`) describing each feature, and a **column** sheet (`colData`) describing each sample, all locked together so they can never drift out of sync.

A `TreeSummarizedExperiment` is exactly those three sheets with one more piece pinned to the side: a **phylogenetic tree** (`rowTree`) that records how the features — the microbes — are related to one another by evolution. When you drop a microbe from the object, it drops from the tree too. So everything you already know about reaching into a `SummarizedExperiment` carries over here; the tree is the only genuinely new idea.
:::


## Loading a microbiome dataset

We will be using the `curatedMetagenomicData` package to load the `FengQ_2015` dataset. This dataset contains microbiome data from a study by [Feng et al. (2015)](https://pubmed.ncbi.nlm.nih.gov/25758642/). The curatedMetagenomicData package provides a convenient interface for accessing microbiome datasets that have been pre-processed and curated for analysis. 

::: {.callout-note}
Take a moment to read at least the title and abstract of the paper. We won't reproduce its analysis, but connecting the code in this workflow to the underlying biology and research question will make everything that follows more meaningful.
:::


```{r eval=FALSE}
# Again, before using a package, we must install it
# just once.
BiocManager::install('curatedMetagenomicData')
```

Once installed, we can use the package. 

Ignore the details of the next few lines of code for now, but suffice it to say that what the code will do for us is to load a dataset for downstream analysis.

```{r}
# Load the FengQ_2015 dataset
library(curatedMetagenomicData)
tse <- curatedMetagenomicData("FengQ_2015.relative_abundance", dryrun = FALSE, rownames = "short")[[1]]
```

This code goes out to the internet and downloads the dataset for you. The `tse` object is a `TreeSummarizedExperiment` object that contains the microbiome data from the Feng et al. (2015) study. The `TreeSummarizedExperiment` class is a specialized data structure for storing microbiome data in R, and it is used by the `mia` package for microbiome analysis.

::: {.callout-tip}
If you are working with your own microbiome data, you will often need to  load it into R using functions like `read.csv()` or `read.table()` to read tabular data files. Make sure your data is properly formatted and cleaned before loading it into R for analysis. 
:::

Once the data is loaded, you can start exploring its structure and contents. The `tse` object is a `TreeSummarizedExperiment`. You can think of it as a richer relative of a `data.frame`: it has many specialized slots for storing the parts of a microbiome experiment, laid out in @fig-tse-anatomy. We'll use a handful of accessor functions to look inside it. See the [TreeSummarizedExperiment](https://bioconductor.org/packages/TreeSummarizedExperiment) package for the full details.

In microbiome data science, containers like this one link a taxonomic abundance table to rich side information about both the features (the microbes) and the samples. That abundance data can come from 16S rRNA amplicon sequencing, metagenomic sequencing, phylogenetic microarrays, or other methods. A single experiment often holds several versions of the data at once — raw counts plus transformed or agglomerated versions derived from them — and the container keeps them all together.


![Anatomy of a TreeSummarizedExperiment object. Compared to the SingleCellExperiment objects, TreeSummarizedExperiment has five additional slots. **rowTree**: the hierarchical structure on the rows of the assays. 
**rowLinks**: the link information between rows of the assays and the rowTree.
**colTree**: the hierarchical structure on the columns of the assays.
**colLinks**: the link information between columns of the assays and the colTree.
**referenceSeq** (optional): the reference sequence data per feature (row).
](images/treesummarizedexperiment.png){#fig-tse-anatomy}


To get an overview of the data, we can simply type the name of the object in the R console:

```{r}
# Print the object
tse
```

Read that printout like the label on a box. The `dim` line tells you how many features (rows) and samples (columns) you have; the `assays`, `colData`, `rowData`, and `rowTree` lines preview the parts described above. Notice that the *numbers* themselves are not printed — the dataset is far too large for that — so R shows you the shape and the index of contents instead.

Next, we'll look at some of the key components of the `TreeSummarizedExperiment` object and how to access and manipulate them.

### Experimental measurements: `assays`

When you work with high-throughput biological data, the result of the workflow usually resembles an Excel spreadsheet: samples down the columns, features (genes, microbes, CpG sites, ...) down the rows. The numbers in that grid are called *assay* data, because they come from running an assay — a measurement of the presence and abundance of something. For a microbiome study, an assay tells you how much of each microbe is in each sample.

The original measurements are count-based, but those counts are often transformed — into logarithms, Centered Log-Ratio (CLR) values, or relative abundances — for different analyses. A single object can hold all of these versions at once, so the `assays` slot is a *list* of matrices rather than a single one.

```{r}
assays(tse)
```

An individual assay is reached with the `assay()` function (singular), which returns a single matrix. Recall that to subset a matrix you use `[<rows>, <columns>]`, so here we ask for just the first five rows and columns to keep the output readable:

```{r}
# just the first 5 rows and columns to make the results easier to see
assay(tse, "relative_abundance")[1:5, 1:5]
```

The values are relative abundances — each column sums to 1 across all microbes in that sample. Choose a different number of rows or columns if you like.

### Sample information: `colData`

The `colData` holds everything you know about the *samples*. For a study like this one, that can include the sample ID, the type of sample (fecal, soil, mock, ...), the subject's age and sex, the disease status, and any experimental details you care to track (prep kit lot number, sequencing run, and so on).

```{r}
colData(tse)
```

There is a lot here. Scan the column names: alongside identifiers you'll find biologically interesting variables such as `study_condition` (whether the subject had colorectal cancer or was a control), `age`, and `gender`. These are exactly the variables you'll group and color by when you compare diversity and composition later in the chapter.

::: {.callout-tip}
## Treat `colData` as your sample database
By keeping a reasonably complete `colData`, you can answer questions about your findings without flipping back to an external Excel spreadsheet or a lab notebook. The metadata travels with the data.
:::


### Measured feature information: `rowData`

The `rowData` describes the *features* — here, the microbes. In microbiome work this is where taxonomic information lives: the kingdom, phylum, class, order, family, genus, and species of each detected organism. That taxonomy is what lets you identify which microbes are present and connect community composition to environmental or health factors.

```{r}
head(rowData(tse))
```

Each row is one microbe, and the columns walk down the taxonomic hierarchy from broad (phylum) to specific (species).

### rowTree

A phylogenetic tree records evolutionary relationships, and microbiome analysis leans on it heavily. A `TreeSummarizedExperiment` tracks features and their positions in the tree through two accessors, `rowTree()` and `rowLinks()`.

The tree itself is reached with `rowTree()`, which returns a `phylo` object — R's standard representation of a tree.

::: {.callout-note}
## What is this tree of which we speak?

The "tree" in a `TreeSummarizedExperiment` is a phylogenetic tree: it describes the inferred evolutionary relationships among species (or other entities) based on similarities and differences in their genetic or physical characteristics. The famous "tree of life" is one such tree, showing how all living organisms on Earth are related.

In microbiome analysis, phylogenetic trees represent how the different microbial taxa in a sample are related by descent. This matters because it lets you ask not just *which* microbes are present, but how closely related they are — two samples can share no exact species yet still host very similar branches of the tree.

The `phylo` class is R's standard structure for holding such trees, and the `ggtree` package is the popular tool for drawing them.
:::

```{r}
rowTree(tse)
```

The printed `phylo` object summarizes the tree: how many tips (one per microbe) and internal nodes it has. We can draw it with the `ggtree` package, which renders trees in several layouts (rectangular, circular, radial). Here's the default rectangular view of this study's tree:

```{r message=FALSE}
library(ggtree)
ggtree(rowTree(tse))
```

Don't try to read individual tips — at this scale the value is the overall shape, which shows how the detected microbes fan out across the branches of life. See the [ggtree](https://bioconductor.org/packages/ggtree) package and the [ggtree book](https://yulab-smu.top/treedata-book/) for much more on tree visualization.

The companion accessor `rowLinks()` records *which* tip of the tree each feature corresponds to — the bookkeeping that keeps the assay rows and the tree in step:

```{r}
rowLinks(tse)
```

Both `rowTree` and `rowLinks` are optional parts of a `TreeSummarizedExperiment`, but when present they carry the evolutionary structure that distinguishes this class from a plain `SummarizedExperiment`.

## Wrangling and subsetting

A major advantage of the `SummarizedExperiment` family is that a single object encapsulates everything needed to describe the samples (`colData`), the measured features (`rowData`, `rowTree`), and the measurements themselves (`assays`). When you subset or manipulate the object, every one of those parts is subset to match — automatically, in lockstep.

The `mia` package adds functions for filtering, transforming, and reshaping microbiome data on top of that foundation.

### Subsetting samples

You subset the samples in a `TreeSummarizedExperiment` much as you would the columns of a `data.frame`. Say you want to focus on older subjects. First, look at the `age` column in `colData`:

```{r}
head(colData(tse)$age)
```

Because reaching into `colData` is such a common operation, you can also grab a column directly with `$`, as if the object were a data frame:

```{r}
head(tse$age)
# and get a summary of all patient ages
summary(tse$age)
```

Age is a numeric variable. Let's focus on older patients.

```{r}
# Include all rows, but only columns (samples) who are older than 50
tse_subset_by_age <- tse[, tse$age > 50]
tse_subset_by_age
```

`tse_subset_by_age` is a copy of the original `tse` holding only the 50-and-older subjects. To double-check, look at the ages in the new object — the minimum should now be above 50:

```{r}
summary(tse_subset_by_age$age)
```

### Agglomerating data

The microbiome features (organisms) that are measured may be measured at different taxonomic levels. For example, the data may be available at the species, genus, family, or phylum level. Agglomerating data is the process of summarizing the data at a higher taxonomic level by combining the abundances of the lower-level taxa. This can be useful for simplifying the data and reducing the dimensionality of the data.

The `agglomerateByRank` function in the `mia` package can be used to agglomerate the data at a specified taxonomic rank. For example, to agglomerate the data at the phylum level, you can use the following code:

```{r}
new_tse <- agglomerateByRank(tse, rank = "phylum")
# observe the new object, particularly the number of features (rows)
new_tse
```

Compare the number of features (rows) in `new_tse` to the original `tse`: collapsing hundreds of species into a few dozen phyla shrinks the data dramatically. That makes the data easier to plot and interpret, at the cost of fine-grained detail — a trade-off you'll make often.

## Community indices

Microbiome ecology offers many indices that summarize a sample, or compare a community of samples. Here we give just a brief introduction to two broad flavors: *alpha* diversity (diversity within a single sample) and *beta* diversity (how different two samples are from each other).

### Alpha diversity

Diversity within a sample is captured by several related indices:

- **species richness** — the total number of species present;
- **equitability** — how evenly the species are distributed;
- **diversity** — a combination of the two.

`mia` provides functions for all of these. The `addAlpha()` function calculates one or more indices and stores the results automatically in `colData`, right alongside your sample metadata.

::: {.callout-note}
## Why do we care about alpha diversity?

Alpha diversity is a single number that captures how complex or "diverse" one microbiome sample is. Plenty of things can shift it — antibiotic use, chemotherapy, autoimmune conditions, dietary changes, and more.

When you compare two groups (say, antibiotic users versus non-users), a significant difference in alpha diversity between them can point you toward real biology, or even toward a useful biomarker.
:::

In the code below we calculate the Shannon and observed diversity indices. 

```{r}
tse <- addAlpha(tse, index = "shannon", assay.type = "relative_abundance")
tse <- addAlpha(tse, index = "observed", assay.type = "relative_abundance")
colnames(colData(tse))
```

The `observed` index simply counts the species present in a sample, while the `shannon` index also accounts for how evenly abundant those species are. Notice from the printed column names that both new indices now live in `colData`, ready to plot or model.

Look at the full `colData()` and you'll see the new diversity columns sitting beside the original sample metadata:

```{r}
colData(tse)
```

The `scater` package provides convenient plotting functions designed to work directly with the `SummarizedExperiment` family, so they can read straight from `colData`.

```{r message=FALSE}
#| label: plotAlphaDiversity
library(scater)
plotColData(tse,
    "observed",             # the y-axis value that we just calculated
    "study_condition",      # The x-axis
    colour_by = "gender"    # Color the points by gender
) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = expression(Richness[Observed]))  # and give a meaningful label
```

Each point is one sample, grouped along the x-axis by study condition. Look at where the clouds of points sit: a noticeable shift in observed richness between the cancer and control groups would be the first hint that disease status is associated with how many species live in the gut.

### Beta diversity

Where alpha diversity looks *within* a single sample, beta diversity compares two samples and asks how *different* their communities are. In microbiome analysis the most common measure is the Bray-Curtis dissimilarity between two samples A and B, defined as:

$$
BC_{ij} = \frac{\sum_{k} |A_{k} - B_{k}|}{\sum_{k} (A_{k} + B_{k})}
$$ 

where $A_{k}$ and $B_{k}$ are the abundances of taxon $k$ in samples A and B, respectively. The Bray-Curtis dissimilarity ranges from 0 (identical communities) to 1 (completely different communities).

The `mia` package provides functions for calculating beta diversity indices, such as the Bray-Curtis dissimilarity, Jaccard similarity, and UniFrac distance. These indices can be used to compare the microbial communities between different samples and to identify patterns and relationships between samples based on their microbial composition.

```{r message=FALSE}
#| label: calculateBrayCurtis
#| output: false

# Run PCoA on relabundance assay with Bray-Curtis distances
library(vegan)

tse <- runMDS(tse,
    FUN = vegdist,
    method = "bray",
    assay.type = "relative_abundance",
    name = "MDS_bray"
)
```

This code is a bit to unpack. It calculates the Bray-Curtis dissimilarity between every pair of samples in the `tse` object using the `vegdist` function from the `vegan` package. The resulting dissimilarity matrix is then used to perform a Principal Coordinate Analysis (PCoA) using the `runMDS` function from the `scater` package. The PCoA is a dimensionality reduction technique that projects the high-dimensional Bray-Curtis dissimilarity matrix onto a lower-dimensional space while preserving the pairwise distances between samples.

The resulting `TreeSummarizedExperiment` object now contains the Bray-Curtis dissimilarity matrix in the `MDS_bray` slot, which can be used to visualize the relationships between samples based on their microbial composition.

We can visualize the result with `plotReducedDim()` from the `scater` package, coloring each sample by its study condition:

```{r}
#| label: showPCoA
# Create ggplot object
p <- plotReducedDim(tse, "MDS_bray", colour_by = "study_condition")

print(p)
```

Each point is now a whole microbial community, and points that sit close together have similar compositions. The question to ask of this plot is whether the colors *separate*: if cancer and control samples fell into distinct regions, that would suggest disease status leaves a fingerprint on the community as a whole, not just on a handful of species. Try coloring by other `colData` variables to explore what else structures the data.

## Microbial composition

Let's now look at the microbial composition of the samples in the `tse` object. The microbial composition refers to the relative abundances of different microbial taxa in each sample. This information can provide insights into the diversity and structure of the microbial community in each sample and help identify patterns and relationships between samples based on their microbial composition.

The `mia` package provides functions for visualizing the microbial composition of samples, such as bar plots, heatmaps, and stacked bar plots. These plots can help you explore the relative abundances of different microbial taxa in each sample and identify patterns and relationships between samples based on their microbial composition.

### Abundance

We can start with the 40 most abundant taxa. This plot shows, for each of those taxa, the spread of its relative abundance across all samples (one jittered point per sample), sorted by abundance.

```{r}
#| label: plotTopTaxa
library(miaViz)
plotAbundanceDensity(tse,
    layout = "jitter",
    assay.type = "relative_abundance",
    n = 40
) +
    scale_x_log10(label = scales::percent)
```

A density layout is handy for zooming in on a few taxa and splitting them by group. Here we take the top five and color by study condition — separated curves would suggest that taxon's abundance differs between cancer and control samples.

```{r}
plotAbundanceDensity(tse,
    layout = "density",
    assay.type = "relative_abundance",
    n = 5, colour_by = "study_condition"
) +
    scale_x_log10()
```


### Prevalence

*Abundance* asks how *much* of a taxon is present; *prevalence* asks in how *many* samples it appears at all. A microbe can be abundant in a few people yet rare across the cohort, or modestly abundant but present in nearly everyone — prevalence tells these apart.

Let's agglomerate to the genus level and list the most prevalent genera, counting a genus as "present" wherever its relative abundance exceeds 1%:


```{r}
tse_genus <- agglomerateByRank(tse, rank = "genus")
```

```{r}
head(getPrevalence(tse_genus,
    detection = 1 / 100,
    sort = TRUE, assay.type = "relative_abundance",
    as.relative = TRUE
))
```

Each value is the fraction of samples in which that genus clears the 1% threshold. A value near 1 marks a near-universal member of the gut community; a small value marks a genus that is only occasionally abundant.

## Heatmaps

A heatmap lets you see abundance across every taxon and every sample at once. To keep it readable, we'll agglomerate to the phylum level first — a handful of phyla, rather than hundreds of species — and then draw the grid, annotating each sample column with its study condition.

```{r message=FALSE}
library(pheatmap)
tse_phylum <- agglomerateByRank(tse, rank = "phylum")
pheatmap(
    assay(tse_phylum, "relative_abundance"),
    annotation_col = as.data.frame(colData(tse_phylum)[, "study_condition", drop = FALSE])
)
```

Read the heatmap as a grid: each row is a phylum, each column a sample, and the color is that phylum's relative abundance. The colored bar across the top marks each sample's study condition. The clustering reorders rows and columns so that similar profiles sit together — if columns of the same condition tend to cluster, you're seeing the same community-level signal the PCoA hinted at, now broken down phylum by phylum.

## Summary

In this chapter you took a real curated dataset and ran a complete first-pass microbiome analysis without ever splitting your data across loose variables. You loaded the `FengQ_2015` study as a `TreeSummarizedExperiment` — a `SummarizedExperiment` with a phylogenetic tree attached — and reached into its parts with `assays()`, `colData()`, `rowData()`, and `rowTree()`. You subset samples by age and agglomerated features up to coarser taxonomic ranks, trusting the object to keep every part aligned. Then you summarized the communities: alpha diversity within each sample, beta diversity (Bray-Curtis) between samples visualized as a PCoA, and composition shown as abundance plots and a heatmap. Each of these is a doorway to a real research question about how the gut community differs between health and disease.

## Exercises

These exercises reuse the `tse` object built earlier in the chapter. If you've started a fresh session, re-run the loading chunk first.

1. How many samples in the dataset come from subjects older than 60? (Hint: subset on `tse$age` as you did for the 50+ subjects, then check the dimensions.)
2. Agglomerate `tse` to the `family` rank and report how many features remain. How does that compare to the phylum-level count?
3. Add the `shannon` alpha-diversity index to `tse` and make a `plotColData()` plot of it against `study_condition`. Does the picture match what you saw for observed richness?

::: {.callout-note collapse="true"}
## Solutions

```{r}
# 1. subset the samples (columns) and read off the new dimensions
tse_over60 <- tse[, tse$age > 60]
ncol(tse_over60)

# 2. agglomerate to family level and count features (rows)
tse_family <- agglomerateByRank(tse, rank = "family")
nrow(tse_family)   # fewer than at species level, more than at phylum level

# 3. add Shannon diversity, then plot it
tse <- addAlpha(tse, index = "shannon", assay.type = "relative_abundance")
plotColData(tse, "shannon", "study_condition", colour_by = "gender") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The key idea in exercises 1 and 2: you subset or agglomerate the *whole object*, and every linked part — assays, `colData`, `rowData`, the tree — follows along automatically. You never have to remember to update them separately.
:::

## Resources

The `mia` developers maintain [Orchestrating Microbiome Analysis with Bioconductor (OMA)](https://microbiome.github.io/OMA/docs/devel/), an online book that goes well beyond this introduction — data loading, preprocessing, differential abundance, and a wide range of visualizations, all worked through with detailed examples. It's the natural next step once you're comfortable with the basics here.

For the phylogenetic-tree side of things, the [ggtree](https://bioconductor.org/packages/ggtree) package and its companion [ggtree book](https://yulab-smu.top/treedata-book/) cover tree visualization in depth.
