---
title: "Building and sharing a SummarizedExperiment"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

In the [previous chapter](bioc-summarizedexperiment.qmd) you worked with a
`SummarizedExperiment` someone else had assembled — the `airway` dataset, ready to
subset and plot. But a great deal of real data does *not* arrive that way. It
arrives as a spreadsheet: a few columns describing the genes, a wall of numeric
columns holding the measurements, and the sample information living only in your
head or a separate file. Turning that loose table into a proper
`SummarizedExperiment` is one of the most useful things you can do, because from
that point on the object keeps itself consistent — and it speaks the same language
as every Bioconductor analysis package you'll reach for next.

In this chapter you'll build a `SummarizedExperiment` from raw parts, add a derived
assay, draw a clustered heatmap, and then save and share the result so a
collaborator gets everything in a single file.

## What you'll learn

By the end of this chapter you will be able to:

- Assemble a `SummarizedExperiment` from a matrix (or several), a feature table,
  and a sample table — and explain the alignment rules the constructor enforces.
- Add a **derived assay** (e.g. a normalized or log-ratio matrix) that stays
  aligned to the originals.
- Select informative (variable) features and draw an unsupervised clustering
  heatmap.
- **Save** an experiment to one file and **share** it, provenance and all, with
  `saveRDS()`/`readRDS()`.
- Explain why this one object plugs into the wider Bioconductor ecosystem.

::: {.callout-tip}
## Installing the packages

```{r eval=FALSE}
BiocManager::install(c("SummarizedExperiment", "pheatmap"))
```
:::

```{r setup, echo=FALSE}
suppressPackageStartupMessages(library(SummarizedExperiment))
```

## The ingredients

To construct a `SummarizedExperiment` you supply:

- `assays`: one or more matrices (or matrix-like objects) of data — the numbers;
- `rowData`: a `DataFrame` describing the features (rows);
- `colData`: a `DataFrame` describing the samples (columns);
- optionally `metadata`: a list of anything else about the experiment.

We'll build one from the **DeRisi** dataset — a classic yeast microarray time
course measuring gene expression as cells exhaust their glucose and switch
metabolic gears (the *diauxic shift*). It begins life as a plain `data.frame`: the
first two columns identify each gene, and the rest are measurements.

```{r load}
library(SummarizedExperiment)
deRisi <- read.csv("https://raw.githubusercontent.com/seandavi/RBiocBook/main/data/derisi.csv")
head(deRisi)
```

We'll pull the three pieces out of this table one at a time.

### rowData — feature information

The first two columns describe each gene, so they become the `rowData`:

```{r rowdata}
rdata <- deRisi[, 1:2]
head(rdata)
```

### colData — sample information

The sample information isn't in the file, so we supply it ourselves. Each of the
seven columns of measurements is a timepoint in the experiment:

```{r coldata}
cdata <- DataFrame(sample = paste("Sample", 0:6), timepoint = 0:6,
                   hours = c(0, 9.5, 11.5, 13.5, 15.5, 18.5, 20.5))
cdata
```

### assays — the data

The DeRisi experiment recorded four *different* measurements for every gene at
every timepoint:

| assay | description |
|-------|-------------|
| `R`  | Red foreground fluorescence |
| `G`  | Green foreground fluorescence |
| `Rb` | Red background fluorescence |
| `Gb` | Green background fluorescence |

Each becomes its own matrix, sliced out of the numeric columns:

```{r matrices}
R  <- as.matrix(deRisi[, 3:9])
G  <- as.matrix(deRisi[, 10:16])
Rb <- as.matrix(deRisi[, 17:23])
Gb <- as.matrix(deRisi[, 24:30])
```

### Aligning the names

Here is the rule that makes the whole container trustworthy. The constructor
**checks that the pieces line up**: the row names of every assay matrix must match
the row names of `rowData`, and the column names of every assay matrix must match
the row names of `colData`. If they don't, you get an error rather than a silent
mismatch — the constructor refuses to build something inconsistent.

So we set those names explicitly. Genes (rows) are keyed by their ORF identifier;
samples (columns) by their sample name:

```{r names}
rownames(rdata) <- rdata$ORF
rownames(cdata) <- cdata$sample
for (m in c("R", "G", "Rb", "Gb")) {
    mat <- get(m)
    rownames(mat) <- rdata$ORF
    colnames(mat) <- cdata$sample
    assign(m, mat)
}
```

### Putting it together

```{r build}
se <- SummarizedExperiment(assays = list(R = R, G = G, Rb = Rb, Gb = Gb),
                           rowData = rdata, colData = cdata)
se
```

One object now holds four aligned assays, the gene table, and the sample table.
The constructor accepted it, which means everything lines up.

::: {.callout-warning}
## When the constructor complains
If you ever see an error about non-matching dimensions or names when building a
`SummarizedExperiment`, that is the container doing its job. Check that each
matrix is genes × samples (not transposed), and that its `rownames`/`colnames`
match `rowData`/`colData`. The error is annoying once; the silent mismatch it
prevents is annoying for the life of the paper.
:::

## Adding a derived assay

Real analysis rarely stops at the raw numbers. A natural quantity here is the
background-corrected **log ratio** of red to green — a standard summary of
two-color microarray signal. Because we compute it from assays that are already
aligned, the result drops straight back into the object as a fifth assay, aligned
forever:

```{r logratios}
logRatios <- log2((assay(se, "R") - assay(se, "Rb")) /
                  (assay(se, "G") - assay(se, "Gb")))
assays(se)$logRatios <- logRatios
assayNames(se)
```

That is a real benefit worth pausing on: you can keep raw and processed
measurements *in the same object*, each guaranteed to refer to the same genes and
samples. Reach for any of them by name:

```{r logratios-peek}
head(assay(se, "logRatios"))
```

```{r fig-derisi-hist, fig.cap="Distribution of background-corrected log ratios across the DeRisi time course."}
hist(assay(se, "logRatios"), breaks = 50,
     main = "DeRisi log ratios", xlab = "log2(R/G)")
```

## A clustering heatmap of informative genes

With 6,400 genes, no heatmap can show them all — nor would you want it to. As in
the previous chapter, we keep the **variable** genes, because a gene whose log
ratio barely moves across the time course carries no information about the
diauxic shift; the genes that rise and fall are the informative ones. We rank by
standard deviation and keep the top 50.

The log-ratio calculation produces some non-finite values (a background-corrected
signal can go negative, and `log2` of a negative number is `NaN`), so we first
keep only genes measured cleanly at every timepoint:

```{r topvar}
lr <- assay(se, "logRatios")
lr <- lr[rowSums(!is.finite(lr)) == 0, ]    # genes finite at all 7 timepoints
top <- head(order(apply(lr, 1, sd), decreasing = TRUE), 50)
lr_top <- lr[top, ]
dim(lr_top)
```

Now the heatmap. Two deliberate choices: we **don't** cluster the columns
(`cluster_cols = FALSE`) so the samples stay in time order and we can read each
gene's trajectory left to right; and we **do** cluster the rows, which groups
genes that move together. The column annotation — the time in hours — comes
straight from `colData`.

```{r fig-derisi-heatmap, fig.cap="The 50 most variable genes across the DeRisi diauxic-shift time course. Columns are held in time order; rows (genes) are clustered, grouping co-regulated genes. The annotation bar (hours) comes directly from colData."}
library(pheatmap)
ann <- as.data.frame(colData(se)[, "hours", drop = FALSE])
pheatmap(lr_top,
         cluster_cols = FALSE,          # keep samples in time order
         cluster_rows = TRUE,           # group co-regulated genes
         annotation_col = ann,
         show_rownames = FALSE,
         scale = "row",
         main = "Top 50 variable genes (DeRisi)")
```

::: {.callout-note}
## Unsupervised clustering, again — now on the genes
Clustering the **rows** here groups genes purely by how similarly their expression
moves over time, with no labels supplied — the same *unsupervised* idea you met in
the [previous chapter](bioc-summarizedexperiment.qmd), now pointed at features
instead of samples. Blocks of genes that rise together (or fall together) across
the diauxic shift show up as bands. That is exactly the structure the
[k-means chapter](kmeans.qmd) sets out to find formally on this very dataset; the
heatmap is the quick, visual version of the same question.
:::

## Saving and sharing

Here is the part that makes a collaborator's day. Your experiment — four raw
assays, a derived one, the gene table, the sample table, and any notes — is a
single R object. Save it to a single file:

```{r save}
# In practice you'd give it a real name, e.g. "derisi_experiment.rds".
path <- tempfile(fileext = ".rds")
saveRDS(se, path)
```

Anyone who receives that file reads it back and has *everything*, already aligned,
with nothing to re-link:

```{r load-back}
se2 <- readRDS(path)
se2
```

No "which spreadsheet had the sample labels?", no "are the columns in the same
order?". The object is self-describing. Stash provenance in `metadata()` and it
rides along too:

```{r provenance}
metadata(se)$processing <- "background-corrected log2 ratios; built in the RBioc Book"
metadata(se)$date <- "2026-06-06"
saveRDS(se, path)
metadata(readRDS(path))            # the notes survive the round trip
```

This is also how published datasets are distributed: many arrive as ready-made
`SummarizedExperiment` objects through Bioconductor's `ExperimentHub` and data
packages (the `airway` package from the last chapter is exactly this). You load
one line and get an analysis-ready object — because the author saved it in this
shape.

## Why this is worth it: the ecosystem payoff

Step back and ask what you actually *gained* by building this object instead of
keeping a few matrices around.

- **It can't silently desync.** Every subset, every new assay, stays aligned.
- **It carries its own context.** Feature annotation, sample annotation, and
  experiment notes travel with the numbers, in one file.
- **It is the common currency of Bioconductor.** This is the big one. Hundreds of
  analysis packages — `DESeq2` and `edgeR` and `limma` for differential
  expression, `scran` and `scater` for single cell, and many more — *take a
  `SummarizedExperiment` and hand one back*. You don't reshape your data for each
  tool; you build the object once and plug it into all of them.
- **It is the root of a family.** The `SingleCellExperiment` you'll meet in the
  [single-cell chapter](single_cell/setup.qmd) and the `TreeSummarizedExperiment`
  in the [microbiome chapter](310_microbiome.qmd) are both `SummarizedExperiment`s
  with extra parts bolted on. Learn this shape once and those feel familiar
  immediately. The [GEOquery chapter](geoquery.qmd) and the
  [k-means chapter](kmeans.qmd) both hand you data in exactly this form.

Build it once; everything downstream speaks the same language.

## Summary

You took a flat spreadsheet and assembled it into a `SummarizedExperiment` —
extracting `rowData`, supplying `colData`, building four assay matrices, and
letting the constructor enforce that they all line up. You added a derived
log-ratio assay that stays aligned, drew an unsupervised clustering heatmap of the
most informative genes, and saved the whole thing to one self-describing file that
carries its provenance. Most importantly, you now hold an object that the rest of
the Bioconductor ecosystem accepts directly.

## Exercises

```{r ex-fresh, echo=FALSE}
suppressPackageStartupMessages(library(SummarizedExperiment))
```

1. **Build a tiny one.** Create a `SummarizedExperiment` from a 4 × 3 matrix of
   your choosing, a `rowData` with a `gene` column, and a `colData` with a
   `group` column. Confirm its `dim()`.
2. **Break it on purpose.** Try to build a `SummarizedExperiment` where the assay
   has 3 columns but `colData` has 4 rows. What happens, and why is that a good
   thing?
3. **Add a derived assay.** To the DeRisi `se`, add an assay called `Rcorr` equal
   to `R - Rb` (red minus red background). Confirm it appears in `assayNames(se)`.
4. **Save and reload.** Save the DeRisi `se` to a temporary file, read it back,
   and confirm the assay names and `metadata()` survived.

::: {.callout-note collapse="true"}
## Solutions

```{r}
# 1. a minimal SummarizedExperiment
m <- matrix(1:12, nrow = 4,
            dimnames = list(paste0("g", 1:4), paste0("s", 1:3)))
rd <- DataFrame(gene = paste0("g", 1:4), row.names = paste0("g", 1:4))
cd <- DataFrame(group = c("a", "a", "b"), row.names = paste0("s", 1:3))
tiny <- SummarizedExperiment(assays = list(counts = m), rowData = rd, colData = cd)
dim(tiny)

# 2. mismatched dimensions -> the constructor errors (a feature, not a bug):
#    it refuses to build an object whose pieces don't line up, so you can never
#    end up with sample labels pointing at the wrong columns.
try(SummarizedExperiment(assays = list(x = m[, 1:3]),
                         colData = DataFrame(group = letters[1:4])))

# 3. a derived assay, aligned automatically
assays(se)$Rcorr <- assay(se, "R") - assay(se, "Rb")
assayNames(se)

# 4. round-trip through a file
f <- tempfile(fileext = ".rds")
saveRDS(se, f)
back <- readRDS(f)
assayNames(back)
metadata(back)
```
:::
