---
title: "Base R versus the tidyverse"
author:
  - "Martin Morgan"
  - "Lori Kern"
---

There is almost never just one way to do something in R. The same task — count a
variable, summarize a column, draw a boxplot — can be written in plain **base R**
or with the **[tidyverse](https://www.tidyverse.org/)**, a family of packages
(`dplyr`, `readr`, `ggplot2`, and friends) built around a shared, readable style.
Neither is "right." They are two dialects, and good R users read and write both.

The best way to feel the difference is to watch the *same* analysis done both
ways, side by side. So in this chapter we revisit the BRFSS dataset from the
exploratory-data-analysis chapter and redo it twice: once with base R's
data frames and graphics, and once with tidyverse tibbles and `ggplot2`. Read
across each pair of columns and you'll start to recognize which dialect you
reach for, and when.

## What you'll learn

- Read the same CSV into a base R `data.frame` and a tidyverse `tibble`, and
  describe how they differ.
- Tabulate and summarize a variable with base R (`table()`, `aggregate()`) and
  with `dplyr` (`count()`, `summarize()`, `group_by()`).
- Reshape a count table with `tidyr::pivot_wider()`.
- Recreate the same plots — boxplot, density, scatter, regression line,
  histogram — in base graphics and in `ggplot2`.
- Read both dialects fluently, and choose the one that fits the task in front of
  you.

## Loading the dataset

First we load the dataset two ways. Base R reads a CSV into a classic
`data.frame` with `read.csv()`; the tidyverse's `readr` package reads the same
file into a **tibble** with `read_csv()`. As in the EDA chapter, point R at the
`BRFSS-subset.csv` file in your working directory — interactively, `file.choose()`
opens a dialog so you don't have to type the path:

```{r setup, include=FALSE}
# Fix the path so the chapter renders non-interactively; readers use file.choose().
path <- "BRFSS-subset.csv"
```

```{r ALL-choose-A3, eval=FALSE}
path <- file.choose()    # opens a dialog; navigate to BRFSS-subset.csv

# We'll use dplyr throughout, so load it now
library(dplyr)
```

```{r, include=FALSE}
library(dplyr)
```

::: {.columns}

::: {.column width="50%"}

```{r ALL-input-A3}
# loading using base R
stopifnot(file.exists(path))
brfss_DF <- read.csv(path)
```
:::
::: {.column width="50%"}

```{r, message = FALSE}
# loading using readr
library(readr)
brfss_tbl <- readr::read_csv(path)
```
:::
:::

Let's examine our objects:

::: {.columns}

::: {.column width="50%"}
```{r}
# Classic data frame
head(brfss_DF)
```
:::
::: {.column width="50%"}

```{r}
# Tidyverse tibble
head(brfss_tbl)
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r}
# Classic data frame
class(brfss_DF)
```
:::
::: {.column width="50%"}

```{r}
# Tidyverse tibble
class(brfss_tbl)
```
:::
:::

::: {.callout-note}
**Note:** A tidyverse tibble object inherits a data.frame class. This means that
most data.frame operations like `dim()`, `colnames()`, `$`, `[`, etc. will work on the tibble
object as well. 
:::

## Clean data

Both 'Sex' and 'Year' are really `factor` values (each can only take on specific
levels, 'Female' and 'Male' for 'Sex', and '1990' and '2010' for 'Year').



::: {.columns}

::: {.column width="50%"}
```{r}
# base R / data.frame
brfss_DF$Year <- factor(brfss_DF$Year)
brfss_DF$Sex <- factor(brfss_DF$Sex)
```
:::
::: {.column width="50%"}

```{r}
# dplyr / tibble 
brfss_tbl <-
brfss_tbl |>
    mutate(
        Sex = factor(Sex,
	      levels = c("Female", "Male")),
        Year = factor(Year,
	      levels = c("1990", "2010"))
    )
```
:::
:::

## Data Exploration

Let's do some basic exploration. `summary()` works the same on both objects, but
let's look at some summary tables and counts instead. The two dialects produce
the same results in slightly different formats.

We'll start with basic table of a single variable:

::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R / data.frame
table(brfss_DF$Year)
```
:::
::: {.column width="50%"}

```{r, eval=FALSE}
# dplyr / tibble
brfss_tbl |> count(Year)
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R / data.frame
table(brfss_DF$Year)
```
:::
::: {.column width="50%"}

```{r, echo=FALSE}
# dplyr / tibble
brfss_tbl |> count(Year)
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R / data.frame
table(brfss_DF$Sex)
```
:::
::: {.column width="50%"}

```{r, eval=FALSE}
# dplyr / tibble
brfss_tbl |> count(Sex)
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R / data.frame
table(brfss_DF$Sex)
```
:::
::: {.column width="50%"}

```{r, echo=FALSE}
# dplyr / tibble
brfss_tbl |> count(Sex)
```
:::
:::


Now let's look at contingency table

::: {.columns}

::: {.column width="50%"}
```{r}
# base R / data.frame
table(brfss_DF$Sex, brfss_DF$Year)
```
:::
::: {.column width="50%"}

```{r}
# dplyr / tibble
brfss_tbl |> count(Sex, Year)
```
:::
:::

We can get the tidy table to look even more similar to the base R table with the
help of the tidyr package's function `pivot_wider`


::: {.columns}

::: {.column width="50%"}
```{r}
# base R / data.frame
table(brfss_DF$Sex, brfss_DF$Year)
```
:::
::: {.column width="50%"}

```{r}
# dplyr / tibble
library(tidyr)
brfss_tbl |> count(Sex, Year) |>
    tidyr::pivot_wider(names_from = "Year", values_from = "n")

```
:::
:::

What about some summary statistics on the columns of data? `summarize()` will
create the new data.frame automatically; base R you have to create your own.


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R / data.frame
data.frame(
  avg_age = mean(brfss_DF$Age, na.rm = TRUE),
  ave_wt  = mean(brfss_DF$Weight, na.rm = TRUE),
  ave_ht  = mean(brfss_DF$Height, na.rm = TRUE)
)
```
:::
::: {.column width="50%"}

```{r, eval=FALSE}
# dplyr / tibble
brfss_tbl |>
    summarize(
        avg_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
	)
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R / data.frame
data.frame(
  avg_age = mean(brfss_DF$Age, na.rm = TRUE),
  ave_wt  = mean(brfss_DF$Weight, na.rm = TRUE),
  ave_ht  = mean(brfss_DF$Height, na.rm = TRUE)
)
```
:::
::: {.column width="50%"}

```{r, echo=FALSE}
# dplyr / tibble
brfss_tbl |>
    summarize(
        avg_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
	)
```
:::
:::

If we want to get more complex with groupings by Year and Sex, `dplyr` uses
`group_by()` where base R would use `aggregate()`.

::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R / data.frame
aggregate(
  cbind(Age, Weight, Height) ~ Sex + Year,
  data = brfss_DF,
  FUN = function(x) mean(x, na.rm = TRUE)
)
```
:::
::: {.column width="50%"}

```{r, eval=FALSE}
# dplyr / tibble
brfss_tbl |>
    group_by(Sex, Year) |>
    summarize(
        avg_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
    )
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R / data.frame
aggregate(
  cbind(Age, Weight, Height) ~ Sex + Year,
  data = brfss_DF,
  FUN = function(x) mean(x, na.rm = TRUE)
)
```
:::
::: {.column width="50%"}

```{r, echo=FALSE, message=FALSE}
# dplyr / tibble
brfss_tbl |>
    group_by(Sex, Year) |>
    summarize(
        avg_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
    )
```
:::
:::


## Visualization

Before we start visualizing, lets create a few different subsets of data.


::: {.columns}

::: {.column width="50%"}
```{r}
# base R / data.frame
brfss_female_DF <-
    brfss_DF[brfss_DF$Sex == "Female",]
brfss_male_DF <-
    brfss_DF[brfss_DF$Sex == "Male",]
brfss_2010_DF <-
    brfss_DF[brfss_DF$Year == "2010",]
```
:::
::: {.column width="50%"}

```{r}
# dplyr / tibble
brfss_male_tbl <-
    brfss_tbl |> filter(Sex == "Male")
brfss_female_tbl <-
    brfss_tbl |> filter(Sex == "Female")
brfss_2010_tbl <-
    brfss_tbl |> filter(Year == "2010")
```
:::
:::

We should also load the ggplot2 package so we can compare base R graphics vs
ggplot2

```{r}
library(ggplot2)
```

Let's start with a boxplot that compares the Weights of Males vs Females for the
2010 dataset.

::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
plot(Weight ~ Sex, brfss_2010_DF)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, eval=FALSE}
# ggplot2
ggplot(brfss_2010_tbl) +
    aes(x = Sex, y = Weight) +
    geom_boxplot()
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
plot(Weight ~ Sex, brfss_2010_DF)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, echo=FALSE}
# ggplot2
ggplot(brfss_2010_tbl) +
    aes(x = Sex, y = Weight) +
    geom_boxplot()
```
:::
:::


Let's look at some density and scatterplots.


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
den_male <- density(brfss_2010_DF$Weight[brfss_2010_DF$Sex == "Male"], na.rm = TRUE)
den_female <- density(brfss_2010_DF$Weight[brfss_2010_DF$Sex == "Female"], na.rm = TRUE)
plot(den_male, 
     col = "skyblue", lwd = 2,
     main = "Density of Weight by Sex",
     xlab = "Weight")

lines(den_female, 
      col = "lightsalmon", lwd = 2)

legend("topright",
    legend = c("Male", "Female"),
    col = c("skyblue", "lightsalmon"), lwd = 2)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Weight, color= Sex) +
    geom_density()
```
:::
:::



::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
den_male <- density(brfss_2010_DF$Weight[brfss_2010_DF$Sex == "Male"], na.rm = TRUE)
den_female <- density(brfss_2010_DF$Weight[brfss_2010_DF$Sex == "Female"], na.rm = TRUE)
plot(den_male, 
     col = "skyblue", lwd = 2,
     main = "Density of Weight by Sex",
     xlab = "Weight")

lines(den_female, 
      col = "lightsalmon", lwd = 2)

legend("topright",
    legend = c("Male", "Female"),
    col = c("skyblue", "lightsalmon"), lwd = 2)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Weight, color= Sex) +
    geom_density()
```
:::
:::

Presumably taller people are heavier than shorter people. Let's examine this
relationship.


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
plot(Weight ~ Height, brfss_2010_DF)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight) +
    geom_point()
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
plot(Weight ~ Height, brfss_2010_DF)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight) +
    geom_point()
```
:::
:::

Let's fit the linear regression


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
plot(Weight ~ Height, brfss_2010_DF)
fit <- lm(Weight ~ Height, brfss_2010_DF)
abline(fit, col="blue", lwd=2)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight) +
    geom_point() +
    geom_smooth(method = "lm")
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
plot(Weight ~ Height, brfss_2010_DF)
fit <- lm(Weight ~ Height, brfss_2010_DF)
abline(fit, col="blue", lwd=2)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight) +
    geom_point() +
    geom_smooth(method = "lm")
```
:::
:::

We saw that there could be a difference based on Sex. Let's add color to the
points

::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
colors <- c("Female" = "lightsalmon", "Male" = "skyblue")
plot(Weight ~ Height, brfss_2010_DF,
  col = colors[Sex], pch = 16)
for (sex in levels(brfss_2010_DF$Sex)) {
  subset_data <- subset(brfss_2010_DF, Sex == sex)
  fit <- lm(Weight ~ Height, data = subset_data)
  abline(fit, col = colors[sex], lwd = 2)
}
legend("topleft", legend = levels(brfss_2010_DF$Sex), 
       col = colors, pch = 16, bty = "n")
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight, color = Sex) +
    geom_point() +
    geom_smooth(method = "lm")
```
:::
:::

::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
colors <- c("Female" = "lightsalmon","Male" = "skyblue")
plot(Weight ~ Height, brfss_2010_DF,
  col = colors[Sex], pch = 16)
for (sex in levels(brfss_2010_DF$Sex)) {
  subset_data <- subset(brfss_2010_DF, Sex == sex)
  fit <- lm(Weight ~ Height, data = subset_data)
  abline(fit, col = colors[sex], lwd = 2)
}
legend("topleft", legend = levels(brfss_2010_DF$Sex), 
       col = colors, pch = 16, bty = "n")
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
brfss_2010_tbl |>
    ggplot() +
    aes(x = Height, y = Weight, color = Sex) +
    geom_point() +
    geom_smooth(method = "lm")
```
:::
:::

Let's look at a histogram of `Weight` for the 2010 males. We didn't make that
subset earlier, so each dialect carves it out on the fly — base R with
`subset()`, the tidyverse with `filter()`.


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
brfss_2010_Male <- subset(brfss_DF,
    Year == 2010 & Sex == "Male")
hist(brfss_2010_Male$Weight)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
brfss_2010_tbl |> filter(Sex == "Male") |>
    ggplot() +
    aes(x = Weight) +
    geom_histogram(col = "white")
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
brfss_2010_Male <- subset(brfss_DF,
    Year == 2010 & Sex == "Male")
hist(brfss_2010_Male$Weight)
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
brfss_2010_tbl |> filter(Sex == "Male") |>
    ggplot() +
    aes(x = Weight) +
    geom_histogram(col = "white")
```
:::
:::

What if we took all the Males and looked to see if the relationship of Height
and Weight changed between 1990 and 2010.


::: {.columns}

::: {.column width="50%"}
```{r, eval=FALSE}
# base R
colors <- c("1990" = "lightsalmon","2010" = "skyblue")
plot(log10(Weight) ~ Height, brfss_male_DF,
  col = colors[Year], pch = 16, ylab = "log10(Weight)")
for (yr in levels(brfss_male_DF$Year)) {
  subset_data <- subset(brfss_male_DF, Year == yr)
  fit <- lm(log10(Weight) ~ Height, data = subset_data)
  abline(fit, col = colors[yr], lwd = 2)
}
legend("topleft", legend = levels(brfss_male_DF$Year), 
       col = colors, pch = 16, bty = "n")

```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# ggplot2
ggplot(brfss_male_tbl) +
    aes(x = Height, y = log10(Weight), color = Year) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "BRFSS Male Subset")
```
:::
:::


::: {.columns}

::: {.column width="50%"}
```{r, echo=FALSE}
# base R
colors <- c("1990" = "lightsalmon","2010" = "skyblue")
plot(log10(Weight) ~ Height, brfss_male_DF,
  col = colors[Year], pch = 16, ylab = "log10(Weight)")
for (yr in levels(brfss_male_DF$Year)) {
  subset_data <- subset(brfss_male_DF, Year == yr)
  fit <- lm(log10(Weight) ~ Height, data = subset_data)
  abline(fit, col = colors[yr], lwd = 2)
}
legend("topleft", legend = levels(brfss_male_DF$Year), 
       col = colors, pch = 16, bty = "n")
```
:::
::: {.column width="50%"}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# ggplot2
ggplot(brfss_male_tbl) +
    aes(x = Height, y = log10(Weight), color = Year) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "BRFSS Male Subset")
```
:::
:::


## Exercises

1. **Count the other way.** We tabulated `Year` with base R's `table()` and with
   `dplyr`'s `count()`. Do the same for the `Sex` column using *both* dialects,
   and confirm they report the same group sizes in different formats.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   # base R
   table(brfss_DF$Sex)
   # dplyr
   brfss_tbl |> count(Sex)
   ```
   `table()` returns a named vector with the counts under each level, while
   `count()` returns a tidy two-column tibble (`Sex` and `n`). Same numbers, two
   shapes.
   :::

2. **A boxplot, both ways.** Earlier we drew a boxplot of `Weight` by `Sex` for
   the 2010 data. Draw the equivalent for `Height` by `Sex`, once with base R's
   `plot()` and once with `ggplot2`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   # base R
   plot(Height ~ Sex, brfss_2010_DF)
   ```
   ```{r}
   # ggplot2
   ggplot(brfss_2010_tbl) +
       aes(x = Sex, y = Height) +
       geom_boxplot()
   ```
   The base R formula `Height ~ Sex` reads "Height as a function of Sex"; in
   `ggplot2` the same mapping is spelled out in `aes()`.
   :::

## Summary

You've now seen the same analysis — loading, tabulating, summarizing, and
plotting — written twice, in two of R's most common dialects:

- **Reading data**: `read.csv()` gives a base `data.frame`; `readr::read_csv()`
  gives a tibble, which is a data frame with friendlier printing and type
  guessing.
- **Summarizing**: base R reaches for `table()` and `aggregate()`; `dplyr` uses
  `count()`, `summarize()`, and `group_by()`, often reshaped with
  `tidyr::pivot_wider()`.
- **Plotting**: base graphics build a plot with `plot()` plus helpers like
  `lines()` and `legend()`; `ggplot2` builds it up in layers from an `aes()`
  mapping and `geom_*()` functions.

Neither dialect is the winner. Base R is always available and concise for quick
checks; the tidyverse shines when a pipeline of steps needs to stay readable.
Knowing both means you can read anyone's code and pick the clearer tool for the
job in front of you.
