---
author: "Lori Kern"
---
# Control Statements

Control statements help determine the flow and execution of commands based on
conditional statements. This chapter will cover a brief overview of the
following:

- Conditional Statements
  * if
  * if-else
  * ifelse

- Loops
  * for
  * while
  * repeat-break
  
- Special
  * break
  * return
  * next

- Other
  * nested
  * try / tryCatch
  * vectorization and apply functions


::: {.callout-important}
Pay attention to syntax. It is important to include all parenthesis and
brackets. In general if you have an open parenthesis or bracket, you will need a
closed parenthesis or bracket.
:::

::: {.callout-tip}
Good coding practices involve consistent indenting and spacing. Once a control
statement is initialized, all code in its brackets are indented to show clearly
what code/statements are being executed for that section. This becomes
especially important if you start nesting control statements.  
:::

## Conditional Statements

### if 

Let's start with an `if` statement. An if statement evaluates an expression and
depending on its result performs a sub-section of select code.

**Syntax:**

```{r, eval=FALSE}
if (expression){
  # additional code to run if expression is TRUE
  ...
}
```
The expression contained in parenthesis will result in a boolean (TRUE/FALSE)
value used to determine if the code in the braces should be executed. 

**Example:**

```{r}
x <- 12
if (x > 0){
  message(x, " is greater than 0")
  x <- 0
}
x
```

::: {.callout-tip}
Notice all lines in the braces are executed including a assignment that changes
our original value
:::

### if-else

An `if-else` statement adds additional code to be executed if the expression is
FALSE.

**Syntax:**

```{r, eval=FALSE}
if (expression){
  # additional code to run if expression is TRUE
  ...
} else {
  # additional code to run if expression if FALSE
  ...
}
```
We could read this allowed as if the expression is true execute this code else
if the expression is false execute this other code.

**Example:**

```{r}
if (x > 0){
  message(x, " is greater than 0")
  x <- 0
} else {
  message(x, " is not greater than 0")
  x <- x + 2
}
```

What is x now?


### ifelse

A specialized if-else statement is the `ifelse`. It is a simplified version
where an object can be coerced into logical form and return values for
true/false.

**Syntax:**

```{r, eval=FALSE}
ifelse(test_expression, yes_value, no_value)
```

**Example**

```{r}
num_vec <- -3:3
ifelse(num_vec >= 0, "positive", "negative")
```

## Loops

Loops are control statements that allow for repeated code execution either for a
set number of times, over a certain set of elements, or until a conditional
statement is met.

### for

A `for` loop will execute commands over a certain set of elements.

**Syntax:**

```{r, eval=FALSE}
for(value in vector){
  ## code to execute for each item in vector
  ...
}
```
value can be utilized in the executed code.

**Examples**

In this example, the vector of names is looped over, printing the number of
characters in each name.

```{r}
names <- c("Donna", "John", "Bradley", "Kara")
for(nm in names){
  print(paste(nm, "has", nchar(nm), "letters"))
}
```
In this example, for each value 1 to 5 (1,2,3,4,5), we take that value and add
to the current value of x. Notice how this updates x each time.

```{r}
x <- 0
for(i in 1:5){
  print(paste("add", i, "to", x))
  x <- x + i
}
```
In this example, we loop over the elements of a list. For each list element we
get the name of the item in the list and how many items that list element contains.

```{r}
my_list <- list(people=names,
                ages=c(54, 78, 40, 5, 25),
		animals=c("dog", "fish"))
for(i in seq_along(my_list)){
  list_element <- my_list[[i]]
  print(paste("List element", names(my_list)[i], "contains",
  length(my_list[[i]]), "values"))
}
```
### while

A `while` loop will execute until an expression is met.

**Syntax:**

```{r, eval=FALSE}
while(expression){
  ## code to execute until the expression is met
  ## be sure to update variable used in expression
  ...
}
```
In this example we will start at the value 1 and as long as that value stays
less than or equal to 5, we will print the value, incrementing by 1 each loop:

**Example**
```{r}
value <- 1
while (value <= 5){
  print(value)
  value = value + 1
}
```
::: {.callout-important}
Notice how we have to update the value that is being checked each time and that
it is logical that it should eventually reach a point where the loop exists. Be
cautious of **infinite loops**. These occur when the loop will never reach a
stopping point because the expression will never be FALSE
:::

### repeat

`repeat` is a indefinite loop. A `break` statement must be used to terminate the
loop.

**Syntax:**

```{r, eval=FALSE}
repeat{
  ## code to be evaluated
  if (condition){
    break
  }
}	
```
In this example we will repeat ourself until the breaking conditions reaches the
number of times we set to repeat. Notice how the code inside the loop alters the
variable used in the conditional statement. 

**Example**
```{r}
i <- 0
times <- 3
repeat{
  print("I am repeating myself")
  i <- i + 1
  if (i == times){
    break
  }
}
```

## Special

There are some special other options used to customize control statements. These
are used within control statements and can be especially useful in nested
statements. 

<center>
**break**
</center>
We have already seen `break` usage with repeat. Break will stop and exit the
control statement immediately when it hits.

<center>
**return**
</center>

`return` is similar to break in that it will stop and ext the control statement
immediately when it is hit, however it will return the result of the given
executed function or variable value upon exiting. It is generally used within
functions.

**Syntax:**
```{r, eval=FALSE}
return(expression)
```

This example creates a function that takes argument x.  If x is equal to 0 it
returns the string "zero". If it does not equal 0 it continues to execute
code. It will add 4 to the value. If that value is less than 0 it returns the
value, otherwise it returns the value multipled by 2. 

**Example**
```{r}
func <- function(x){
  if(x == 0){
    return("zero")
  }
  x <- x+4
  if (x <= 0){
    return(x)
  }else{
    return(x*2)
  } 
}

func(0)
func(-8)
func(6)
```

<center>
**next**
</center>

`next` will skip the current iteration of a loop without executing any further
statements without terminating the loop.

**Example**
```{r}
for(i in 1:10){
  if(i%%2 != 0){
    next
  }
  print(i)
}
```

## Other

### nesting

We already saw an example of a nesting built into the design of repeat-break;
nesting an if statement inside the repeat. All control statements can have
multiple nesting.
Nesting multiple for loops to loop over the row and columns of a matrix is such
an example.

**Example:**

In this example we create a numeric matrix with 5 rows and 3 columns and fill
the numbers 1 through 15 by column. Let us loop over by row and print out each
cell of the matrix. 

```{r}
mat <- matrix(1:15, ncol=3)
mat
for (r in seq(nrow(mat))) {
  for (c in seq(ncol(mat))) {
    print(mat[r, c])
  }
}

```

### try / tryCatch

This isn't necessarily a control statement but fits well enough to disucss. A
`tryCatch` statement is a way to handle code that may produce errors, warnings,
or other conditions that may arise.

**Syntax:**
```{r, eval=FALSE}
tryCatch(expr,
         error = function(e) {
	     ## error handling code 
	 },
         warning = function(w) {
	     ## warning handling code 
	 },
         finally = {
	    ## code to execute regardless of conditions
	 }
)
```
It can include all error, warning, finally or any subset of the three. expr is
an expression or block of code that is attempted to be executed. The error if a
function that is triggered if running the expr resulted in an error. The
argument e is generally an error object that contains details about the error
for reference or parsing if necessary. warning similar to error but handles
warnings with argument w. finally is code that executed regardless of what
happens. This is particularly useful for when the expr opens a connection such
as a database or file; finally can be used to close or clean up these open
connections.

### vectorization and apply functions

Some most common and simply implementations of control functions are to perform
functions over a vector or applying a function over the columns or rows of a
matrix. While control statement can be utilized, they may not be the most
efficient way to accomplish these tasks. It is encouraged to invesetigate if an
already existing vectorized function exists for common tasks (mean, average, etc
are already vectorized) or to utilized apply functions. Apply functions
include: `apply()`, `lapply()`, `sapply()`, `mapply()`, and `tapply()`. 

