---
title: "Control statements"
author: "Lori Kern"
---

So far your code has run straight down the page: every line executes, once, in
order. Real analyses rarely stay that simple. You want R to *decide* — drop a
sample only if its read count falls below a threshold — and to *repeat* — run the
same quality check across every one of a hundred samples without writing the line
a hundred times. **Control statements** are how you express those decisions and
repetitions.

This chapter introduces the two workhorses, conditionals (`if` / `else`) and
loops (`for`, `while`, `repeat`), along with the keywords that steer them
(`break`, `next`, `return`) and a first look at error handling with `tryCatch()`.
We close with a note on when *not* to write a loop at all — because in R, a
vectorized function is usually the better tool.

## What you'll learn

- Branch your code with `if`, `if`/`else`, and the vectorized `ifelse()`.
- Repeat work with `for`, `while`, and `repeat` loops, and choose the right one.
- Steer a loop with `break`, `next`, and `return`.
- Handle errors and warnings gracefully with `tryCatch()`.
- Recognize when a vectorized function or an `apply()`-family call beats a loop.

::: {.callout-important}
## Mind your parentheses and braces

Control statements lean heavily on punctuation. Every open parenthesis `(` needs
a matching close `)`, and every open brace `{` needs a matching `}`. A single
missing bracket is the most common reason a control statement refuses to run, so
keep them balanced.
:::

::: {.callout-tip}
## Indent the body

Once a control statement opens a `{ }` block, indent everything inside it. The
indentation has no effect on R, but it makes it obvious at a glance which lines
belong to the block — which matters enormously once you start nesting one
statement inside another.
:::

## Conditional statements

### `if`

An `if` statement evaluates an expression and, **only if** the result is `TRUE`,
runs the block of code in its braces.

**Syntax:**

```{r, eval=FALSE}
if (expression) {
  # code that runs only when expression is TRUE
  ...
}
```

The expression in parentheses must boil down to a single logical value —
`TRUE` or `FALSE` — and that value decides whether the braced code runs.

**Example:**

```{r}
x <- 12
if (x > 0) {
  message(x, " is greater than 0")
  x <- 0
}
x
```

::: {.callout-note}
## Every line in the block runs

When the condition is `TRUE`, *all* the lines in the braces execute — including
the assignment that reset `x` to `0`. That's why printing `x` afterward shows
`0`, not `12`.
:::

### `if`/`else`

An `if`/`else` statement adds a second block that runs when the expression is
`FALSE`. Exactly one of the two blocks always runs.

**Syntax:**

```{r, eval=FALSE}
if (expression) {
  # code that runs when expression is TRUE
  ...
} else {
  # code that runs when expression is FALSE
  ...
}
```

Read it aloud: *if* the expression is true, run this code; *else* (otherwise) run
that other code.

**Example:**

```{r}
if (x > 0) {
  message(x, " is greater than 0")
  x <- 0
} else {
  message(x, " is not greater than 0")
  x <- x + 2
}
x
```

Because `x` was `0` going in, the condition `x > 0` is `FALSE`, so the `else`
branch runs and `x` becomes `2`.

### `ifelse()`

`ifelse()` is a *vectorized* cousin of `if`/`else`. Instead of testing one value,
it tests an entire vector at once and returns a vector of results — one "yes" or
"no" value per element.

**Syntax:**

```{r, eval=FALSE}
ifelse(test_expression, yes_value, no_value)
```

**Example:**

```{r}
num_vec <- -3:3
ifelse(num_vec >= 0, "positive", "negative")
```

::: {.callout-warning}
## `if` is not `ifelse()`

A plain `if` expects a *single* `TRUE`/`FALSE`. Hand it a whole vector and, in
recent versions of R, it errors. When you want to test many values at once — a
column of a data frame, say — reach for `ifelse()`.
:::

## Loops

Loops repeat a block of code: a fixed number of times, once for each element of a
collection, or until some condition is met.

### `for`

A `for` loop runs its body once for each element of a vector or list, assigning
the current element to a variable you can use inside the loop.

**Syntax:**

```{r, eval=FALSE}
for (value in vector) {
  # code that runs once per element; `value` holds the current one
  ...
}
```

**Examples:**

Here we loop over a vector of names, printing how many characters each one has:

```{r}
names <- c("Donna", "John", "Bradley", "Kara")
for (nm in names) {
  print(paste(nm, "has", nchar(nm), "letters"))
}
```

Here, for each value `1` through `5`, we add it to a running total. Notice how
`x` is updated on every pass:

```{r}
x <- 0
for (i in 1:5) {
  print(paste("add", i, "to", x))
  x <- x + i
}
x
```

We can also loop over the elements of a list. `seq_along()` gives us the
positions `1, 2, 3, ...`, which we use both to pull each element out and to look
up its name:

```{r}
my_list <- list(
  people  = names,
  ages    = c(54, 78, 40, 5, 25),
  animals = c("dog", "fish")
)
for (i in seq_along(my_list)) {
  print(paste("List element", names(my_list)[i], "contains",
              length(my_list[[i]]), "values"))
}
```

### `while`

A `while` loop keeps running *as long as* its condition stays `TRUE`. It checks
the condition before each pass, so you must change something inside the loop that
eventually makes the condition `FALSE`.

**Syntax:**

```{r, eval=FALSE}
while (expression) {
  # code that runs while expression is TRUE
  # remember to update the variable the expression tests
  ...
}
```

**Example:**

Starting at `1`, we print the value and add `1` each pass, stopping once the
value passes `5`:

```{r}
value <- 1
while (value <= 5) {
  print(value)
  value <- value + 1
}
```

::: {.callout-warning}
## Beware the infinite loop

A `while` loop only ends when its condition becomes `FALSE`. If nothing inside
the loop ever makes that happen, the loop runs forever — an **infinite loop**.
Always make sure the body updates the variable the condition tests. (If you do
get stuck in one, press `Esc` or the stop button to interrupt R.)
:::

### `repeat`

A `repeat` loop has no condition of its own — it runs forever until a `break`
statement stops it. That makes `break` mandatory.

**Syntax:**

```{r, eval=FALSE}
repeat {
  # code to evaluate
  if (condition) {
    break
  }
}
```

**Example:**

Here we repeat ourselves until we've done it the set number of times. The body
increments `i`, and the `if` checks whether we've hit the limit:

```{r}
i <- 0
times <- 3
repeat {
  print("I am repeating myself")
  i <- i + 1
  if (i == times) {
    break
  }
}
```

## Steering a loop: `break`, `next`, and `return`

These keywords let you change a loop's flow from the inside. They're especially
handy in longer or nested statements.

### `break`

We've already seen `break` with `repeat`. It stops the loop immediately and exits
the moment it's reached — no further iterations run.

### `next`

`next` skips the rest of the *current* iteration and jumps straight to the next
one, without ending the loop. Here we print only the even numbers from 1 to 10,
using `next` to skip the odd ones:

```{r}
for (i in 1:10) {
  if (i %% 2 != 0) {
    next
  }
  print(i)
}
```

### `return`

`return` exits the surrounding **function** immediately and hands back a value.
It's used inside functions rather than bare loops.

**Syntax:**

```{r, eval=FALSE}
return(expression)
```

This function takes an argument `x`. If `x` is `0` it returns `"zero"`.
Otherwise it adds `4`; if the result is `0` or less it returns that value, and if
not it returns the value doubled:

```{r}
classify <- function(x) {
  if (x == 0) {
    return("zero")
  }
  x <- x + 4
  if (x <= 0) {
    return(x)
  } else {
    return(x * 2)
  }
}

classify(0)
classify(-8)
classify(6)
```

## Other useful patterns

### Nesting

You can place one control statement inside another — we already did, when the
`repeat` loop held an `if` block. Any statement can nest, to any depth. A classic
example is a pair of `for` loops walking the rows and columns of a matrix.

Here we build a numeric matrix with 5 rows and 3 columns, filled with `1` through
`15` by column, then loop over it row by row and print each cell:

```{r}
mat <- matrix(1:15, ncol = 3)
mat
for (i in seq(nrow(mat))) {
  for (j in seq(ncol(mat))) {
    print(mat[i, j])
  }
}
```

::: {.callout-tip}
## Don't name a loop variable `c`

It's tempting to use `r` and `c` for "row" and "column", but `c` is also R's
function for combining values (`c(1, 2, 3)`). Naming a variable `c` shadows that
function and invites confusing bugs. Using `i` and `j` keeps the built-in `c()`
available and is the conventional choice for loop counters.
:::

### `try()` / `tryCatch()`

Not strictly a control statement, but it fits the theme: `tryCatch()` lets you
*handle* code that might throw an error or a warning, instead of letting it halt
your whole script. This matters when you're looping over many files or samples
and don't want one bad case to stop the rest.

**Syntax:**

```{r, eval=FALSE}
tryCatch(
  expr,
  error = function(e) {
    # what to do if expr raises an error
  },
  warning = function(w) {
    # what to do if expr raises a warning
  },
  finally = {
    # code that runs no matter what
  }
)
```

You can supply any subset of `error`, `warning`, and `finally`. `expr` is the
code you're attempting. The `error` handler runs if `expr` fails; its argument
`e` is an error object carrying the details. The `warning` handler works the same
way for warnings, via `w`. The `finally` block runs regardless of the outcome —
which makes it the right place to close a file or database connection that `expr`
may have opened.

### Vectorization and the `apply()` family

Here's the twist: a lot of the work people reach for loops to do, R can do
*without* a loop — and faster. Many functions are already **vectorized**, meaning
they operate on a whole vector at once. To add `1` to every element of a vector,
you don't loop; you just write `x + 1`.

```{r}
x <- 1:5
x + 1          # vectorized: no loop needed
mean(x)        # already summarizes the whole vector
```

When there isn't a ready-made vectorized function, the `apply()` family applies a
function across a structure for you: `apply()` over the rows or columns of a
matrix, and `lapply()`, `sapply()`, `mapply()`, and `tapply()` over vectors and
lists. Before writing a loop, it's worth asking whether a vectorized function or
an `apply()` call would be clearer and quicker.

## Exercises

1. **Even or odd.** Write an `if`/`else` statement that prints `"even"` if a
   number `n` is even and `"odd"` otherwise. (Hint: `n %% 2` gives the remainder
   after dividing by 2.)

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   n <- 7
   if (n %% 2 == 0) {
     print("even")
   } else {
     print("odd")
   }
   ```
   `n %% 2` is `0` for even numbers and `1` for odd ones, so testing it against
   `0` tells the two apart.
   :::

2. **Sum a vector with a loop.** Using a `for` loop, add up the numbers in
   `c(4, 8, 15, 16, 23, 42)`. Then check your answer with the built-in `sum()`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   nums <- c(4, 8, 15, 16, 23, 42)
   total <- 0
   for (n in nums) {
     total <- total + n
   }
   total
   sum(nums)   # the vectorized way — same answer, far less typing
   ```
   The loop accumulates the running total one element at a time; `sum()` does the
   same thing in one vectorized call. This is exactly the kind of task where a
   vectorized function beats a hand-written loop.
   :::

3. **Skip the missing values.** Loop over `c(3, NA, 7, NA, 2)` and print only the
   values that are *not* `NA`, using `next` to skip the missing ones. (Hint:
   `is.na(x)` is `TRUE` when `x` is missing.)

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   vals <- c(3, NA, 7, NA, 2)
   for (v in vals) {
     if (is.na(v)) {
       next
     }
     print(v)
   }
   ```
   When `v` is `NA`, `next` jumps to the following iteration before the `print()`
   line is reached, so only the real numbers are printed.
   :::

## Summary

You can now control the flow of your code rather than running straight down the
page:

- **Branch** with `if` and `if`/`else` for single values, and `ifelse()` to test a
  whole vector at once.
- **Repeat** with `for` (once per element), `while` (until a condition flips), and
  `repeat` (until you `break`).
- **Steer** loops with `break` (stop), `next` (skip), and `return` (exit a
  function with a value).
- **Guard** risky code with `tryCatch()` so one error doesn't sink an entire run.
- **Reach for vectorization first.** Many tasks that look like they need a loop
  are better done with a vectorized function or an `apply()`-family call.
