---
title: "Data Frames"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

Almost every dataset you will meet in genomics arrives as a table: genes down the
side, samples across the top, measurements in the middle, and a few columns of
annotation bolted on. In R, the structure built for exactly this shape is the
`data.frame`. Get comfortable with it and most of the rest of the book — loading
expression data, filtering genes, summarizing by condition — becomes variations
on a theme you already know.

A data.frame looks a bit like an R matrix in that it has two dimensions, rows and
columns. The difference that matters is this: a matrix **must** hold a single data
type, while a data.frame may hold a *different* data type in each column — numbers
in one, gene names in another, a logical flag in a third. That is exactly what a
real dataset looks like, which is why the data.frame, not the matrix, is the
workhorse for tabular data in R.

## What you'll learn

By the end of this chapter you will be able to:

- Explain how a `data.frame` differs from a matrix, and why that difference matters.
- Build a `data.frame` from scratch with `data.frame()`.
- Inspect a data frame with `str()`, `head()`, `dim()`, `nrow()`, `ncol()`, and `summary()`.
- Pull out columns with `$` and `[[ ]]`, and rows and columns with `[rows, columns]`.
- Add a new column and select rows by a logical condition.
- Save a data frame back to disk with `write.csv()`.

We'll do all of this on a small gene-expression table that we build ourselves, so
there is nothing to download and every value is in plain sight.

## Building a data.frame from scratch

The most direct way to understand a data.frame is to make one. Recall from the
[vectors](vectors.qmd) chapter that a column of a data frame is just a vector. So a
data.frame is really a set of equal-length vectors stood up side by side, each one
becoming a named column. The `data.frame()` function does exactly that assembly.

Imagine we have measured a handful of genes: each gene has a symbol, the
chromosome it sits on, an average expression value, the log~2~ fold change between
a tumor and a normal sample, and a flag for whether it is a known cancer gene.
Each of those is a vector, and they are all the same length — one entry per gene.

```{r}
genes <- data.frame(
  symbol     = c("TP53", "BRCA1", "EGFR", "MYC", "PTEN", "KRAS", "ALK", "RB1"),
  chromosome = c("17", "17", "7", "8", "10", "12", "2", "13"),
  expression = c(8.2, 5.1, 11.4, 9.8, 6.7, 7.3, 4.2, 5.9),
  log2fc     = c(-1.4, -0.8, 2.1, 3.0, -2.2, 1.1, 0.3, -1.7),
  is_cancer  = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE)
)
genes
```

Notice that R lined up our five vectors into five columns and numbered the rows
`1` through `8` down the left edge. Those row numbers are the default *row names*.
Each column kept its own type: `symbol` and `chromosome` are character, `expression`
and `log2fc` are numeric, and `is_cancer` is logical. That mix of types in one
object is the whole point of a data.frame — and the one thing a matrix could never
do.

We can confirm that R really did give us a `data.frame`:

```{r}
class(genes)
```

## Inspecting a data.frame

Our table is small enough to print in full, but real datasets have thousands or
tens of thousands of rows and you will not be able to look at them all at once.
Fortunately, R gives you a toolkit for inspecting a data.frame a piece at a time.

- Structure at a glance
  + `str()` to see the columns, their types, and a preview of each
- Overviews of content
  + `head()` to show the first few rows
  + `tail()` to show the last few rows
- Size
  + `dim()` for dimensions (rows, columns)
  + `nrow()` for the number of rows
  + `ncol()` for the number of columns
- Names and summaries
  + `colnames()` for the column names
  + `rownames()` for the row "names" (often just the row numbers)
  + `summary()` for a per-column thumbnail of the data

The single most useful one is `str()`, which you met in the
[lists](lists.qmd) chapter. It reports the structure: how many rows ("obs."), how
many columns ("variables"), and for each column its name, type, and first few values.

```{r}
str(genes)
```

`head()` and `tail()` show the top and bottom of the table — invaluable when the
full thing is too long to read:

```{r}
head(genes, 3)
tail(genes, 3)
```

The size functions answer "how big is this?" at a glance. `dim()` returns the rows
and columns together; `nrow()` and `ncol()` return them one at a time, which is
handy inside other calculations:

```{r}
dim(genes)
nrow(genes)
ncol(genes)
```

The name functions tell you what the rows and columns are called. `colnames()` is
the one you'll reach for constantly — it's a menu of everything you can ask for:

```{r}
colnames(genes)
rownames(genes)
```

Finally, `summary()` walks across the columns and gives each an appropriate
thumbnail: numeric ranges and quartiles for the measurement columns, simple counts
for the others.

```{r}
summary(genes)
```

::: {.callout-note}
## Look before you leap with `View()`

In RStudio there is one more inspection tool, `View()` (note the capital "V"), that
opens the data frame in a spreadsheet-style window you can scroll and sort. It is a
great way to *get a feel* for a dataset before you start writing code against it.
We don't run it here because it only works in interactive RStudio, not when the
book is rendered.

```{r eval=FALSE}
View(genes)
```
:::

## Accessing columns

Because pulling out a single column is such a common operation, R gives data.frames
two shortcuts for it. The first, the `$` operator, takes the data frame on the left
and a column name on the right:

```{r}
genes$symbol
genes$expression
```

Each of these hands you back a plain vector — exactly the vector you put in when
you built the data frame. So everything you learned about vectors applies directly.
You can, for example, take the mean of the expression column:

```{r}
mean(genes$expression)
```

The second shortcut comes from a fact worth remembering: in R, a data.frame is also
a list, where each column is one element. So the list notation `[[ ]]` works too,
either by position or by name:

```{r}
genes[[3]]               # the third column, by position
genes[["expression"]]    # the same column, by name
```

::: {.callout-note}
## `$` versus `[[ ]]`

Both pull out a single column as a plain vector, so why two notations? `$` is the
quick, everyday form when you know the column name as you type it
(`genes$expression`). `[[ ]]` is the form to reach for when the column is chosen by
a *value* — a number (`genes[[3]]`) or a name stored in a variable
(`col <- "expression"; genes[[col]]`), which `$` cannot do. Reach for `$` by default
and `[[ ]]` when you're programming with the column choice.
:::

## Subsetting with `[rows, columns]`

A data.frame has two dimensions, so to pull out a rectangular piece we use the
square brackets with **two** values inside, separated by a comma: `[rows, columns]`.
Leaving one side of the comma blank means "all of them."

To get the first three rows and *all* columns, fill in the rows and leave the
columns blank:

```{r}
genes[1:3, ]
```

To get specific rows *and* specific columns, fill in both:

```{r}
genes[1:3, c("symbol", "expression")]
```

And to get one whole column while leaving the rows blank, name the column on the
right of the comma:

```{r}
genes[, "chromosome"]
```

::: {.callout-tip}
## The comma is the whole trick

With a one-dimensional vector you write `x[i]` — one value in the brackets. With a
two-dimensional data.frame you write `df[i, j]` — two values, separated by a comma,
meaning `[rows, columns]`. If you ever see an error like *"incorrect number of
dimensions,"* you've almost always forgotten the comma.
:::

## Adding a column

Adding a column works just like assigning to a column that doesn't exist yet: name
it with `$` on the left of the assignment arrow, and give it a vector of the right
length. Here we add an `abs_log2fc` column holding the *absolute* fold change,
which is often what we sort on when we care about the size of a change regardless
of its direction:

```{r}
genes$abs_log2fc <- abs(genes$log2fc)
genes
```

The data frame now has six columns instead of five. Because `abs()` is vectorized
(see the [vectors](vectors.qmd) chapter), one short line computed the new value for
every row at once.

## Selecting rows by a logical condition

This is where data.frames earn their keep. Recall the logical-indexing pattern from
the vectors chapter: ask a yes/no question of a column, get back a `TRUE`/`FALSE`
vector, then use it to keep only the rows where the answer is `TRUE`. With a
data.frame, that logical vector goes in the *row* position of `[rows, columns]`.

For example, keep only the genes with expression above 7:

```{r}
genes[genes$expression > 7, ]
```

Read the inside first: `genes$expression > 7` produces a logical vector — one
`TRUE`/`FALSE` per gene — and `genes[..., ]` keeps the rows where it is `TRUE`, with
the blank after the comma keeping all the columns.

You can combine conditions with `&` ("and") and `|` ("or"). Here we ask for cancer
genes whose expression is also above 7:

```{r}
genes[genes$is_cancer & genes$expression > 7, ]
```

And of course you can select rows and columns at the same time — keep the up-regulated
genes (`log2fc > 0`) but show only their symbol and fold change:

```{r}
genes[genes$log2fc > 0, c("symbol", "log2fc")]
```

::: {.callout-tip}
## `subset()`: the same idea, less typing

Writing `genes$` in front of every column name gets repetitive. The `subset()`
function lets you refer to the columns by their bare names, because it knows to look
them up inside the data frame:

```{r}
subset(genes, is_cancer & expression > 7)
```

It reads almost like English: *subset the genes where `is_cancer` is true and
`expression` exceeds 7.* Use whichever form is clearer to you; they do the same job.
:::

## Saving a data.frame

Once you have a data.frame worth keeping, you'll want to save it. The most portable
way is to write it to a `.csv` file (comma-separated values), which any other
program — R, Python, even a spreadsheet — can read. The `write.csv()` function does
this. We pass `row.names = FALSE` so the default row numbers aren't written as an
extra unnamed column:

```{r eval=FALSE}
write.csv(genes, "genes.csv", row.names = FALSE)
```

Reading it back later is just as direct, with `read.csv()`:

```{r eval=FALSE}
genes <- read.csv("genes.csv")
```

The function names follow a pattern worth noticing: `read.csv()` and `write.csv()`
are mirror images, and the `.csv` in each name tells you the file format. As always,
`help('read.csv')` is worth a look for the many options these functions accept.

## Exercises

These all use the `genes` data frame we built earlier in the chapter (including the
`abs_log2fc` column we added).

1. **How many genes, how many columns?** Report the number of rows and the number
   of columns in `genes` two ways: with a single call to `dim()`, and with `nrow()`
   and `ncol()` separately.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   dim(genes)
   nrow(genes)
   ncol(genes)
   ```
   `dim()` returns both numbers as a length-2 vector (rows first, then columns);
   `nrow()` and `ncol()` return each one on its own, which is handier inside other
   calculations.
   :::

2. **Extract a column and summarize it.** Pull the `log2fc` column out as a plain
   vector and compute its mean. Confirm the extracted object is numeric.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   fc <- genes$log2fc
   class(fc)
   mean(fc)
   ```
   `$` extracts the column as the plain numeric vector you put in, so ordinary
   vector functions like `mean()` work on it directly.
   :::

3. **Subset by condition.** Keep only the rows where `is_cancer` is `TRUE` *and*
   `log2fc` is negative (down-regulated cancer genes). How many are there?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   down_cancer <- genes[genes$is_cancer & genes$log2fc < 0, ]
   down_cancer
   nrow(down_cancer)
   ```
   The two conditions are joined with `&`; the combined logical vector goes in the
   *row* position of `[rows, columns]`, and the blank after the comma keeps every
   column. `nrow()` then counts the surviving rows.
   :::

4. **Add a column.** Add a logical column called `strong` that is `TRUE` when a
   gene's absolute fold change (`abs_log2fc`) is at least 2. Then count how many
   genes are `strong`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   genes$strong <- genes$abs_log2fc >= 2
   sum(genes$strong)
   ```
   Assigning to `genes$strong` creates the new column from a logical comparison.
   Because `TRUE` counts as 1, `sum()` of a logical column tallies the `TRUE`s — the
   same count-by-summing trick from the vectors chapter.
   :::

## Summary

You can now build a `data.frame` from a set of equal-length vectors, inspect its
shape and contents, pull out the rows and columns you care about, add new columns,
and write the result back to disk. Specifically, you can:

- **Build** a data frame with `data.frame(name = vector, ...)`, mixing column types
  freely — the one thing a matrix cannot do.
- **Inspect** it with `str()` (structure), `head()`/`tail()` (content), `dim()` /
  `nrow()` / `ncol()` (size), and `summary()` (per-column thumbnail).
- **Extract a column** with `$` or `[[ ]]`, getting back a plain vector.
- **Subset** a rectangle with `[rows, columns]`, where a blank means "all."
- **Filter rows** by a logical condition in the row position — the single most
  important move, and the same logical-indexing pattern you learned for vectors.
- **Add** a column by assigning to `df$newname`, and **save** with `write.csv()`.

The logical-indexing pattern — ask a yes/no question of a column, then use the
answer to keep rows — is the move you'll repeat on every dataset for the rest of the
book.

## Reflection

- Can you state one thing a `data.frame` can do that a matrix cannot?
- Given the column `genes$expression`, can you write the subset that keeps only rows
  where `expression < 6`?
- What's the difference between `genes[2]`, `genes[[2]]`, and `genes[, 2]`?
