---
title: "Data Frames"
---

Almost every dataset you will meet in genomics arrives as a table: genes down the
side, samples across the top, measurements in the middle, and a few columns of
annotation bolted on. In R, the structure built for exactly this shape is the
`data.frame`. Get comfortable with it and most of the rest of the book — loading
expression data, filtering genes, summarizing by condition — becomes variations
on a theme you already know.

A data.frame looks a bit like an R matrix in that it has two dimensions, rows and
columns. The difference that matters is this: a matrix **must** hold a single data
type, while a data.frame may hold a *different* data type in each column — numbers
in one, gene names in another, an experimental condition in a third. That is
exactly what a real dataset looks like, which is why the data.frame, not the
matrix, is the workhorse for tabular data in R.

## What you'll learn

By the end of this chapter you will be able to:

- Explain how a `data.frame` differs from a matrix, and why that difference matters.
- Load tabular data from a file or URL with `read.csv()`.
- Inspect a data frame with `head()`, `dim()`, `summary()`, and friends.
- Pull out columns and rows by position, name, and logical condition.
- Summarize measurements by category with `aggregate()`.
- Build a `data.frame` from scratch and save it back to disk.

We'll do all of this on a real yeast gene-expression dataset.

## Dataset

The data used here are borrowed directly from the [fantastic
Bioconnector tutorials](http://bioconnector.org/index.html) and are a
cleaned up version of the data from [Brauer et al. Coordination of
Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in
Yeast (2008) Mol Biol Cell
19:352-367](http://www.ncbi.nlm.nih.gov/pubmed/17959824). These data
are from a gene expression microarray, and in this paper the authors
examine the relationship between growth rate and gene expression in
yeast cultures limited by one of six different nutrients (glucose,
leucine, ammonium, sulfate, phosphate, uracil). If you give yeast a
rich media loaded with nutrients except restrict the supply of a
single nutrient, you can control the growth rate to any rate you
choose. By starving yeast of specific nutrients you can find genes
that:

1. Raise or lower their expression in response to growth
   rate. Growth-rate dependent expression patterns can tell us a lot
   about cell cycle control, and how the cell responds to stress. The
   authors found that expression of >25% of all yeast genes is
   linearly correlated with growth rate, independent of the limiting
   nutrient. They also found that the subset of negatively
   growth-correlated genes is enriched for peroxisomal functions, and
   positively correlated genes mainly encode ribosomal functions.
2. Respond differently when different nutrients are being limited. If
   you see particular genes that respond very differently when a
   nutrient is sharply restricted, these genes might be involved in
   the transport or metabolism of that specific nutrient.

The dataset can be downloaded directly from:

- [brauer2007_tidy.csv](https://raw.githubusercontent.com/bioconnector/workshops/master/data/brauer2007_tidy.csv)

We are going to read this dataset into R and then use it as a playground for learning about data.frames.

## Reading in data

R has many capabilities for reading in data. Many of the functions
have names that help us to understand what data format is to be
expected. In this case, the filename that we want to read ends in
`.csv`, meaning comma-separated-values. The `read.csv()` function
reads in `.csv` files.  As usual, it is worth reading
`help('read.csv')` to get a better sense of the possible
bells-and-whistles.

The `read.csv()` function can read directly from a URL, so we do not
need to download the file directly.  This dataset is relatively large
(about 16MB), so this may take a bit depending on your network
connection speed.

```{r}
options(width=60)
```


```{r cache=TRUE}
url <- paste0(
    'https://raw.githubusercontent.com',
    '/bioconnector/workshops/master/data/brauer2007_tidy.csv'
)
ydat <- read.csv(url)
```

Our variable, `ydat`, now "contains" the downloaded and read data. We can check to see what data type
read.csv gave us:

```{r}
class(ydat)
```

R reports `"data.frame"`, so `read.csv()` did what its name promises: it turned a
table of text into the structure we want to work with.

## Inspecting data.frames

Your `ydat` variable is a data.frame. The dataset is fairly large, so you will not be
able to look at it all at once on the screen. Fortunately, R gives you many tools to
inspect a data.frame a piece at a time.

- Overviews of content
  + `head()` to show first few rows
  + `tail()` to show last few rows
- Size
  + `dim()` for dimensions (rows, columns)
  + `nrow()`
  + `ncol()`
  + `object.size()` for power users interested in the memory used to store an object
- Data and attribute summaries
  + `colnames()` to get the names of the columns
  + `rownames()` to get the "names" of the rows--may not be present
  + `summary()` to get per-column summaries of the data in the data.frame.
  
```{r dataframeinspection}
head(ydat)
tail(ydat)
dim(ydat)
nrow(ydat)
ncol(ydat)
colnames(ydat)
summary(ydat)
```

The `dim()` output tells you the shape at a glance — tens of thousands of rows and a
handful of columns — and `summary()` gives a per-column thumbnail: numeric ranges for
the measurement columns and counts for the categorical ones.

::: {.callout-note}
## Look before you leap with `View()`

In RStudio there is one more inspection tool, `View()` (note the capital "V"), that
opens the first 1000 rows in a spreadsheet-style window you can scroll and sort. It
is a great way to *get a feel* for a dataset before you start writing code against
it. We don't run it here because it only works in interactive RStudio, not when the
book is rendered.

```{r eval=FALSE}
View(ydat)
```
:::

## Accessing variables (columns) and subsetting

In R, data.frames can be subset similarly to other two-dimensional data structures. The `[` in R is used 
to denote subsetting of any kind. When working with two-dimensional data, we need two values inside the `[ ]`
to specify the details. The specification is `[rows, columns]`. For example, to get the first three rows of `ydat`,
use:

```{r}
ydat[1:3, ]
```

Note how the second number, the columns, is blank. R takes that to mean "all the columns". Similarly, we can
combine rows and columns specification arbitrarily.

```{r}
ydat[1:3, 1:3]
```

Because selecting a single variable, or column, is such a common operation, there are two shortcuts for
doing so *with data.frames*. The first, the `$` operator works like so:

```{r}
# Look at the column names, just to refresh memory
colnames(ydat)
# Note that I am using "head" here to limit the output
head(ydat$symbol)
# What is the actual length of "symbol"?
length(ydat$symbol)
```

The second is related to the fact that, in R, data.frames are also lists. We subset a list by using `[[]]` notation.
To get the second column of `ydat`, we can use:

```{r}
head(ydat[[2]])
```

Alternatively, we can use the column name:

```{r}
head(ydat[["systematic_name"]])
```

::: {.callout-note}
## `$` versus `[[ ]]`

Both pull out a single column as a plain vector, so why two notations? `$` is the
quick, everyday form when you know the column name as you type it
(`ydat$symbol`). `[[ ]]` is the form to reach for when the column is chosen by a
*value* — a number (`ydat[[2]]`) or a name stored in a variable
(`col <- "symbol"; ydat[[col]]`), which `$` cannot do. Reach for `$` by default and
`[[ ]]` when you're programming with the column choice.
:::

## Aggregating and exploring

There are a couple of columns that hold numeric values. You can confirm which by
asking each column its `class()`:

```{r}
class(ydat$symbol)
class(ydat$rate)
class(ydat$expression)
```

So `symbol` is a character column, while `rate` and `expression` are numeric. We'll
put those numeric columns to work in the exercises at the end of the chapter.

### More advanced indexing and subsetting

We can use, for example, logical values (TRUE/FALSE) to subset data.frames. Here we
keep only the rows where the `symbol` column equals `LEU1`:

```{r}
head(ydat[ydat$symbol == 'LEU1', ])
tail(ydat[ydat$symbol == 'LEU1', ])
```

Notice the rows full of `NA` that came along for the ride. That is not a bug in your
code — it is one of the most common surprises in R, and worth understanding once so
it never trips you up again.

::: {.callout-warning}
## Logical subsetting and missing values

When you filter with a condition like `ydat$symbol == 'LEU1'`, any row whose
`symbol` is `NA` returns `NA` for the test — not `FALSE` — so R keeps it, and you
get phantom rows full of `NA`. The fix is to exclude the missing values explicitly
with `!is.na()`:

```{r}
head(ydat[ydat$symbol == 'LEU1' & !is.na(ydat$symbol), ])
```

Now only the genuine `LEU1` rows remain.
:::

Sometimes, looking at the data themselves is not that important. Using `dim()` is one possibility to look at the
number of rows and columns after subsetting.

```{r}
dim(ydat[ydat$expression > 3, ])
```

Find the high expressed genes when leucine-starved. For this task we can also use `subset` which allows 
us to treat column names as R variables (no `$` needed).

```{r}
subset(ydat, nutrient == 'Leucine' & rate == 0.05 & expression > 3)
```

## Aggregating data

Aggregating data, or summarizing by category, is a common way to look
for trends or differences in measurements between categories. Use `aggregate`
to find the mean expression by gene symbol.

```{r}
head(aggregate(ydat$expression, by = list(ydat$symbol), mean))
# or
head(aggregate(expression ~ symbol, mean, data = ydat))
```

Both calls do the same thing: collapse the many measurements for each gene into a
single mean expression value, one row per `symbol`. The formula form
(`expression ~ symbol`) reads as "expression *by* symbol" and is usually the easier
to type and to read.

## Creating a data.frame from scratch

Sometimes it is useful to combine related data into one object. For example, let's simulate some data.

```{r}
set.seed(42)
smoker <- factor(rep(c("smoker", "non-smoker"), each = 50))
smoker_numeric <- as.numeric(smoker)
x <- rnorm(100)
risk <- x + 2 * smoker_numeric
```

We set a seed first with `set.seed(42)` so the random draws are reproducible — you
and a reader will see the same numbers. We now have two related variables, `risk`
and `smoker`. We can make a data.frame out of them:

```{r}
smoker_risk <- data.frame(smoker = smoker, risk = risk)
head(smoker_risk)
```

R also has plotting shortcuts that work with data.frames. Because `smoker` is a
factor, `plot()` recognizes the `risk ~ smoker` formula and draws a boxplot of risk
in each group:

```{r}
plot(risk ~ smoker, data = smoker_risk)
```

As we built the data, smokers carry the higher simulated risk — and the boxplot
shows exactly that.

## Saving a data.frame

Once we have a data.frame of interest, we may want to save it. The most
portable way to save a data.frame is to use one of the `write` functions. In
this case, let's save the data as a `.csv` file.

```{r eval=FALSE}
write.csv(smoker_risk, "smoker_risk.csv")
```

## Exercises

These all use the yeast dataset, `ydat`, from earlier in the chapter.

1. Which columns of `ydat` are numeric? Check `rate` and `expression`.
2. Make a histogram of the `expression` values, then of the `rate` values.
3. `rate` has only a handful of distinct values. Use `table()` to count how many
   rows fall at each rate. Which `rate` corresponds to the most nutrient-starved
   condition?

::: {.callout-note collapse="true"}
## Solutions

```{r}
class(ydat$rate); class(ydat$expression)   # both numeric
hist(ydat$expression)
hist(ydat$rate)
table(ydat$rate)                            # the smallest rate = most starved
```

Growth rate is the *limiting nutrient* dialed down, so the **lowest** `rate`
(0.05) is the most starved condition.
:::

## Summary

You can now load a real dataset into a `data.frame`, inspect its shape and
contents, pull out the rows and columns you care about (by position, name, or a
logical test), summarize values by category with `aggregate()`, and write results
back to disk. These are the moves you'll repeat on every dataset for the rest of the
book.

## Reflection

- Can you state one thing a `data.frame` can do that a matrix cannot?
- Given a column `ydat$rate`, can you write the subset that keeps only rows where
  `rate < 0.1`?
- When would you reach for `aggregate()` instead of subsetting?
