---
title: "Reshaping tables with dplyr"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

Open almost any biological analysis and a table is sitting at the center of it.
A sample sheet has one row per sample. A count matrix has one row per gene. A
differential-expression result has one row per transcript, with columns for the
fold change, the p-value, and the adjusted p-value. The data changes from study
to study, but the *moves* you make on those tables hardly change at all: keep the
rows that matter, throw away the columns that don't, sort by some number, compute
a derived quantity, and boil many rows down to a handful of summary numbers.

The **dplyr** package gives each of those moves a name and a function. The
functions are small, they each do one job, and they snap together. Once you can
do these moves on a table of mammal sleep times, you can do them — unchanged — on
a table of gene counts. That is the whole bet of this chapter: learn the moves on
something easy to picture, then carry them straight over to data you actually
care about.

## What you'll learn

- Keep and drop columns with `select()`.
- Keep rows that pass a test with `filter()`.
- Sort rows with `arrange()`, including largest-first with `desc()`.
- Build new columns from old ones with `mutate()`.
- Reduce a column to one number with `summarise()`.
- Compute those summaries one-per-group with `group_by()` — the
  *split-apply-combine* idea.
- Chain all of these together with the pipe, `|>`, so an analysis reads top to
  bottom.

::: {.callout-tip}
## Why mammal sleep in a genomics book?

We'll practice on `msleep`, a small table of how long different mammals sleep.
It's a stand-in, picked because it's tiny enough to see all at once and the
columns need no explaining. But hold the genomics version in your head the whole
time: filtering to mammals that sleep more than 16 hours is the *same operation*
as filtering to genes with an adjusted p-value below 0.05; pulling out two
columns is the same as pulling `sample_id` and `condition` from a sample sheet;
and averaging sleep within each taxonomic order is the same as averaging
expression within each treatment group. Same verbs, bigger stakes.
:::

## A small table to practice on

The `msleep` dataset records sleep times, brain weights, and body weights for 83
mammals, drawn from a study by Savage and West[^msleep]. It ships inside
**ggplot2**, so there is nothing to download — install ggplot2 once if you don't
already have it, then load it to make the data available.

```{r eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
data(msleep)
```

A few of the columns will do most of the work for us: `name` (the common name),
`order` and `vore` (taxonomic order and feeding type — carnivore, herbivore,
omnivore, or insectivore), `sleep_total` and `sleep_rem` (hours of total and REM
sleep), and `bodywt` and `brainwt` (weights in kilograms). The full set of 11
columns is described on the help page:

```{r eval=FALSE}
?msleep
```

Glancing at the first few rows is a good habit before you touch any table:

```{r}
head(msleep)
```

The verbs we're about to meet live in dplyr. Install it once, then load it:

```{r eval=FALSE}
install.packages("dplyr")
```

```{r}
library(dplyr)
```

## `select()`: choose columns

`select()` returns a table with only the columns you ask for, in the order you
ask for them. Notice that you write the column names bare — no quotes — as if they
were already variables in scope:

```{r}
select(msleep, name, sleep_total)
```

A leading minus sign means "everything *except* this." To drop the `name` column
and keep the rest:

```{r}
head(select(msleep, -name))
```

A colon picks a contiguous run of columns by name, the same way `1:5` picks a run
of numbers:

```{r}
head(select(msleep, name:order))
```

When column names follow a pattern, dplyr's selection helpers save typing. To
grab every column whose name begins with `"sleep"`:

```{r}
head(select(msleep, starts_with("sleep")))
```

There are companions to `starts_with()` — `ends_with()`, `contains()` (a literal
substring), and `matches()` (a regular expression) — that all select columns by
their names. They matter more than they look: a real count matrix might have
hundreds of columns, and naming them one at a time is not an option.

## `filter()`: choose rows

Where `select()` works on columns, `filter()` works on rows. You give it a
logical test, and it keeps the rows where the test is `TRUE`. Here are the
mammals that sleep at least 16 hours a day:

```{r}
filter(msleep, sleep_total >= 16)
```

The test is an ordinary comparison of the kind you met in the Vectors chapter —
`sleep_total >= 16` is `TRUE` or `FALSE` for each row, and `filter()` keeps the
`TRUE` ones. Any comparison works: `>`, `<`, `>=`, `<=`, `==`, `!=`.

List several conditions, separated by commas, and a row must satisfy *all* of
them — the comma acts as "and." Mammals that both sleep a lot *and* are not tiny:

```{r}
filter(msleep, sleep_total >= 16, bodywt >= 1)
```

To test membership in a set of allowed values, reach for `%in%`. It asks, for
each row, "is this value one of these?" Here we keep two taxonomic orders at once:

```{r}
filter(msleep, order %in% c("Perissodactyla", "Primates"))
```

This is exactly the operation behind "keep only the genes that passed
significance" or "keep only the samples from these three batches."

## The pipe: reading an analysis top to bottom

So far each call has stood alone. Real work strings these verbs together, and
nesting the calls — `head(select(filter(msleep, ...), ...))` — gets unreadable
fast, because you have to read it inside-out. The **pipe** fixes this. Written
`|>`, it takes whatever is on its left and feeds it as the *first argument* to the
function on its right.

::: {.callout-tip}
## Read `|>` as "and then"

`x |> f() |> g()` means: start with `x`, **and then** apply `f`, **and then**
apply `g`. The steps appear in the order they actually happen, top to bottom,
instead of turned inside-out by nesting. Older code uses `%>%` from the magrittr
package for the same idea; `|>` is built into R itself, and the two behave the
same for everything we do here.
:::

Compare the nested form with the piped form. Nested, read inside-out:

```{r}
head(select(msleep, name, sleep_total))
```

Piped, read top to bottom — take `msleep`, **and then** select two columns,
**and then** show the head:

```{r}
msleep |>
    select(name, sleep_total) |>
    head()
```

Both produce the same answer. The piped version wins the moment a chain grows
past two steps, so we'll use it from here on.

## `arrange()`: sort rows

`arrange()` reorders the rows by one or more columns, smallest first by default.
Sort by taxonomic order:

```{r}
msleep |> arrange(order) |> head()
```

Name more than one column and `arrange()` sorts by the first, breaking ties with
the second. Here we select three columns, then sort by order and, within each
order, by total sleep:

```{r}
msleep |>
    select(name, order, sleep_total) |>
    arrange(order, sleep_total) |>
    head()
```

Wrap a column in `desc()` to sort it from largest to smallest. The sleepiest
mammals within each order, heaviest sleepers first:

```{r}
msleep |>
    select(name, order, sleep_total) |>
    arrange(order, desc(sleep_total)) |>
    head()
```

Sorting a results table by p-value, or a count table by total expression, is this
same verb.

## `mutate()`: build new columns

`mutate()` adds columns computed from the ones already there, leaving the original
columns in place. REM sleep as a fraction of total sleep is a natural quantity to
derive:

```{r}
msleep |>
    mutate(rem_fraction = sleep_rem / sleep_total) |>
    select(name, sleep_total, sleep_rem, rem_fraction) |>
    head()
```

You can compute several new columns in one call, separated by commas, and a later
one may even use a column the same `mutate()` just created. Here we convert body
weight to grams and flag the heavyweights:

```{r}
msleep |>
    mutate(bodywt_g = bodywt * 1000,
           is_large = bodywt_g > 10000) |>
    select(name, bodywt, bodywt_g, is_large) |>
    head()
```

Turning raw counts into counts-per-million, or a fold change into a log fold
change, is `mutate()` doing exactly this job on a bigger table.

## `summarise()`: reduce a column to one number

`summarise()` collapses a whole column down to a single summary value. The mean
amount of sleep across all 83 mammals:

```{r}
msleep |>
    summarise(avg_sleep = mean(sleep_total))
```

Ask for several summaries at once and you get a one-row table with one column per
answer. Alongside `mean()`, the usual reducers — `min()`, `max()`, `median()`,
`sd()`, `sum()`, and `n()` (which simply counts the rows) — are all fair game:

```{r}
msleep |>
    summarise(avg_sleep = mean(sleep_total),
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              n_mammals = n())
```

That single row reports the average, the extremes, and the sample size together.

## `group_by()`: one summary per group

`summarise()` on its own gives one number for the whole table. Pair it with
`group_by()` and you get one number *per group* instead — and almost every
interesting summary is a per-group summary.

::: {.callout-note}
## Split, apply, combine

`group_by()` doesn't change the data; it tags the rows with which group each
belongs to. Picture the table sorted into piles, one pile per taxonomic order. The
next `summarise()` runs *separately* on each pile (**apply**) and stacks the
per-pile answers back into one tidy table (**combine**). Splitting the data,
applying a calculation to each part, and combining the results is such a common
shape that it has a name: **split-apply-combine**. Base R does it with
`aggregate()` and `tapply()`; dplyr just makes it read like a sentence.
:::

Group by taxonomic order, then ask for the same summaries as before. The result
has one row per order:

```{r}
msleep |>
    group_by(order) |>
    summarise(avg_sleep = mean(sleep_total),
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              n_mammals = n())
```

Swap "order" for "treatment condition" and "sleep" for "expression" and you have a
standard table from a genomics paper, produced with the same three lines.

## Exercises

For each exercise, with ggplot2 and dplyr loaded, try writing the pipe yourself
before opening the solution.

1.  **The right two verbs.** You want only the mammals that sleep *fewer than* 5
    hours, showing just `name` and `sleep_total`. Which two verbs, and in what
    order?

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    msleep |>
        filter(sleep_total < 5) |>
        select(name, sleep_total)
    ```
    `filter()` chooses the rows; `select()` chooses the columns. Either order
    returns the same answer, but filtering first means `select()` has fewer rows
    to carry — the habit that pays off on large tables.
    :::

2.  **Sleepiest carnivores.** Keep only the carnivores (`vore == "carni"`), then
    sort them from most to least total sleep, showing `name`, `vore`, and
    `sleep_total`.

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    msleep |>
        filter(vore == "carni") |>
        select(name, vore, sleep_total) |>
        arrange(desc(sleep_total))
    ```
    `filter()` selects the rows, `select()` trims the columns, and
    `arrange(desc(...))` sorts largest-first. Reading the pipe top to bottom
    tells the whole story.
    :::

3.  **A derived column.** Add a `rem_fraction` column (REM sleep divided by total
    sleep), then sort so the mammals with the *highest* REM fraction come first,
    and look at their `bodywt`. Do the heaviest animals sit at the top?

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    msleep |>
        mutate(rem_fraction = sleep_rem / sleep_total) |>
        select(name, bodywt, rem_fraction) |>
        arrange(desc(rem_fraction)) |>
        head()
    ```
    `mutate()` builds the new column and `arrange(desc(...))` ranks by it. The
    heaviest mammals are *not* at the top, hinting that large-bodied animals spend
    a smaller fraction of their sleep in REM — a pattern worth a proper plot to
    confirm.
    :::

4.  **Summarize by group.** For each feeding type (`vore`), compute the average
    total sleep and the number of species. Which feeding type sleeps the most on
    average?

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    msleep |>
        group_by(vore) |>
        summarise(avg_sleep = mean(sleep_total),
                  n = n())
    ```
    Split-apply-combine: split by `vore`, apply `mean()` and `n()` to each group,
    combine into one row per group. The `NA` row gathers mammals whose feeding
    type is unrecorded — a standing reminder to check for missing values in real
    data.
    :::

## Summary

You now have the core of dplyr, and every verb maps onto something you'll do to a
biological table:

- `select()` keeps the columns you want (e.g. `sample_id` and `condition` from a
  sample sheet).
- `filter()` keeps the rows that pass a test (e.g. genes below a p-value cutoff).
- `arrange()` sorts rows, with `desc()` for largest-first.
- `mutate()` adds derived columns (e.g. log fold changes).
- `summarise()` reduces a column to a summary statistic.
- `group_by()` turns `summarise()` into one-summary-per-group — the
  split-apply-combine pattern at the heart of countless analyses.
- The pipe, `|>`, threads these verbs so an analysis reads top to bottom as a
  sequence of "and then" steps.

Master these six verbs on `msleep` and they carry over, unchanged, to count
matrices, sample sheets, and results tables.

[^msleep]: A quantitative, theoretical framework for understanding mammalian sleep. Van M. Savage, Geoffrey B. West. Proceedings of the National Academy of Sciences Jan 2007, 104 (3) 1051-1056; DOI: [10.1073/pnas.0610080104](https://doi.org/10.1073/pnas.0610080104)
