---
title: 'Introduction to dplyr: mammal sleep dataset'
author: "Stephen Turner (with modifications by Sean Davis)"
---

Almost every analysis you'll do in biology starts with a table: a sample sheet
with one row per sample, a count matrix with one row per gene, a results table
with one row per differentially expressed transcript. Over and over you'll want
to do the same handful of things to those tables — keep only the rows that pass a
filter, pull out a few columns, sort by a statistic, add a derived column, or
collapse many rows into a per-group summary.

That recurring pattern has a name: **split-apply-combine**. You *split* a table
into groups (say, by treatment condition), *apply* a calculation to each group
(say, the mean expression), and *combine* the answers back into one tidy table.
The **dplyr** package gives you a small, consistent vocabulary of verbs for
exactly these moves, and they read almost like English.

We'll learn those verbs on a gentle, friendly dataset — the sleep habits of
mammals — before you turn them loose on genomics tables. The dataset is small
enough to see the whole thing, so you can watch each verb do its job. Every move
you make here (filtering rows, selecting columns, summarizing by group) is the
same move you'll later make on a count matrix or a sample sheet.

## What you'll learn

- Filter rows of a table with `filter()` and select columns with `select()`.
- Re-order rows with `arrange()` and add derived columns with `mutate()`.
- Collapse a table to summary statistics with `summarise()`.
- Use `group_by()` to compute those summaries *per group* — the split-apply-combine pattern.
- Chain verbs together with the pipe operator, `|>`, to read an analysis left to right.

::: {.callout-tip}
## Why a sleep dataset for a genomics book?

`msleep` is a warm-up, chosen because it is small and the columns are easy to
reason about. Keep the *biological* payoff in view as you go: filtering mammals
that sleep more than 16 hours is the same operation as filtering genes with an
adjusted p-value below 0.05; selecting the `name` and `sleep_total` columns is
the same as pulling `sample_id` and `condition` out of a sample sheet; and
summarizing sleep by taxonomic order is the same as summarizing mean expression
by treatment group. Learn the verbs here; reuse them on your own data.
:::

## What is dplyr?

The *dplyr* package is a specialized package for working with `data.frame`s (and
the closely related `tibble`) to transform and summarize tabular data with rows
and columns. For another explanation, see the package vignette,
[Introduction to dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html).

## Why is dplyr useful?

dplyr provides a set of functions — commonly called the dplyr "verbs" — that
perform the common data manipulations: filtering rows, selecting columns,
re-ordering rows, adding new columns, and summarizing data. It also makes the
split-apply-combine pattern straightforward through grouping.

Compared with base R functions, the dplyr verbs are often easier to work with.
They are more consistent in their syntax and are designed around data frames
rather than individual vectors, which is exactly the shape biological data
usually arrives in.

## Data: mammals sleep

The `msleep` (mammals sleep) dataset is an updated and expanded version of the
mammals sleep dataset, with sleep times and weights taken from Savage and
West[^msleep]. It contains the sleep times and weights for a set of mammals, with
83 rows and 11 variables. It ships as a dataset inside the *ggplot2* package, so
there is nothing to download — you just install ggplot2 once if you don't already
have it.

```{r eval=FALSE}
install.packages("ggplot2")
```

Then load the library, which makes `msleep` available.

```{r}
library(ggplot2)
data(msleep)
```

As with many datasets in R, a help page describes the dataset itself.

```{r eval=FALSE}
?msleep
```

The columns are described on that help page, and are listed here for convenience.

column name | Description
--- | ---
name | common name
genus | taxonomic rank
vore | carnivore, omnivore or herbivore?
order | taxonomic rank
conservation | the conservation status of the mammal
sleep_total | total amount of sleep, in hours
sleep_rem | rem sleep, in hours
sleep_cycle | length of sleep cycle, in hours
awake | amount of time spent awake, in hours
brainwt | brain weight in kilograms
bodywt | body weight in kilograms

## The dplyr verbs

dplyr has many functions, but the six verbs below do the bulk of everyday work.
The rest of this chapter takes them one at a time.

dplyr verbs | Description
--- | ---
`select()` | select columns
`filter()` | filter rows
`arrange()` | re-order or arrange rows
`mutate()` | create new columns
`summarise()` | summarise values
`group_by()` | allows for group operations in the "split-apply-combine" concept

Before going further, install dplyr (once) and load it.

```{r eval=FALSE}
install.packages("dplyr")
```

```{r}
library(dplyr)
```

## Selecting columns with `select()`

`select()` keeps only the columns you name. Here we keep the `name` and
`sleep_total` columns — think of pulling just the sample identifier and one
measurement out of a wider table.

```{r}
sleepData <- select(msleep, name, sleep_total)
head(sleepData)
```

To keep all columns *except* one, put a minus sign in front of it (negative
indexing). For example, every column except `name`:

```{r}
head(select(msleep, -name))
```

To select a range of columns by name, use the `:` operator. Notice that dplyr
lets you write the column names without quotes, almost as if they were variables:

```{r}
head(select(msleep, name:order))
```

To select every column whose name *starts with* "sl", use the helper
`starts_with()`:

```{r}
head(select(msleep, starts_with("sl")))
```

Other helpers let you select columns by other criteria:

1. `ends_with()` — columns that end with a character string
2. `contains()` — columns that contain a character string
3. `matches()` — columns that match a regular expression
4. `one_of()` — columns whose names are in a given group of names

## Selecting rows with `filter()`

`filter()` keeps only the rows that *match* a condition. For example, keep the
mammals that sleep a total of 16 or more hours:

```{r}
filter(msleep, sleep_total >= 16)
```

You can give `filter()` more than one condition, separated by commas; a row must
satisfy *all* of them. Here we want mammals that sleep at least 16 hours *and*
weigh more than one kilogram:

```{r}
filter(msleep, sleep_total >= 16, bodywt >= 1)
```

You can also match against a set of values with the `%in%` operator, which
returns `TRUE` for each element of the first vector that appears *in* the second.
Here we keep mammals in either the Perissodactyla or Primates taxonomic order:

```{r}
filter(msleep, order %in% c("Perissodactyla", "Primates"))
```

Any of the comparison operators (`>`, `<`, `>=`, `<=`, `!=`, `%in%`) can form the
logical test. This is exactly how you would later keep only the genes passing a
significance cutoff.

## Chaining verbs with the pipe, `|>`

Real analyses string several of these operations together. The pipe operator,
`|>`, lets you "pipe" the output of one function straight into the first argument
of the next. There is nothing magical about it — but it lets you read a chain of
operations from left to right, in the order they happen.

::: {.callout-tip}
## Read the pipe as "and then"

`x |> f() |> g()` means: take `x`, **and then** apply `f`, **and then** apply
`g`. Without the pipe you'd either save each step in a temporary variable and
pass it along, or *nest* the calls inside out — `g(f(x))` — which is hard to read
once the chain grows. The pipe keeps the steps in the order you think about them.
:::

Here is an example we have already written the nested way:

```{r}
head(select(msleep, name, sleep_total))
```

With the pipe, you read the same thing top to bottom: take `msleep`, **and then**
select two columns (`name` and `sleep_total`), **and then** show the head of the
result.

```{r}
msleep |>
    select(name, sleep_total) |>
    head()
```

You'll see how much this helps once we start combining many verbs. From here on,
we'll use the pipe throughout.

## Re-ordering rows with `arrange()`

`arrange()` sorts the rows by one or more columns. To sort by the taxonomic
order, name the column:

```{r}
msleep |> arrange(order) |> head()
```

Now let's combine verbs: select three columns, sort the rows first by taxonomic
order and then by `sleep_total`, and finally show the head of the result.

```{r}
msleep |>
    select(name, order, sleep_total) |>
    arrange(order, sleep_total) |>
    head()
```

The same chain, but instead of showing the head we filter to the mammals that
sleep 16 or more hours:

```{r}
msleep |>
    select(name, order, sleep_total) |>
    arrange(order, sleep_total) |>
    filter(sleep_total >= 16)
```

To sort a column in *descending* order, wrap it in `desc()`. Here we sort by
order ascending, then by `sleep_total` descending:

```{r}
msleep |>
    select(name, order, sleep_total) |>
    arrange(order, desc(sleep_total)) |>
    filter(sleep_total >= 16)
```

## Creating new columns with `mutate()`

`mutate()` adds new columns computed from existing ones. Here we add a column
`rem_proportion`, the ratio of REM sleep to total sleep:

```{r}
msleep |>
    mutate(rem_proportion = sleep_rem / sleep_total) |>
    head()
```

You can add several columns at once, separated by commas. Here we add
`rem_proportion` and also `bodywt_grams`, the body weight expressed in grams:

```{r}
msleep |>
    mutate(rem_proportion = sleep_rem / sleep_total,
           bodywt_grams = bodywt * 1000) |>
    head()
```

Computing a derived column like this is the same move as turning raw counts into
counts-per-million, or a fold change into a log fold change.

## Summarizing with `summarise()`

`summarise()` collapses a column down to a single summary statistic. To get the
average amount of sleep, apply `mean()` to `sleep_total` and name the result
`avg_sleep`:

```{r}
msleep |>
    summarise(avg_sleep = mean(sleep_total))
```

Mammals in this dataset sleep about ten and a half hours a day, on average. You
can compute several summaries at once. Besides `mean()`, useful summary functions
include `sd()`, `min()`, `max()`, `median()`, `sum()`, `n()` (the number of
rows), `first()`, `last()`, and `n_distinct()` (the number of distinct values):

```{r}
msleep |>
    summarise(avg_sleep = mean(sleep_total),
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              total = n())
```

That single row tells you the average, the extremes, and the sample size all at
once.

## Grouping with `group_by()`

`group_by()` is what makes `summarise()` powerful. On its own, `summarise()`
collapses the whole table to one row. Combined with `group_by()`, it gives you
one summary row *per group* — this is split-apply-combine in action.

::: {.callout-note}
## Split-apply-combine

Picture the table sorted into piles, one pile per group (here, one per taxonomic
order). dplyr **splits** the data into those piles, **applies** your summary to
each pile independently, and **combines** the per-pile answers into one tidy
result table. It's the same idea as base R's `aggregate()`, but it fits naturally
into a dplyr pipe.
:::

Let's split `msleep` by taxonomic order, then ask for the same summary
statistics as before. We expect one row of statistics per order:

```{r}
msleep |>
    group_by(order) |>
    summarise(avg_sleep = mean(sleep_total),
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              total = n())
```

Each row is one taxonomic order, with its average, minimum, and maximum sleep and
the number of species. Swap "order" for "treatment condition" and "sleep" for
"expression" and you have a standard genomics summary.

## Exercises

Use the `msleep` dataset (with ggplot2 and dplyr loaded) for each of these. Try
to write the pipe before opening the solution.

1. **Pick a verb.** You want only the mammals that sleep *fewer than* 5 hours,
   showing just their `name` and `sleep_total`. Which two verbs do you need, and
   in what order? Write the pipe.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   msleep |>
       filter(sleep_total < 5) |>
       select(name, sleep_total)
   ```
   You need `filter()` to choose the rows and `select()` to choose the columns.
   Either order works here, but filtering first means `select()` operates on
   fewer rows — the habit that scales to large tables.
   :::

2. **Sleepiest carnivores.** Keep only the carnivores (`vore == "carni"`), then
   sort them from most to least total sleep. Show `name`, `vore`, and
   `sleep_total`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   msleep |>
       filter(vore == "carni") |>
       select(name, vore, sleep_total) |>
       arrange(desc(sleep_total))
   ```
   `filter()` chooses the rows, `select()` trims the columns, and
   `arrange(desc(...))` sorts largest-first. Reading the pipe top to bottom tells
   the whole story.
   :::

3. **A derived column.** Is there a relationship between `rem_proportion` (REM
   sleep divided by total sleep) and body weight? Add a `rem_proportion` column,
   then sort the table so the mammals with the *highest* REM proportion come
   first, and look at their `bodywt`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   msleep |>
       mutate(rem_proportion = sleep_rem / sleep_total) |>
       select(name, bodywt, rem_proportion) |>
       arrange(desc(rem_proportion)) |>
       head()
   ```
   `mutate()` builds the new column, then `arrange(desc(...))` ranks by it. The
   heaviest mammals are not at the top, hinting that large-bodied animals tend to
   spend a *smaller* fraction of their sleep in REM — a pattern worth a proper
   plot to confirm.
   :::

4. **Summarize by group.** For each feeding type (`vore`), compute the average
   total sleep and the number of species. Which group sleeps the most on average?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   msleep |>
       group_by(vore) |>
       summarise(avg_sleep = mean(sleep_total),
                 n = n())
   ```
   This is the split-apply-combine pattern: split by `vore`, apply `mean()` and
   `n()`, combine into one row per group. The `NA` row collects mammals whose
   feeding type is unknown — a reminder to check for missing values in real data.
   :::

## Summary

You now have the core dplyr toolkit, and every verb maps onto something you'll do
with biological tables:

- `select()` keeps the columns you want (e.g. pulling `sample_id` and `condition`
  from a sample sheet).
- `filter()` keeps the rows that pass a condition (e.g. genes below a p-value
  cutoff).
- `arrange()` sorts rows, with `desc()` for largest-first.
- `mutate()` adds derived columns (e.g. log fold changes).
- `summarise()` collapses a column to a summary statistic.
- `group_by()` turns `summarise()` into per-group summaries — the
  split-apply-combine pattern at the heart of so many analyses.
- The pipe, `|>`, chains these verbs so you can read an analysis left to right,
  as a sequence of "and then" steps.

Master these six verbs on `msleep` and you can carry them straight over to count
matrices, sample sheets, and results tables.

[^msleep]: A quantitative, theoretical framework for understanding mammalian sleep. Van M. Savage, Geoffrey B. West. Proceedings of the National Academy of Sciences Jan 2007, 104 (3) 1051-1056; DOI: [10.1073/pnas.0610080104](https://doi.org/10.1073/pnas.0610080104)
