---
title: "Case Study: Behavioral Risk Factor Surveillance System"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
  - name: Martin Morgan
    affiliation: Roswell Park Comprehensive Cancer Center<br/>Bioconductor Core Team
    affiliation-url: https://www.roswellpark.org/
---

::: {.callout-note appearance="minimal"}
Much of the material in this chapter is adapted from teaching material by Martin
Morgan.
:::

```{r style-A3, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

The Behavioral Risk Factor Surveillance System (BRFSS) is a large health survey
run every year by the U.S. Centers for Disease Control and Prevention (CDC). Each
year it phones hundreds of thousands of adults and asks about their health
behaviors, chronic conditions, and use of preventive care. Public-health
researchers lean on it to track risk factors, shape policy, and judge whether
prevention programs are working.

In this chapter you take on the role of one of those researchers. You'll work
with a small slice of the BRFSS data and walk it through a full **exploratory
data analysis (EDA)**: load it, look at it, summarize it, plot it, and ask whether
the variables relate to one another. EDA is the first thing you do with any new
dataset — before any fancy modeling — because it's how you learn the shape of your
data and catch problems early. By the end you'll go one step further and fit a
simple statistical model to a question the data can actually answer.

Everything here uses **base R** — no extra packages — so you can follow along on a
fresh R install.

## What you'll learn

- Download a dataset and load a CSV into R with `download.file()` and `read.csv()`.
- Inspect a data frame's size and structure with `head()`, `dim()`, and `summary()`.
- Visualize single variables with `hist()` and `boxplot()`, and pairs of variables
  with `plot()`.
- Measure the strength of a linear relationship with `cor()`, and handle the
  missing-value trap that makes it return `NA`.
- Compare a measurement between two groups with a `t.test()`, and fit and read a
  simple linear regression with `lm()`.

## Loading the dataset

First you need the data. You can download it from
[this link](data/BRFSS-subset.csv) by hand, or — better — let R fetch it for you
straight from the command line. The `download.file()` function takes a URL and a
local filename to save it under:

```{r}
download.file('https://raw.githubusercontent.com/seandavi/ITR/master/BRFSS-subset.csv',
              destfile = 'BRFSS-subset.csv')
```

That writes a file called `BRFSS-subset.csv` into your working directory. Now you
point R at that file and read it in with `read.csv()`, storing the result in a
data frame called `brfss`:

```{r}
path <- "BRFSS-subset.csv"
stopifnot(file.exists(path))
brfss <- read.csv(path)
```

The `stopifnot(file.exists(path))` line is a small safety check: if the file isn't
where you said it is, R stops immediately with a clear message instead of failing
in some confusing way two steps later.

::: {.callout-tip}
## Working interactively? Let R pop up a file picker

If you downloaded the file by hand and aren't sure of its exact path, you can let
R open a file-chooser dialog instead of typing the path:

```{r eval=FALSE}
path <- file.choose()    # opens a dialog; navigate to BRFSS-subset.csv
brfss <- read.csv(path)
```

That's handy when you're poking around interactively, but for a script or a
reproducible document you want a fixed path like the one above so it runs the same
way every time.
:::

## Inspecting the data

Whenever you load a dataset, your very first move should be to *look at it*. Start
with `head()`, which prints the first six rows:

```{r}
head(brfss)
```

That gives you a feel for what each column holds and what type of value it is.
Here you can see an `Age`, a `Weight`, a `Height`, a `Sex`, and a `Year`.

Next, check how big the dataset is with `dim()`, which reports the number of rows
and columns:

```{r}
dim(brfss)
```

Knowing the size matters: it tells you whether you have a handful of records or
many thousands, which shapes what analyses make sense.

## Summary statistics

Now that you know the structure, get a quick numeric overview of every column at
once with `summary()`:

```{r}
summary(brfss)
```

For each **numeric** column (`Age`, `Weight`, `Height`) this prints the minimum,
first quartile, median, mean, third quartile, and maximum. For text/categorical
columns you get a simpler count. Two things to notice right away: the `Weight` and
`Height` columns list a number of `NA`'s (missing values), and the spread between
the median and the maximum hints at the *shape* of each distribution. Both of
those observations will matter shortly.

## Visualizing single variables

Numbers in a summary table only take you so far. A picture of a single variable —
its **distribution** — tells you far more about its shape, center, and spread.
Start with a histogram of `Age` using `hist()`:

```{r}
#| label: fig-age-hist
#| fig-cap: "Distribution of respondent ages in the BRFSS subset."
hist(brfss$Age, main = "Age Distribution",
     xlab = "Age", col = "lightblue")
```

Each bar counts how many respondents fall in that age range. You can read the
center, the spread, and whether the distribution is roughly symmetric or lopsided
straight off the picture.

::: {.callout-tip}
## A function has many options — read its help page

`hist()` has many options: the number of bins, the bar color, the title, the axis
labels, and more. You can pull up the full list any time by typing `?hist` in the
R console.

This is a habit worth building for *every* function you use. The documentation for
any function `f` is one keystroke away with `?f` or `help("f")`, and it's the
fastest way to learn what a function can do.
:::

Next, compare a variable *across groups*. A boxplot is perfect for this. Here you
compare the `Weight` distribution between the two sexes. The formula
`Weight ~ Sex` reads as "weight broken down by sex":

```{r}
#| label: fig-weight-by-sex
#| fig-cap: "Weight distribution compared between the sexes."
boxplot(brfss$Weight ~ brfss$Sex, main = "Weight Distribution by Sex",
        xlab = "Sex", ylab = "Weight", col = c("pink", "lightblue"))
```

Each box spans the middle 50% of weights, the thick line is the median, and the
whiskers reach out toward the extremes. At a glance you can see whether one group
tends to weigh more than the other and whether the spreads differ.

## Analyzing relationships between variables

So far you've looked at variables one at a time. EDA gets really interesting when
you ask whether *two* variables move together. Start with a scatterplot of age
against weight using `plot()`:

```{r}
#| label: fig-age-weight-scatter
#| fig-cap: "Scatterplot of age versus weight."
plot(brfss$Age, brfss$Weight, main = "Scatterplot of Age and Weight",
     xlab = "Age", ylab = "Weight", col = "darkblue")
```

Each point is one respondent. A cloud with no obvious slope means little
relationship; a tilted, tightening band means the two variables tend to rise or
fall together.

To put a single number on that, compute the **correlation coefficient** with
`cor()`. It runs from −1 (perfect negative relationship) through 0 (none) to +1
(perfect positive relationship):

```{r}
cor(brfss$Age, brfss$Weight)
```

That's not a number — it's `NA`. This is one of the most common surprises in R,
so it's worth pausing on.

::: {.callout-warning}
## Missing values make `cor()` return `NA`

Remember those `NA`'s the `summary()` flagged in the `Weight` column? By default
`cor()` refuses to guess what a missing value should be, so if *any* pair has a
missing value the whole result comes back `NA`. A quick look at `help("cor")`
points to the fix: the `use` argument. Setting `use = "complete.obs"` tells `cor()`
to drop the rows with missing values and compute the correlation on the rest.
:::

```{r}
cor(brfss$Age, brfss$Weight, use = "complete.obs")
```

Now you get a real number. It's small and positive, telling you age and weight are
only weakly related in this dataset.

## Comparing two groups: a t-test

EDA naturally raises questions that a picture alone can't settle. Here's one: the
BRFSS subset spans two survey years, 1990 and 2010. **Did the average weight of
female respondents change between those years?**

R read the `Year` column as a plain integer, but for grouping it's really a
**category** with two levels. Convert it to a factor so R treats it that way:

```{r}
brfss$Year <- factor(brfss$Year)
```

Now pull out just the female respondents into their own data frame so you can
compare them across years:

```{r}
brfssFemale <- brfss[brfss$Sex == "Female", ]
summary(brfssFemale)
```

A boxplot of weight by year shows the comparison at a glance. The formula
`Weight ~ Year` again reads "weight broken down by year":

```{r}
#| label: fig-female-weight-year
#| fig-cap: "Female respondent weight in 1990 versus 2010."
plot(Weight ~ Year, brfssFemale)
```

The 2010 box looks shifted up from the 1990 box — average weight appears higher in
the later year. But "appears higher" isn't an answer. To decide whether that gap is
larger than you'd expect from chance alone, use a **two-sample t-test**. (We cover
the t-test in depth in @sec-ttest; here we just put it to work.)

```{r}
t.test(Weight ~ Year, brfssFemale)
```

Read the output from the bottom up. The two group means show 2010 females weigh
more on average than 1990 females. The **p-value** is tiny — far below the usual
0.05 threshold — so the difference is *statistically significant*: a gap this large
is very unlikely to be a fluke of sampling. The **confidence interval** for the
difference doesn't include zero, telling the same story. You can now say, with
evidence, that average female weight was higher in 2010 than in 1990.

## Fitting a model: weight and height

A t-test compares groups. A different kind of question asks how one *continuous*
variable depends on another: **for 2010 males, how does weight change with
height?** That's a job for **linear regression**.

First, subset to 2010 males with `subset()`, then look at the pieces:

```{r}
brfss2010Male <- subset(brfss, Year == 2010 & Sex == "Male")
summary(brfss2010Male)
```

Look at each variable on its own first, then look at the relationship:

```{r}
#| label: fig-male-weight-hist
#| fig-cap: "Weight distribution for 2010 male respondents."
hist(brfss2010Male$Weight)
```

```{r}
#| label: fig-male-height-hist
#| fig-cap: "Height distribution for 2010 male respondents."
hist(brfss2010Male$Height)
```

```{r}
#| label: fig-male-weight-height
#| fig-cap: "Weight versus height for 2010 male respondents."
plot(Weight ~ Height, brfss2010Male)
```

The scatterplot shows the expected upward trend: taller men tend to weigh more.
Now quantify that trend by fitting a straight line through the cloud with `lm()`
("linear model"). The formula `Weight ~ Height` reads "model weight as a function
of height":

```{r}
fit <- lm(Weight ~ Height, brfss2010Male)
fit
```

The printout gives two **coefficients**. The `(Intercept)` is where the line
crosses zero height (not physically meaningful here, just where the math anchors
the line). The `Height` coefficient is the **slope**: for each additional
centimeter of height, predicted weight goes up by that many kilograms. That single
number is the relationship, summarized.

You can lay out the same fit as an **ANOVA table**, which partitions the
variability in weight and tests whether height explains a significant share of it:

```{r}
anova(fit)
```

The small p-value on the `Height` row confirms what the scatterplot suggested:
height explains a statistically significant amount of the variation in weight.

To *see* the fitted model, redraw the scatterplot and superimpose the regression
line with `abline()`, which knows how to read the line straight out of a fitted
model object:

```{r}
#| label: fig-male-fit-line
#| fig-cap: "The fitted regression line through the weight-versus-height data, with one point marked in red."
plot(Weight ~ Height, brfss2010Male)
abline(fit, col = "blue", lwd = 2)
# Substitute your own weight and height to see where you land...
points(73 * 2.54, 178 / 2.2, col = "red", cex = 4, pch = 20)
```

The blue line is the model's best guess at average weight for any given height. The
red dot is one person's height and weight (converted to centimeters and kilograms);
swap in your own numbers to see where you'd fall relative to the trend.

::: {.callout-note}
## Reading the model's diagnostics

A fitted model is a *noun*; the things you can do to it are *verbs* (R calls them
**methods**). When you call `plot()` on a linear model, R quietly routes it to a
special version, `plot.lm()`, that draws diagnostic plots instead of a scatterplot.
The first one, **Residuals vs Fitted**, is the most useful: it plots the model's
errors against its predictions. You want a shapeless cloud centered on zero — any
strong pattern (a curve, a funnel) is a sign the straight-line model is missing
something.

```{r}
#| label: fig-male-diagnostics
#| fig-cap: "Residuals-versus-fitted diagnostic plot for the weight-versus-height model."
plot(fit, which = 1)
```

You can see the full set of diagnostics with `plot(fit)` (it draws four), and read
about them with `?plot.lm`.
:::

## Exercises

1.  What is the mean weight in this dataset? How about the median? What is the
    difference between the two, and what does that tell you about the shape of the
    weight distribution?

    ```{r}
    #| code-fold: true
    #| code-summary: Show answer
    mean(brfss$Weight, na.rm = TRUE)
    median(brfss$Weight, na.rm = TRUE)
    mean(brfss$Weight, na.rm = TRUE) - median(brfss$Weight, na.rm = TRUE)
    ```

    The mean sits above the median, which is the signature of a distribution with a
    long right tail — a few very heavy individuals pull the mean up. (Note the
    `na.rm = TRUE`: just like `cor()`, these functions need to be told to ignore the
    missing values.)

2.  Using what you just found about the mean and median, draw a histogram of the
    weight distribution with `hist()`. How would you describe its shape?

    ```{r}
    #| code-fold: true
    #| code-summary: Show answer
    hist(brfss$Weight, xlab = "Weight (kg)", breaks = 30)
    ```

    The histogram is **right-skewed**: most respondents cluster at lower weights
    with a tail stretching toward the high end — exactly what the mean-above-median
    result predicted.

3.  Use `plot()` to examine the relationship between height and weight.

    ```{r}
    #| code-fold: true
    #| code-summary: Show answer
    plot(brfss$Height, brfss$Weight)
    ```

    Taller people tend to be heavier, so the points drift upward to the right.

4.  What is the correlation between height and weight? What does it tell you about
    their relationship?

    ```{r}
    #| code-fold: true
    #| code-summary: Show answer
    cor(brfss$Height, brfss$Weight, use = "complete.obs")
    ```

    A moderate-to-strong positive correlation, confirming numerically what the
    scatterplot showed. (Don't forget `use = "complete.obs"` to skip the missing
    values.)

5.  Draw a histogram of the height distribution. How would you describe its shape
    compared to weight?

    ```{r}
    #| code-fold: true
    #| code-summary: Show answer
    hist(brfss$Height, xlab = "Height (cm)", breaks = 30)
    ```

    Height is far more symmetric than weight — closer to the bell-shaped normal
    curve, with no strong tail to either side.

## Summary

You've now run a complete exploratory data analysis on a real public-health
dataset, end to end:

- You **loaded** the data with `download.file()` and `read.csv()`, and guarded the
  load with `stopifnot(file.exists(path))`.
- You **inspected** it with `head()`, `dim()`, and `summary()`, which flagged the
  missing values that mattered later.
- You **visualized** single variables with `hist()` and `boxplot()`, and pairs of
  variables with `plot()`.
- You **measured** a relationship with `cor()` — and learned that missing values
  force you to add `use = "complete.obs"`.
- You went **beyond description** to answer real questions: a `t.test()` showed
  female weight rose between 1990 and 2010, and an `lm()` quantified how male weight
  rises with height.

EDA is the beginning of the data-analysis process, not the end — but it's the part
that keeps every later step honest. Knowing the shape of your data, its missing
values, and its obvious relationships is what lets you choose the right model and
trust the answer it gives. For more on the t-test you used here, see @sec-ttest.
