---
title: "Factors"
author: "Sean Davis & Lori Kern"
---

Open almost any biological dataset and you will meet columns that are not
numbers but *labels*: a sample is **treated** or **control**, a mouse is
**wild-type**, **heterozygous**, or **homozygous**, a tumor is stage **I**,
**II**, **III**, or **IV**. These categorical variables are the backbone of
experiments — they say which group each sample belongs to. R has a special tool
built exactly for them: the **factor**.

You could store these labels as ordinary text, and R would let you. But the
moment you want to count how many samples are in each group, run an analysis of
variance, or fit a model that compares treatment to control, R needs to know
that "treated" and "control" are *categories* — a fixed set of possible values —
not free-form strings. A factor is how you tell it that. This chapter shows you
what factors are, how they behave, and the one quirk that trips up nearly every
beginner.

## What you'll learn

- Create a factor from a character vector with `factor()`.
- Explain how a factor stores its data as integer codes plus a lookup table of
  **levels**.
- Use `table()` to count how many observations fall in each category.
- Control the set of levels, and understand how missing values arise.
- Avoid the classic `as.numeric()` trap that returns codes instead of labels.

## A vector with a built-in dictionary

Imagine you are recording the country of origin for nine samples in a study. The
raw data is just text:

```{r}
# country of origin for nine samples
citizen <- c("uk", "us", "no", "au", "uk", "us", "us", "no", "au")
citizen
```

That is a plain character vector — nine strings, with the quotation marks R
always uses to show text. Now turn it into a factor:

```{r}
citizen_f <- factor(citizen)
citizen_f
```

Two things changed. The values printed *without* quotation marks, and a new line
appeared: `Levels: au no uk us`. That `Levels:` line is the heart of the whole
idea.

::: {.callout-important}
## The mental model: codes plus a lookup table

A factor *looks* like text but is stored as **integers** behind a **lookup
table**. Think of a coat check at a theater: you hand over your coat and get a
numbered ticket. The tickets (1, 2, 3, …) are small and easy to handle; the
coats themselves stay on a rack, each hanging at the number on its ticket.

A factor works the same way. The **levels** are the rack — the unique categories,
stored once, in alphabetical order by default (`au`, `no`, `uk`, `us`). Each
element of your data is just a *ticket number* pointing at one of those levels.
So `"uk"` is not stored nine times; it is stored as the integer `3`, and R looks
up level 3 to show you `uk`.
:::

We can pull back the curtain and see the tickets directly. Every R object can
carry hidden **attributes**, and a factor's class and levels live there:

```{r}
attributes(citizen_f)
class(citizen_f)
```

The `attributes()` call shows the `levels` (the rack) and confirms the object's
`class` is `"factor"`. If we strip that class away with `unclass()`, the
disguise drops and the raw ticket numbers appear:

```{r}
unclass(citizen_f)
```

Read that output against the levels `au no uk us`. The first sample was `"uk"`,
which is the **3rd** level, so its code is `3`. The third sample was `"no"`, the
**2nd** level, so its code is `2`. Each number is simply a pointer into the
lookup table.

## Counting categories with `table()`

The single most useful thing you will do with a factor is count how many samples
fall in each category. That is exactly what `table()` does:

```{r}
table(citizen_f)
```

At a glance you can see the shape of your sample set: two samples each from
Australia and Norway, two from the UK, and three from the US. For a real
experiment this is how you check that your treatment and control groups are
balanced before you analyze anything.

## Converting a factor back

You will often need to turn a factor back into something else — and here is where
people get burned. There are two ways to convert, and they give very different
answers.

To recover the original labels, use `as.character()`:

```{r}
as.character(citizen_f)
```

That gives back the text you started with — `"uk"`, `"us"`, and so on. But watch
what happens if you reach for `as.numeric()` instead:

```{r}
as.numeric(citizen_f)
```

::: {.callout-warning}
## The classic gotcha: `as.numeric()` returns the codes, not the labels

`as.numeric()` on a factor returns the **ticket numbers**, not the values. You
get `3 4 2 1 3 4 4 2 1` — the integer codes — *not* the countries, and certainly
not any number the labels might have looked like.

This bites hardest when a factor's labels are themselves numbers. Suppose a dose
column reads `"10"`, `"20"`, `"50"` but was accidentally stored as a factor.
`as.numeric()` would return `1, 2, 3` (the codes), not `10, 20, 50` (the doses) —
a silent, dangerous error. The safe recipe when labels are numeric is to go
through character first: `as.numeric(as.character(dose_f))`.
:::

## Controlling the levels

By default, the levels are every unique value in your data, sorted
alphabetically. But you are allowed to specify the levels yourself — and this is
genuinely useful. Maybe you only care about the US and UK samples for one
analysis:

```{r}
citizen_f2 <- factor(citizen, levels = c("us", "uk"))
citizen_f2
```

Look closely: the values that were not in your chosen levels — the `no` and `au`
samples — turned into `<NA>`, R's marker for a missing value. By naming only `us`
and `uk` as legal levels, you told R that every other value has no category to
belong to, so it became missing.

```{r}
table(citizen_f2)
```

By default, `table()` quietly drops those missing values, so the count reflects
only the four samples that matched. If you *want* to see the missing ones, add
them as an explicit level with `addNA()`:

```{r}
table(addNA(citizen_f2))
```

Now the `<NA>` category appears with its own count of 5 — the five samples that
fell outside your chosen levels.

::: {.callout-note}
## Why specifying levels matters

Setting the levels yourself does more than filter. It also fixes their **order**,
which controls how categories appear in tables and plots and which group a model
treats as the baseline. For a treatment study you might write
`factor(group, levels = c("control", "treated"))` so that "control" is the
reference everything else is compared against. The default settings are not
always what your analysis needs — knowing what they are, and overriding them on
purpose, is part of doing the analysis correctly.
:::

## Exercises

1. **Genotypes.** A small genotyping study records these calls for eight mice:
   `"wt"`, `"het"`, `"hom"`, `"wt"`, `"wt"`, `"het"`, `"hom"`, `"wt"`. Make a
   factor from this vector and use `table()` to count how many mice carry each
   genotype.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   genotype <- c("wt", "het", "hom", "wt", "wt", "het", "hom", "wt")
   genotype_f <- factor(genotype)
   table(genotype_f)
   ```
   `table()` counts each category for you: four `wt`, two `het`, and two `hom`.
   The factor turned a list of labels into a tidy summary of the sample set.
   :::

2. **Read the codes.** Using the `genotype_f` factor from Exercise 1, predict
   what `unclass(genotype_f)` will print *before* you run it. Then run it and
   check.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   unclass(genotype_f)
   ```
   The levels are alphabetical — `het`, `hom`, `wt` — so `het` is code `1`, `hom`
   is `2`, and `wt` is `3`. The first mouse (`wt`) prints as `3`. Each integer is
   a ticket pointing into the levels table, not a measurement.
   :::

3. **The numeric trap.** A colleague stored a dose column as a factor with labels
   `"5"`, `"10"`, `"20"` and ran `as.numeric()` on it to get the doses back. Why
   are their numbers wrong, and what should they do instead?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   dose_f <- factor(c("5", "10", "20", "5", "20"))
   as.numeric(dose_f)               # WRONG: returns the integer codes
   as.numeric(as.character(dose_f)) # RIGHT: recovers the real doses
   ```
   `as.numeric()` returns the level codes, and because levels are sorted, `"10"`
   sorts before `"5"` — so the codes do not even match the order you expect.
   Converting to character first turns each label back into its text, which
   `as.numeric()` can then read as the true number.
   :::

4. **Set the reference group.** You have a `group` vector of `"treated"` and
   `"control"` labels. Build a factor whose levels are ordered so that `control`
   comes first (the baseline a model would compare against).

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   group <- c("treated", "control", "treated", "control", "treated")
   group_f <- factor(group, levels = c("control", "treated"))
   group_f
   ```
   By naming `control` first in `levels`, you make it level 1 — the reference
   category. Most modeling functions in R compare every other group against the
   first level, so setting it deliberately keeps your results interpretable.
   :::

## Summary

You can now recognize and build the categorical variables at the center of nearly
every biological experiment:

- **Create** a factor from a character vector with `factor()`, and recognize it
  by the `Levels:` line and the missing quotation marks.
- **Picture** it as codes plus a lookup table — integer tickets pointing at a rack
  of levels — which explains every behavior in this chapter.
- **Count** categories with `table()` to check group sizes and spot imbalance.
- **Control** the levels to filter categories, set a reference group, and decide
  how missing values are handled.
- **Convert** safely: `as.character()` gives the labels back, while
  `as.numeric()` gives the codes — so reach for `as.numeric(as.character(x))`
  whenever a factor's labels are really numbers.

Keep the coat-check picture in mind and factors stop being mysterious: a factor
is just a vector of tickets with a table of what each ticket stands for.
