# Genomic Ranges Introduction

## Introduction

Genomic ranges are fundamental data structures in bioinformatics that represent intervals on chromosomes. They are essential for analyzing ChIP-seq peaks, gene annotations, regulatory elements, and other genomic features. In this tutorial, we'll explore the BED file format and demonstrate practical genomic range operations using R's rtracklayer package.

This tutorial will cover:

- Understanding the BED file format
- Loading genomic ranges from BED files
- Basic exploration of genomic ranges
- Accessing and manipulating genomic coordinates

## The dataset

We'll use CTCF ChIP-seq peak data from the ENCODE project. CTCF (CCCTC-binding factor) is a key architectural protein that helps organize chromatin structure. The data is available in BED format, which we will load and analyze. Each peak represents the results of a ChIP-seq experiment, indicating regions where CTCF binds to DNA. The sample was sequenced, aligned, and peaks were called using standard ChIP-seq analysis pipelines. The peaks are stored in a BED file, which we will import into R for analysis.

## The BED File Format

The Browser Extensible Data (BED) format is a standard way to represent genomic intervals. BED files are tab-delimited text files where each line represents a genomic feature.

### BED Format Structure

The BED format has several required and optional fields:

**Required fields (BED3):**

- **chrom**: Chromosome name (e.g., chr1, chr2, chrX)
- **chromStart**: Start position (0-based, inclusive)
- **chromEnd**: End position (0-based, exclusive)

**Optional fields:**

- **name**: Feature name/identifier
- **score**: Score (0-1000)
- **strand**: Strand (+ or - OR '*')
- **thickStart**: Start of thick drawing
- **thickEnd**: End of thick drawing
- **itemRgb**: RGB color values
- **blockCount**: Number of blocks
- **blockSizes**: Block sizes
- **blockStarts**: Block start positions

### Key Concepts

- **0-based coordinate system**: BED uses 0-based coordinates where the first base is position 0
- **Half-open intervals**: chromStart is inclusive, chromEnd is exclusive
- **Width calculation**: Width = chromEnd - chromStart

Example BED entry:
```
chr1    1000    2000    peak1    500    +
```
This represents a feature named "peak1" on chromosome 1, from position 1000 to 1999 (width = 1000 bp).

## Loading Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(rtracklayer)
library(GenomicRanges)
library(ggplot2)
library(dplyr)

# Set up plotting theme
theme_set(theme_minimal())
```


## Loading CTCF ChIP-seq Data

We'll work with CTCF ChIP-seq peak data from the ENCODE project. CTCF (CCCTC-binding factor) is a key architectural protein that helps organize chromatin structure. 

```{r load-data}
# URL for the CTCF ChIP-seq BED file
bed_url <- "https://www.encodeproject.org/files/ENCFF960ZGP/@@download/ENCFF960ZGP.bed.gz"

# Let's first load this file using readr::read_table to check its structure
ctcf_peaks_raw <- readr::read_table(bed_url, col_names=FALSE)

# Display the first few rows of the raw data
head(ctcf_peaks_raw)
```

Bioconductor's `rtracklayer` package provides a convenient way to import BED files directly into R as `GRanges` objects, which are optimized for genomic range operations. We'll use this package to load the CTCF peaks data.


```{r}
# Load the BED file using rtracklayer
ctcf_peaks <- rtracklayer::import(bed_url, format="narrowPeak")
```

Let's take a look at the loaded CTCF peaks data. The `ctcf_peaks` object is a `GRanges` object that contains genomic ranges representing the CTCF ChIP-seq peaks.

```{r}
ctcf_peaks
```

The `ctcf_peaks` object now contains the genomic ranges of CTCF peaks, including chromosome names, start and end positions, and additional metadata such as peak scores and strand information. 

## Understanding GRanges Objects

To get information from a `GRanges` object, we can use various accessor functions. For example, `seqnames()` retrieves the chromosome names, `start()` and `end()` get the start and end positions, and `width()` calculates the width of each peak.


```{r granges-accessors}
# Accessing chromosome names, start, end, and width
length(ctcf_peaks) # Total number of peaks
seqnames(ctcf_peaks) # Chromosome names (or contig names, etc.)
head(start(ctcf_peaks)) # Start positions (0-based)
head(end(ctcf_peaks)) # End positions (1-based)
head(width(ctcf_peaks)) # equivalent to end(ctcf_peaks) - start(ctcf_peaks)
```

What is the distribution of peak widths? We can visualize this using a histogram.
```{r peak-widths}
#| fig-cap: "Histogram of CTCF Peak Widths. Why is there a large peak at around 200 bp?"
# Create a histogram of peak widths
hist(width(ctcf_peaks), breaks=50, main="CTCF Peak Widths", xlab="Width (bp)", col="lightblue")
```


The "metadata" part of a `GRanges` object can be accessed using the `mcols()` function, which returns a data frame-like structure containing additional information about each genomic range.

```{r granges-metadata}
mcols(ctcf_peaks)
```

The `mcols()` function returns a data frame-like structure containing additional information about each genomic range, such as peak scores and strand orientation. This metadata can be useful for filtering or annotating peaks.

```{r mcols-cols}
mcols(ctcf_peaks)[1:5, ]
mcols(ctcf_peaks)$signalValue[1:5]  # Accessing the score column directly
hist(mcols(ctcf_peaks)$signalValue, breaks=50, main="CTCF Peak Scores", xlab="Score", col="lightblue")
```


## Exploring Peak Characteristics

### Basic Peak Statistics

How many peaks do we have? What chromosomes are represented? What is the range of peak widths? Let's calculate some basic statistics about the CTCF peaks.

```{r basic-stats}
# Calculate basic statistics
cat("=== CTCF Peak Statistics ===\n")
cat("Total number of peaks:", length(ctcf_peaks), "\n")
cat("Number of chromosomes represented:", length(unique(seqnames(ctcf_peaks))), "\n")
cat("Peak width range:", min(width(ctcf_peaks)), "-", max(width(ctcf_peaks)), "bp\n")
cat("Median peak width:", median(width(ctcf_peaks)), "bp\n")
cat("Mean peak width:", round(mean(width(ctcf_peaks)), 1), "bp\n")

# Show chromosome names
cat("\nChromosomes present:\n")
print(sort(unique(as.character(seqnames(ctcf_peaks)))))
```

### Peaks Per Chromosome

```{r peaks-per-chromosome}
# Count peaks per chromosome
peaks_per_chr <- table(seqnames(ctcf_peaks))
peaks_per_chr_df <- as.data.frame(peaks_per_chr)
peaks_per_chr_df
```


## Accessing Peak Coordinates

### Finding Starts and Ends

```{r starts-ends}
# Extract start and end coordinates
peak_starts <- start(ctcf_peaks)
peak_ends <- end(ctcf_peaks)
peak_centers <- start(ctcf_peaks) + (width(ctcf_peaks)/2)

head(peak_starts, 10)

head(peak_ends, 10)

head(peak_centers, 10)
```

## Manipulating Peak Ranges

### Shifting Peaks

Shifting peaks is useful for various analyses, such as creating flanking regions or adjusting peak positions.

```{r shifting-peaks}
# Shift peaks by different amounts
peaks_shifted_100bp <- shift(ctcf_peaks, 100)  # Shift right by 100bp
peaks_shifted_neg50bp <- shift(ctcf_peaks, -50)  # Shift left by 50bp

cat("=== Peak Shifting Examples ===\n")
cat("Original peak 1:", as.character(ctcf_peaks[1]), "\n")
cat("Shifted +100bp:", as.character(peaks_shifted_100bp[1]), "\n")
cat("Shifted -50bp:", as.character(peaks_shifted_neg50bp[1]), "\n")

# Demonstrate that width is preserved during shifting
cat("\nWidths after shifting (should be unchanged):\n")
cat("Original width:", width(ctcf_peaks[1]), "\n")
cat("Shifted +100bp width:", width(peaks_shifted_100bp[1]), "\n")
cat("Shifted -50bp width:", width(peaks_shifted_neg50bp[1]), "\n")
```

### Setting Peak Widths

Resizing peaks is common when standardizing peak sizes or creating fixed-width windows around peak centers.

```{r setting-widths}
# Resize peaks to fixed width (200bp) centered on original peak center
peaks_200bp <- resize(ctcf_peaks, width = 200, fix = "center")

# Resize peaks to 500bp, keeping the start position fixed
peaks_500bp_start <- resize(ctcf_peaks, width = 500, fix = "start")

# Resize peaks to 300bp, keeping the end position fixed
peaks_300bp_end <- resize(ctcf_peaks, width = 300, fix = "end")

cat("=== Peak Resizing Examples ===\n")
cat("Original peak 1:", as.character(ctcf_peaks[1]), "\n")
cat("Resized to 200bp (center):", as.character(peaks_200bp[1]), "\n")
cat("Resized to 500bp (start fixed):", as.character(peaks_500bp_start[1]), "\n")
cat("Resized to 300bp (end fixed):", as.character(peaks_300bp_end[1]), "\n")

# Verify that all peaks now have the specified width
cat("\nWidth verification:\n")
cat("200bp resize - all widths 200?", all(width(peaks_200bp) == 200), "\n")
cat("500bp resize - all widths 500?", all(width(peaks_500bp_start) == 500), "\n")
cat("300bp resize - all widths 300?", all(width(peaks_300bp_end) == 300), "\n")
```

### Creating Flanking Regions

```{r flanking-regions}
# Create flanking regions around peaks
upstream_1kb <- flank(ctcf_peaks, width = 1000, start = TRUE)
downstream_1kb <- flank(ctcf_peaks, width = 1000, start = FALSE)

# Create regions extending in both directions
extended_peaks <- resize(ctcf_peaks, width = width(ctcf_peaks) + 2000, fix = "center")

cat("=== Flanking Region Examples ===\n")
cat("Original peak 1:", as.character(ctcf_peaks[1]), "\n")
cat("1kb upstream:", as.character(upstream_1kb[1]), "\n")
cat("1kb downstream:", as.character(downstream_1kb[1]), "\n")
cat("Extended ±1kb:", as.character(extended_peaks[1]), "\n")
```


## Coverage

Coverage refers to the number of times a genomic region is covered by ranges on a chromosome. A common use case is calculating coverage from ChIP-seq data, where we want to know how many reads overlap with each peak or for doing "peak calling" analysis.

We can make a toy example by simulating random reads of length 50 bp across a chromosome and then calculating coverage.

```{r coverage-example}
# Simulate random reads on chromosome 1
set.seed(42)  # For reproducibility
chrom_length <- 1000000  # Length of chromosome 1
num_reads <- 100000
read_length <- 50  # Length of each read
random_starts <- sample(seq_len(chrom_length - read_length + 1), num_reads, replace = TRUE)
random_reads <- GRanges(seqnames = "chr1",
                         ranges = IRanges(start = random_starts, end = random_starts + read_length-1))
random_reads
```

We are now ready to calculate the coverage of these random reads across the chromosome. The `coverage()` function from the `GenomicRanges` package computes the coverage of ranges in a `GRanges` object.

```{r coverage-calculation}
random_coverage <- coverage(random_reads, width = chrom_length)
head(random_coverage)
```
We can visualize the coverage of the random reads on chromosome 1. The `coverage()` function returns a `Rle` object, which is a run-length encoding of the coverage values.
```{r coverage-plot}
plot(random_coverage[[1]][1:10000], main="Coverage of Random Reads on Chromosome 1", xlab="Position", ylab="Coverage", type="h", col="blue")
```

This plot shows the coverage of random reads across the first 10,000 bases of chromosome 1. The y-axis represents the number of reads covering each position, while the x-axis represents the genomic positions.

```{r coverage-density}
# Calculate coverage density
hist(as.numeric(random_coverage[[1]]), main="Coverage Density", xlab="Coverage", ylab="Frequency", col="lightblue")
```

## Key Takeaways

This tutorial demonstrated several important concepts:

1. **BED file format**: Understanding the structure and coordinate system of BED files
2. **Loading genomic data**: Using rtracklayer to import BED files into R
3. **Basic exploration**: Counting features, examining distributions, and summarizing data
4. **Coordinate manipulation**: Accessing starts, ends, and performing coordinate arithmetic
5. **Range operations**: Shifting, resizing, and creating flanking regions
6. **Advanced analysis**: Finding overlaps and performing grouped operations

### Common Use Cases

- **Peak calling analysis**: Examining ChIP-seq peaks, ATAC-seq peaks, etc.
- **Annotation overlap**: Finding genes or regulatory elements near peaks
- **Comparative analysis**: Comparing peak sets between conditions or samples
- **Motif analysis**: Creating sequences around peak centers for motif discovery
- **Visualization**: Preparing data for genome browser tracks or custom plots

### Best Practices

1. Always check coordinate systems (0-based vs 1-based)
2. Verify chromosome naming conventions match your reference genome
3. Consider peak width distributions when setting analysis parameters
4. Use appropriate genome builds for all analyses
5. Document coordinate transformations and filtering steps

This foundation in genomic ranges and BED file manipulation will serve as a basis for more advanced genomic analyses in R.
