---
title: "Genomic ranges with real ChIP-seq peaks"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

CTCF is one of the most important architectural proteins in the human genome. It
binds thousands of sites across the chromosomes, where it helps fold DNA into
loops and marks the boundaries between active and silent regions. So a natural
question for a biologist is: **where does CTCF actually bind?**

A ChIP-seq experiment answers that question by handing you a list of genomic
intervals — the *peaks* where the protein was detected. In this chapter you'll
take a real CTCF peak set from the ENCODE project, load it into R, and learn the
tools you need to explore it: how genomic coordinates are stored on disk, how to
import them, and how to ask questions like "how wide are these peaks?" and "where
exactly is the center of each one?"

This chapter focuses on the **real-data workflow** — file formats, importing, and
exploring. For the full catalog of range operations (set operations, overlaps,
nearest-feature searches, gene models), see the *Genomic ranges and features*
chapter, which builds `GRanges` objects by hand and works through those methods
systematically. Here we keep our hands on a real dataset.

## What you'll learn

- Read the BED / narrowPeak file format and explain its coordinate system.
- Import a peak file from a URL into a `GRanges` object with `rtracklayer`.
- Use the accessor functions (`seqnames()`, `start()`, `width()`, `mcols()`) to
  pull information out of a `GRanges`.
- Interpret the distribution of peak widths and what it tells you about the peak
  caller.
- Manipulate ranges with `shift()`, `resize()`, and `flank()` to build the
  windows downstream analyses need.
- Compute and visualize read coverage along a chromosome.

## The BED file format

Genomic intervals are most often stored in **BED** (Browser Extensible Data)
format: a tab-delimited text file where each line is one feature. The first three
columns are required and define the interval itself:

- **chrom** — chromosome name (e.g. `chr1`, `chrX`).
- **chromStart** — start position.
- **chromEnd** — end position.

Optional columns can follow: a feature `name`, a `score`, the `strand`, and
several more. Peak callers extend BED with extra columns (signal values,
p-values); that flavor is called **narrowPeak**, and it's what our ENCODE file
uses.

::: {.callout-warning}
## BED is 0-based; R is 1-based

This is the single most common source of off-by-one bugs in genomics, so it is
worth burning into memory. BED files use a **0-based, half-open** coordinate
system: counting starts at 0, the start is *included*, and the end is *excluded*.
R — and the `GRanges` objects you'll build from these files — use a **1-based,
fully-closed** system: counting starts at 1 and both ends are included.

Take the BED line `chr1  1000  2000`. In BED that means the 1000 bases spanning
the interval `[1000, 2000)`. When `rtracklayer` imports it, the start becomes
**1001** (shifted up by one) and the end stays **2000**, giving the same 1000
bases but now written as `1001-2000` in 1-based coordinates. The width is
identical (`2000 - 1000 = 1000`); only the *numbering* differs. The good news:
`rtracklayer` does this conversion for you. The trap is doing the arithmetic
yourself on the raw file and being off by one.
:::

## Loading the libraries

```{r setup, message=FALSE, warning=FALSE}
library(rtracklayer)
library(GenomicRanges)
library(ggplot2)
library(dplyr)

theme_set(theme_minimal())
```

## Importing the CTCF peaks

We'll use a CTCF ChIP-seq peak file from ENCODE.^[ENCODE accession
[ENCFF960ZGP](https://www.encodeproject.org/files/ENCFF960ZGP/), originally at
<https://www.encodeproject.org/files/ENCFF960ZGP/@@download/ENCFF960ZGP.bed.gz>.
We keep a copy in this book's repository and read it from the GitHub URL below,
so the code runs unchanged on any machine.] It lives at a URL, and
`rtracklayer::import()` can read straight from that URL — no need to download it
by hand first.

Before we use the convenient importer, let's peek at the raw file so you can see
that it really is just a table of numbers and text:

```{r load-data}
# The peak file lives in this book's GitHub repository; we read it straight from
# its URL so this code runs unchanged anywhere (no local checkout needed).
bed_url <- "https://raw.githubusercontent.com/seandavi/RBiocBook/main/data/ctcf_peaks_ENCFF960ZGP.bed.gz"

# Read the raw file as a plain table just to see its shape
ctcf_peaks_raw <- readr::read_table(bed_url, col_names = FALSE)
head(ctcf_peaks_raw)
```

Each row is one peak, and you can see the chromosome, start, and end in the first
three columns followed by the narrowPeak extras. Now let's import it properly into
a `GRanges` object, which is the data structure built for genomic intervals:

```{r import-peaks}
ctcf_peaks <- rtracklayer::import(bed_url, format = "narrowPeak")
ctcf_peaks
```

::: {.callout-note}
## What `format = "narrowPeak"` gives you

Telling `import()` that the file is `narrowPeak` does two helpful things. First,
it applies the 0-based-to-1-based conversion described above, so the `start`,
`end`, and `width` you see are already in R's coordinate system. Second, it parses
the peak-caller columns into named metadata: `signalValue`, `pValue`, `qValue`,
and `peak` (the offset of the summit within the peak). Those land in the metadata
columns of the `GRanges`, where you can reach them with `$`.
:::

## Looking inside a GRanges

A `GRanges` keeps the *ranges* (chromosome, start, end, strand) separate from the
*metadata* (everything else). You pull each part out with an accessor function.

```{r granges-accessors}
length(ctcf_peaks)         # how many peaks?
head(seqnames(ctcf_peaks)) # which chromosomes
head(start(ctcf_peaks))    # start positions (1-based, after import)
head(end(ctcf_peaks))      # end positions
head(width(ctcf_peaks))    # width = end - start + 1
```

So this file contains `r length(ctcf_peaks)` CTCF peaks. The widths vary, but
let's look at the whole distribution rather than the first six values:

```{r peak-widths}
#| label: fig-peak-widths
#| fig-cap: "Distribution of CTCF peak widths. The sharp spike reflects the peak caller's fixed-width output."
hist(width(ctcf_peaks), breaks = 50,
     main = "CTCF peak widths", xlab = "Width (bp)", col = "lightblue")
```

@fig-peak-widths has a striking feature: a tall spike at one particular width
rather than a smooth spread. That's not biology — it's the peak caller. Many
ChIP-seq pipelines report peaks at a **fixed width** (here around 200 bp) centered
on the detected summit, so the same number shows up over and over. When you see a
histogram like this, read it as a signature of how the data were processed, not as
a property of CTCF itself.

The metadata sits in a separate table you reach with `mcols()`:

```{r granges-metadata}
head(mcols(ctcf_peaks))
```

These are the narrowPeak columns. You can pull one out directly with `$` — for
example the `signalValue`, which measures how strong the binding signal was at
each peak:

```{r signal-hist}
#| label: fig-signal
#| fig-cap: "Distribution of CTCF peak signal values. Most peaks are weak; a long tail of strong sites stretches to the right."
hist(mcols(ctcf_peaks)$signalValue, breaks = 50,
     main = "CTCF peak signal values", xlab = "Signal value", col = "lightblue")
```

@fig-signal is **right-skewed**: most peaks carry a modest signal, with a long
tail of much stronger binding sites. That shape is typical of ChIP-seq — a handful
of high-confidence sites and many weaker ones — and it's why people often filter
on `signalValue` before downstream work.

## Some quick summaries

How many peaks, on how many chromosomes, and how wide? A few small expressions
answer this more clearly than a wall of printed text:

```{r basic-stats}
length(ctcf_peaks)                      # total peaks
length(unique(seqnames(ctcf_peaks)))    # chromosomes represented
range(width(ctcf_peaks))                # narrowest and widest peak
median(width(ctcf_peaks))               # typical peak width
```

The median width sits right at the spike we saw in @fig-peak-widths, confirming
that most peaks share the caller's fixed width. We can also count peaks per
chromosome:

```{r peaks-per-chromosome}
peaks_per_chr <- table(seqnames(ctcf_peaks))
head(as.data.frame(peaks_per_chr), 10)
```

Larger chromosomes generally carry more peaks, which makes sense: more DNA means
more potential binding sites.

## Where is each peak?

Three quantities you'll reach for constantly are the start, the end, and the
**center** of each peak. The center is especially useful — motif analyses and
fixed-width windows are usually built around it.

```{r starts-ends}
peak_starts  <- start(ctcf_peaks)
peak_ends    <- end(ctcf_peaks)
peak_centers <- start(ctcf_peaks) + width(ctcf_peaks) / 2

head(peak_centers, 5)
```

`peak_centers` now holds the midpoint of every peak, ready to anchor a window.

## Reshaping ranges

Three functions cover most of the range-reshaping you'll do on a peak set:
`shift()` slides a range, `resize()` changes its width, and `flank()` builds a
neighbouring region. They all return a new `GRanges` and leave the original
untouched.

### Shifting

`shift()` moves every range by a fixed number of bases without changing its width:

```{r shifting-peaks}
peaks_shifted <- shift(ctcf_peaks, 100)   # slide right by 100 bp

ctcf_peaks[1]
peaks_shifted[1]
```

The start and end of the first peak both moved up by 100, and the width is
unchanged — exactly what "shift" should do.

### Resizing

`resize()` sets every range to a chosen width. The `fix` argument says which point
to hold still while the width changes. Resizing all peaks to a common width is the
standard way to build comparable windows:

```{r setting-widths}
peaks_200bp <- resize(ctcf_peaks, width = 200, fix = "center")

peaks_200bp[1]
all(width(peaks_200bp) == 200)   # every peak is now exactly 200 bp
```

Fixing on `"center"` keeps each peak's midpoint in place and grows or shrinks it
symmetrically — so a downstream tool sees uniform 200 bp windows centered on the
binding sites.

### Flanking

`flank()` returns the region *beside* a range — useful for grabbing the DNA just
upstream or downstream of a peak:

```{r flanking-regions}
upstream_1kb <- flank(ctcf_peaks, width = 1000, start = TRUE)

ctcf_peaks[1]
upstream_1kb[1]
```

The flanking range is the 1000 bp sitting immediately before the start of the
first peak — the kind of window you might scan for a neighbouring promoter.

::: {.callout-note}
## Overlaps and grouped operations live next door

Once you have peaks and other features (genes, promoters, other peak sets), the
next question is usually *which ones overlap?* The `findOverlaps()` and
`nearest()` functions, set operations like `intersect()`, and grouped operations
over a `GRangesList` are covered in depth in the *Genomic ranges and features*
chapter. We don't repeat them here so we can stay focused on the real-data import
workflow.
:::

## Coverage

**Coverage** counts how many ranges overlap each position along a chromosome. It's
the foundation of peak calling: pile up sequencing reads, and the tall stacks are
where the protein bound. Let's build a toy example by scattering random reads
across a one-megabase stretch and counting the pile-up.

```{r coverage-example}
set.seed(42)   # reproducible random reads
chrom_length <- 1000000
num_reads    <- 100000
read_length  <- 50

random_starts <- sample(seq_len(chrom_length - read_length + 1),
                         num_reads, replace = TRUE)
random_reads  <- GRanges(seqnames = "chr1",
                         ranges = IRanges(start = random_starts,
                                          end = random_starts + read_length - 1))
random_reads
```

The `coverage()` function turns those reads into a per-base count. It returns an
`Rle` (run-length encoded) object — a compact way to store long stretches of
repeated values:

```{r coverage-calculation}
random_coverage <- coverage(random_reads, width = chrom_length)
head(random_coverage)
```

Let's look at the coverage across the first 10,000 bases:

```{r coverage-plot}
#| label: fig-coverage
#| fig-cap: "Per-base coverage from randomly placed reads across the first 10,000 bases of a simulated chromosome."
plot(random_coverage[[1]][1:10000], type = "h", col = "blue",
     main = "Coverage of random reads", xlab = "Position", ylab = "Coverage")
```

@fig-coverage looks like noise, and that's the point: because we placed the reads
*at random*, the coverage hovers around a constant average with no real peaks.
Real ChIP-seq data would show tall, localized spikes where the protein bound —
those spikes are what a peak caller hunts for, and their absence here is exactly
what "no signal" should look like.

We can also ask how the coverage is distributed across all positions:

```{r coverage-density}
#| label: fig-coverage-density
#| fig-cap: "How often each coverage depth occurs across the simulated chromosome."
hist(as.numeric(random_coverage[[1]]),
     main = "Coverage distribution", xlab = "Coverage", ylab = "Frequency",
     col = "lightblue")
```

@fig-coverage-density is a tidy bell-ish hump centered on the average depth. With
`r format(num_reads, big.mark = ",")` reads of length `r read_length` spread over
`r format(chrom_length, big.mark = ",")` bases, the expected coverage is about
`r round(num_reads * read_length / chrom_length, 1)`x — right where the hump sits.
Random reads give random, evenly-spread coverage; biological signal is what makes
that distribution lumpy.

## Exercises

Use the `ctcf_peaks` object you imported above. Each solution names the accessor
or function that does the job.

1. **Wide peaks.** How many CTCF peaks are wider than 250 bp?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r ex-wide, eval=FALSE}
   sum(width(ctcf_peaks) > 250)
   ```
   `width()` returns a numeric vector, `> 250` turns it into TRUE/FALSE, and
   `sum()` counts the TRUEs. Most peaks sit at the caller's fixed width, so only a
   minority exceed 250 bp.
   :::

2. **Fixed-width windows.** Build a new `GRanges` of 500 bp windows centered on
   each peak, and confirm every window is exactly 500 bp wide.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r ex-resize, eval=FALSE}
   windows_500 <- resize(ctcf_peaks, width = 500, fix = "center")
   all(width(windows_500) == 500)
   ```
   `fix = "center"` keeps each peak's midpoint fixed while the width grows to 500,
   giving uniform windows for a downstream analysis.
   :::

3. **Peak centers.** Compute the center position of every peak and report the
   median center.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r ex-centers, eval=FALSE}
   centers <- start(ctcf_peaks) + width(ctcf_peaks) / 2
   median(centers)
   ```
   The center is the start plus half the width. (You could also use
   `start(resize(ctcf_peaks, width = 1, fix = "center"))` to let `GRanges` do the
   arithmetic.)
   :::

4. **Fraction on chr1.** What fraction of all CTCF peaks fall on `chr1`?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r ex-chr1, eval=FALSE}
   mean(seqnames(ctcf_peaks) == "chr1")
   ```
   Comparing `seqnames()` to `"chr1"` gives a TRUE/FALSE vector; `mean()` of a
   logical vector is the *proportion* that are TRUE — a handy idiom for "what
   fraction satisfy this condition?"
   :::

## Summary

You can now take a real peak set from file to insight:

- **Read a BED/narrowPeak file** and explain its 0-based, half-open coordinate
  system — and why R's 1-based coordinates differ.
- **Import peaks into a `GRanges`** straight from a URL with
  `rtracklayer::import(..., format = "narrowPeak")`, which handles the coordinate
  conversion and parses the metadata columns.
- **Explore the object** with `seqnames()`, `start()`, `end()`, `width()`, and
  `mcols()`, and read its width and signal distributions as fingerprints of the
  peak caller and the biology.
- **Reshape ranges** with `shift()`, `resize()`, and `flank()` to build the
  windows downstream tools expect.
- **Compute coverage** and recognize that random reads give flat coverage, while
  biological signal makes it lumpy.

For overlaps, nearest-feature searches, set operations, and gene models, continue
to the [Genomic ranges and features](ranges_and_signals.qmd) chapter, which treats
those operations in full.
