---
title: "Accessing public omics data with GEOquery"
---

Someone has already run the experiment you need. Thousands of expression
studies — microarray and RNA-seq, across nearly every tissue and disease — sit
in a public archive called the **Gene Expression Omnibus (GEO)**, free to
download and reanalyze. Suppose you read a paper comparing tumor and normal
tissue across several cancers and you want to *see the data for yourself*:
pull it down, check how the samples group, and ask whether tumor and normal
really separate. That is exactly the kind of reuse GEO exists for.

In this chapter you'll do that end to end. You'll fetch a real published
dataset with the `GEOquery` package, load it into a `SummarizedExperiment` (the
container from the previous chapter), keep the most variable genes, and run a
**principal component analysis (PCA)** to visualize how the samples relate. The
data come from a study of immune gene expression in matched normal and tumor
tissue [@brouwer-visser_regulatory_2018], deposited as accession
[GSE103512][GSE103512].

## What you'll learn

By the end of this chapter you will be able to:

- Download a published dataset from GEO with `getGEO()`.
- Convert the result into a `SummarizedExperiment` and inspect its parts.
- Summarize sample annotations with `table()`.
- Filter an expression matrix down to its most variable genes.
- Run PCA with `prcomp()` and plot the samples to see how they group.

This chapter builds directly on the `SummarizedExperiment` chapter. If the words
`assay()`, `colData()`, or "rows are genes, columns are samples" don't feel
familiar yet, read that chapter first — here we'll lean on it rather than
re-explain the object's anatomy.

::: {.callout-tip}
## Installing GEOquery

`GEOquery` is a Bioconductor package. Install it once with `BiocManager`; you
won't need to run this again:

```{r geoquery00, eval=FALSE}
BiocManager::install("GEOquery")
```
:::

## Fetching a dataset from GEO

`GEOquery` has essentially one function you'll reach for: `getGEO()`. You hand it
a GEO **accession number** — the unique ID for a dataset, here `"GSE103512"` —
and it downloads the data over the network.

```{r geoquery10, echo=TRUE, cache=TRUE, message=FALSE}
library(GEOquery)
gse <- getGEO("GSE103512")[[1]]
```

::: {.callout-note}
## Why the `[[1]]` and the conversion later?

Two small historical quirks are worth a moment, because they trip up everyone
the first time.

First, `getGEO()` returns a **list** of datasets, not a single one. In the early
days a single accession could bundle several datasets, so the function always
returns a list. That's uncommon now, but the convention stayed because so much
older data is still useful — hence the `[[1]]` to grab the first (and here,
only) element.

Second, `GEOquery` was written in 2007 and hands you data in the older
Bioconductor container, the `ExpressionSet`. We'll convert it to the modern
`SummarizedExperiment` in the next step so we can use the accessors you already
know.
:::

Now convert the `ExpressionSet` into a `SummarizedExperiment` with `as()`:

```{r convert, message=FALSE}
library(SummarizedExperiment)
se <- as(gse, "SummarizedExperiment")
se
```

Printing `se` confirms what you have: a `SummarizedExperiment` whose rows are
genes (features) and whose columns are samples. The `assay()` holds the
expression matrix, and `colData()` holds the per-sample annotations — the same
three-part structure from the previous chapter.

```{r dims, echo=TRUE}
dim(se)
```

So this experiment has `r nrow(se)` genes measured across `r ncol(se)` samples.
The expression values live in the assay named `"exprs"` (a leftover name from the
`ExpressionSet` days), and you reach them with `assay(se, "exprs")`.

## Summarizing the sample annotations

Before any analysis, look at what the samples *are*. The interesting variables
here are the cancer type and whether each sample is tumor or normal tissue.
They live in `colData(se)` under the names `cancer.type.ch1` and `normal.ch1`.

The `table()` function counts unique values, and `with()` lets you refer to
those column names directly instead of typing `colData(se)$...` for each one:

```{r geoquery20, echo=TRUE, cache=TRUE}
with(colData(se), table(`cancer.type.ch1`, `normal.ch1`))
```

Read this as a grid: each row is a cancer type, and the two columns split it
into tumor (`no`, i.e. *not* normal) and normal samples. Most cancer types here
have both tumor and matched normal tissue, with tumor samples outnumbering
normals — exactly the design you'd want for a tumor-vs-normal comparison. A
quick table like this is the cheapest possible sanity check that the dataset
contains what the paper claimed.

## Keeping the most variable genes

We want to see how the samples group. Most of the genes barely change from
sample to sample, and those flat genes only add noise. A standard move before
dimensionality reduction is to keep just the **most variable** genes — the ones
actually carrying the signal that distinguishes samples.

::: {.callout-note}
## Why filter to high-variance genes before PCA?

PCA looks for the directions in which your samples differ most. A gene whose
expression is nearly constant everywhere contributes almost nothing to those
differences — it's dead weight that dilutes the signal and slows the
computation. Ranking genes by their **standard deviation** and keeping the top
slice focuses PCA on the genes that vary, which usually makes the biological
structure (tumor vs. normal, one cancer type vs. another) far easier to see.
There's no single correct cutoff; 500 genes is a reasonable starting point, and
you're encouraged to try others.
:::

Compute each gene's standard deviation with `apply()` (which applies a function
across the rows or columns of a matrix — here, `sd` to each row, `1` meaning
"by row"), then use `order()` to rank genes from most to least variable and keep
the top 500:

```{r sds, cache=TRUE, echo=TRUE}
sds <- apply(assay(se, "exprs"), 1, sd)
dat <- assay(se, "exprs")[order(sds, decreasing = TRUE)[1:500], ]
```

`order(sds, decreasing = TRUE)` returns the row positions sorted by standard
deviation, largest first; `[1:500]` takes the top 500; and we use those to
subset the expression matrix. `dat` is now a 500-gene by `r ncol(se)`-sample
matrix.

## Running and interpreting the PCA

PCA finds new axes (principal components) that capture the largest patterns of
variation in the data. `prcomp()` expects **samples as rows**, so we transpose
the matrix with `t()` before handing it over:

```{r pca, echo=TRUE, cache=TRUE}
pca_results <- prcomp(t(dat))
pca_df <- as.data.frame(pca_results$x)
pca_df$Type <- factor(colData(se)[, "cancer.type.ch1"])
pca_df$Normal <- factor(colData(se)[, "normal.ch1"])
```

`pca_results$x` holds each sample's coordinates along every principal component.
We pull those into a data frame and attach the cancer type and tumor/normal
label so we can color and shape the plot.

How much of the variation does each component capture? Each component's variance
is its squared standard deviation; divide by the total to get a fraction:

```{r pca-variance, echo=TRUE}
pca_var <- pca_results$sdev^2 / sum(pca_results$sdev^2)
plot(pca_var, xlab = "Principal component", ylab = "Variance explained",
     main = "Variance explained by each principal component", type = "b")
```

The curve drops off steeply: the first principal component (PC1) explains far
more of the variation than any other, and after the first few components each
one adds little. That steep drop is good news — it means a two-dimensional plot
of PC1 and PC2 captures most of what's going on, so it's worth plotting.

```{r pca-plot, echo=TRUE, fig.align='center'}
#| label: fig-pca
#| fig-cap: "Samples projected onto the first two principal components, colored by cancer type and shaped by tumor/normal status."
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, shape = Normal, color = Type)) +
    geom_point(alpha = 0.6) +
    theme(text = element_text(size = 18))
```

In @fig-pca the x-axis is PC1 and the y-axis is PC2, each point is one sample,
color marks the cancer type, and the shape marks tumor versus normal. The
dominant pattern is **cancer type**: samples of the same color cluster together,
so PC1 and PC2 are mostly separating tissues of origin from one another. Within
each cancer type, the tumor and normal shapes tend to occupy slightly different
regions, reflecting the tumor-vs-normal expression differences the original
study set out to characterize. In other words, a handful of unsupervised
components, computed from just 500 genes, already recover biology you'd expect:
different cancers look different, and within a cancer, tumor looks different from
normal.

## Summary

You just reused a published dataset from start to finish. You:

- Downloaded GSE103512 from GEO with `getGEO()` and grabbed the dataset with
  `[[1]]`.
- Converted the legacy `ExpressionSet` into a `SummarizedExperiment` and
  inspected its dimensions and sample annotations.
- Summarized the cancer-type-by-tumor/normal design with `table()`.
- Filtered to the 500 most variable genes, ran PCA with `prcomp()`, and read
  the variance-explained plot.
- Plotted PC1 vs. PC2 and saw the samples separate by cancer type, with
  tumor and normal differing within each type.

The same recipe — fetch, load into a `SummarizedExperiment`, filter, reduce,
plot — works for thousands of other GEO datasets.

## Exercises

::: {.callout-note collapse="true"}
## 1. Inspect the object

What is the class of `se`, and what are the dimensions of its `assay` and
`colData`? Which variables are stored in `colData`?

```{r ex1, eval=FALSE}
class(se)
dim(assay(se, "exprs"))   # genes x samples
dim(colData(se))          # samples x annotation variables
colnames(colData(se))     # the available sample variables
```

`se` is a `SummarizedExperiment`. The assay has one row per gene and one column
per sample; `colData` has one row per sample and one column per annotation
variable, including `cancer.type.ch1` and `normal.ch1`.
:::

::: {.callout-note collapse="true"}
## 2. Change the number of top genes

Re-run the analysis keeping the top 100 genes, and again with the top 2000. Does
the PC1-vs-PC2 plot change much?

```{r ex2, eval=FALSE}
dat100 <- assay(se, "exprs")[order(sds, decreasing = TRUE)[1:100], ]
pca100 <- prcomp(t(dat100))
plot(pca100$x[, 1], pca100$x[, 2], xlab = "PC1", ylab = "PC2")
```

The broad structure — samples separating by cancer type — is usually robust to
the exact cutoff. Too few genes can lose finer distinctions; too many let
low-variance noise creep back in. The stability of the big picture across
cutoffs is itself reassuring.
:::

::: {.callout-note collapse="true"}
## 3. Plot more components with `ggpairs()`

Use the `GGally` package's `ggpairs()` to view the first few principal
components together, colored by cancer type.

```{r ex3, eval=FALSE}
library(GGally)
ggpairs(pca_df, columns = 1:4, mapping = aes(color = Type))
```

`ggpairs()` draws every pairwise combination of the chosen columns (here PC1–PC4)
in a grid, so you can spot structure that PC1-vs-PC2 alone might hide — for
example, a cancer type that only separates along PC3 or PC4.
:::

[R]: https://cran.r-project.org/
[GEOquery]: https://bioconductor.org/packages/GEOquery
[GSE103512]: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE103512
[multidimensional scaling]: https://en.wikipedia.org/wiki/Multidimensional_scaling
