---
title: "Literate programming with Quarto"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

You run an analysis in the console, copy the key number and a figure into a Word
document, and email it off. A week later the data changes, you re-run the code —
but the document still shows last week's number and last week's plot, and nobody
notices. That gap between *the analysis* and *the writeup* is where a huge fraction
of scientific errors live.

**Literate programming** closes the gap. Instead of keeping code in one place and
prose in another, you write a single document that contains the prose, the code,
*and* the results the code produces — and the whole thing regenerates from the data
every time you render it. The numbers and figures can't go stale, because they are
never pasted: they are computed fresh, on the spot.

You are reading literate programming right now. This entire book is a
[Quarto](https://quarto.org) project: every code block you've seen actually *ran*,
every plot was generated straight from the data at high quality, and every figure
reference, citation, and cross-reference resolves itself automatically. The book
*is* its own source code. That's the idea this chapter teaches you to use.

## What you'll learn

By the end of this chapter you will be able to:

- Explain what literate programming is and the problem it solves.
- Name its concrete benefits — reproducibility, live in-text results, one source to
  many formats.
- Create, develop, render, and preview a Quarto document in RStudio, using the
  console-and-document loop you already know.
- Decide when literate programming is the right tool, and when a plain script or a
  pipeline is the better choice.
- (with a GitHub account) publish a rendered document to the web with one command.

## What is literate programming?

The term comes from Donald Knuth [@doi:10.1093/comjnl/27.2.97], who argued that a
program should be written first for *humans* to read — an explanatory narrative
with the code woven into it — and only incidentally for the machine to run. The
modern incarnation for data analysis is **Quarto** (and its predecessor, R
Markdown).

A Quarto document is a plain-text file with the extension `.qmd` that interleaves
three things:

- **prose**, written in Markdown (headings, lists, **bold**, links);
- **code**, in fenced *chunks*;
- the **output** that code produces — printed values, tables, and plots —
  generated when you *render* the document.

There is no separate "the analysis" and "the report." They are the same file, and
that file is the single source of truth.

Here is what a tiny `.qmd` looks like:

````qmd
---
title: "My first analysis"
format: html
---

Here is some prose describing what I'm about to do.

```{r}
x <- c(2, 4, 6, 8)
mean(x)
```

And a sentence with a value computed inline: the mean is `r mean(x)`.
````

The header between the `---` lines is **YAML** metadata (title, output format).
Below it, prose and ```` ```{r} ```` code chunks alternate freely.

## Why bother? The benefits

```{r demo-vector, echo=FALSE}
x <- c(2, 4, 6, 8)
```

- **Reproducibility, with no copy-paste drift.** Rendering re-runs *all* the code
  from the data. A figure or number in the document can't be stale, because it is
  regenerated rather than pasted. Reproducibility like this is now considered a
  baseline expectation of computational work [@doi:10.1371/journal.pcbi.1003285].
- **Live results in the text.** You can drop a computed value straight into a
  sentence — an *inline* code expression. For example, the mean of the vector
  `x <- c(2, 4, 6, 8)` is `r mean(x)`, and that number was calculated when this page
  was rendered, not typed by hand. Change the data, re-render, and the sentence
  updates itself.
- **One source, many formats.** The same `.qmd` renders to HTML, PDF, or Word —
  you don't rewrite anything to change the output.
- **Beautiful, consistent rendering.** Clean typography, high-quality plots
  straight from your code, tables, mathematics, citations, and automatic
  figure/section **cross-references** that number and link themselves.

::: {.callout-tip}
## Exhibit A: the book in your hands
You don't have to take this on faith — you're holding the proof. This book is a
Quarto project, so its code blocks **always work** (they are executed and checked
on every build; if a chunk broke, the build would fail), its figures are generated
fresh at high quality, and its citations and `@fig-`/`@sec-` cross-references all
resolve automatically. Literate programming, demonstrating itself.
:::

## Your first Quarto document in RStudio

The best part: developing a Quarto document is the same console workflow you
already use, with the code saved in a document and the output captured alongside
it.

1. **Create it.** In RStudio: **File → New File → Quarto Document…**, give it a
   title, choose **HTML**, and click *Create*. You'll get a starter `.qmd` with a
   YAML header and a sample chunk.
2. **Recognize the parts.** The `---` header sets the title and format; the body
   alternates Markdown prose with ```` ```{r} ```` chunks.
3. **Develop in the mirrored console.** Click the green ▶ button at the top-right of
   a chunk (or press `Ctrl`/`Cmd`+`Shift`+`Enter`) and the chunk runs *in the R
   console you already know* — the same console from the
   [R mechanics chapter](r_intro_mechanics.qmd) — with its results shown inline,
   right under the chunk. You can also run a single line with
   `Ctrl`/`Cmd`+`Enter` to experiment, exactly as in a script. This is the key
   mental shift: a `.qmd` is your interactive session, made permanent.
4. **Render.** Click the **Render** button. Quarto runs the *entire* document
   top-to-bottom in a **fresh** R session and produces the output, shown in the
   preview pane.
5. **Edit how you like.** RStudio offers a **Source** editor (raw Markdown) and a
   **Visual** editor (a word-processor-like view); switch between them with the
   buttons above the document. Use whichever you prefer — the `.qmd` file is the
   same either way.

::: {.callout-important}
## Why "fresh session" is a feature, not a nuisance
Rendering starts from a blank R session, so it only sees what the document itself
defines. That catches the most common reproducibility bug: code that secretly
relied on a variable you created in the console an hour ago and forgot to write
down. If it renders, it reproduces — on your machine *and* on someone else's.
:::

## When *not* to use literate programming

Literate programming is for the **analysis and communication layer** of your work.
Underneath that often sits a **heavy data-processing or pipeline layer**, and that
layer usually wants different tools. Reach for a plain script (or a pipeline tool)
instead when:

- **A step is slow or compute-heavy.** If a chunk takes an hour, you don't want it
  re-running every time you fix a typo in a sentence. Either cache it (see below) or
  move it to a script that runs once and saves its result.
- **The work is an independent heavy-lift pipeline step.** Aligning sequencing
  reads, processing huge files, multi-stage bioinformatics workflows — these belong
  in scripts orchestrated by a `Makefile`, the `targets` package, Snakemake, or
  Nextflow, not in a render-on-every-build document.
- **You're still exploring.** Poke around in the console first; write it up *after*
  you know what you've found. Don't fight a document while you're still thinking.
- **You're building reusable software.** Functions and packages live in `.R` files,
  not in narrative documents.
- **The deliverable isn't a document** — a trained model, a cleaned dataset, an API.
  Produce those with scripts.

A one-line rule of thumb: **if the output is meant for a human to read, reach for a
Quarto document; if the output is data, a model, or a pipeline artifact, reach for
a script.**

## Keeping heavy steps literate anyway

The boundary above isn't a wall. Three tools let you keep an expensive analysis
inside a literate document without paying the cost on every render:

- **Chunk caching.** Add `#| cache: true` to a chunk and its results are saved and
  reused until *that chunk's code* changes — ideal for one slow model fit.
- **Freeze.** A Quarto book can set `execute: freeze: auto`, which re-executes a
  document only when its source changes and reuses stored results otherwise. This
  book uses exactly that: heavy Bioconductor chapters are frozen, so a normal build
  reuses their results instead of re-downloading and re-computing.
- **Precompute and load.** Run the heavy step once in a separate script, `saveRDS()`
  the result, and have the document `readRDS()` it. The document stays literate and
  fast; the heavy lifting happens out of band.

## Publishing to GitHub Pages

If you have a GitHub account, putting a rendered document on the public web is one
command. From the project's directory:

```bash
quarto publish gh-pages
```

This renders your document, pushes the output to a special `gh-pages` branch of
your repository, and switches on GitHub Pages — giving you a public address like
`https://USERNAME.github.io/REPO/`. Run the same command again whenever you want to
update the live version. (See the [Git and GitHub](git_and_github.qmd) appendix if
you're new to repositories.)

::: {.callout-note}
## No GitHub account? No problem
`quarto publish quarto-pub` publishes to [Quarto Pub](https://quartopub.com), a free
hosting service, with the same one-command flow and no GitHub required. And for a
document or book you update often, you can have **GitHub Actions** render and
publish automatically on every push — which is precisely how this book is built and
deployed.
:::

## "Document or script?" — a quick checklist

| The work… | Use |
|---|---|
| produces a report, figure, or notebook **for people to read** | a Quarto document |
| produces a **dataset, model, or pipeline artifact** | a script |
| takes seconds-to-minutes and belongs in the narrative | a document (cache if needed) |
| takes hours, or is shared across many analyses | a script + a saved artifact |
| is open-ended **exploration** | the console/script now; write it up later |

## Summary

Literate programming keeps code, prose, and results in one document that
regenerates from the data, so your writeup can never drift out of sync with your
analysis. In Quarto that document is a `.qmd` you develop with the same RStudio
console loop you already use, then **Render** in a fresh session — which both
produces a polished, multi-format report and guarantees it reproduces. It is the
right tool for the analysis-and-communication layer, while heavy processing and
pipelines belong in scripts (with caching, freeze, and precomputed artifacts to
bridge the two). And with one command you can publish the result to the web.

## Exercises

1. **Make one.** In RStudio, create a new Quarto document with a title, a sentence
   of prose, and a code chunk that computes something (e.g. the mean of a vector).
   Render it to HTML.
2. **Watch it update.** Add an *inline* value to your prose so a number in the
   sentence is computed from your data. Change the data, re-render, and confirm the
   sentence changed on its own.
3. **Classify.** For each task, say whether it belongs in a Quarto document or a
   script, and why: (a) fit a quick model and report the result; (b) align 200 GB of
   sequencing reads; (c) explore a brand-new dataset for the first time; (d) write a
   reusable plotting function you'll call from several analyses.
4. **(Optional, needs GitHub.)** Publish a document with `quarto publish gh-pages`
   and visit the resulting URL.

::: {.callout-note collapse="true"}
## Solutions

1. New File → Quarto Document, write a title and a ```` ```{r} ```` chunk such as
   `mean(c(2, 4, 6, 8))`, then click **Render**. The HTML appears in the preview.
2. Define a value in a chunk, then reference it inline in a sentence with an `r`
   expression in single backticks; re-rendering after editing the value updates the
   text automatically — nothing was retyped.
3. (a) **Document** — it's a short analysis meant to be read. (b) **Script** — a
   heavy, independent processing step; orchestrate it with a pipeline tool and have
   a later document load its summarized output. (c) **Script/console first** —
   explore, then write up once you know the story. (d) **Script** (`.R`) — reusable
   software, not a narrative.
4. `quarto publish gh-pages` renders and pushes to the `gh-pages` branch; the live
   page appears at `https://USERNAME.github.io/REPO/`.
:::

## Resources

- [Quarto documentation](https://quarto.org) — the authoritative guide, with
  tutorials for RStudio, VS Code, and Jupyter.
- [*R for Data Science*](https://r4ds.hadley.nz/), the Quarto chapters — a gentle,
  practical introduction in book form.
- Knuth, D. E. (1984). *Literate Programming* [@doi:10.1093/comjnl/27.2.97] — the
  original essay that named the idea.
- Sandve et al. (2013). *Ten Simple Rules for Reproducible Computational Research*
  [@doi:10.1371/journal.pcbi.1003285] — why this workflow matters.
