---
title: "Working with distribution functions"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
```

Heights, blood pressures, measurement errors, log-transformed gene-expression
values — measure enough of almost anything in biology and the values tend to pile
up into the same bell-shaped curve: the **normal distribution**. Once your data
follow that curve, you'll want to ask questions of it. *What fraction of samples
fall below some cutoff? Which value marks the top 5%? Give me twenty random draws
to simulate an experiment.* R answers each of those with a different function.

The catch is remembering which function does which job. This chapter is a
**visual guide**: for every function we pair the R call with a picture of exactly
what it measures on the bell curve, because the difference between a *probability*,
a *density*, and a *quantile* is far easier to see than to memorize. It's genuinely
hard to reason about these functions without a shape to point at — so a shape is
what we'll build for each one.

## What you'll learn

- Use `pnorm()` to turn a value into a probability (the area under the curve to its
  left).
- Use `dnorm()` to get the height (density) of the curve at a value.
- Use `qnorm()` to go the other direction — from a probability back to a value.
- Use `rnorm()` to draw random samples from a normal distribution.
- Combine these functions to answer real questions about normally distributed
  data, such as IQ scores.

::: {.callout-tip}
## One naming rule for *every* distribution

R builds these names from a one-letter prefix plus the distribution: **d** for
density, **p** for probability, **q** for quantile, **r** for random. Learn it once
and it transfers everywhere — `dbinom`/`pbinom`/`qbinom`/`rbinom` for the binomial,
`dt`/`pt`/`qt`/`rt` for the t-distribution we meet in the next chapter,
`dpois`/`ppois`/… for the Poisson. Here we use the `…norm` family for the normal.
:::

```{r echo=FALSE}
# Prep the distribution data
x <- seq(-3, 3, 0.005)
y <- dnorm(x)
normal <- as.data.frame(cbind(x, y))
```

The four functions, at a glance:

| Function | Input | Output |
|----------|-------|--------|
| `pnorm`  | a value `q` | P(X < q): area under the curve to the **left** of `q` |
| `dnorm`  | a value `x` | the **height** of the density curve at `x` |
| `qnorm`  | a probability from 0 to 1 | the value `x` where P(X < x) equals that probability |
| `rnorm`  | a count `n` | `n` random samples from the distribution |

: Functions for the normal distribution {#tbl-norm-funcs}

In one sentence each: `pnorm` gives the area to the left of a point, `dnorm` gives
the height of the curve at a point, `qnorm` is the inverse of `pnorm` (probability
back to value), and `rnorm` generates random samples. The rest of the chapter
unpacks them one at a time.

## pnorm

`pnorm` answers the question "what fraction of the distribution lies to the
**left** of this value?" If you don't supply a mean and standard deviation, R
assumes the *standard normal* (mean 0, standard deviation 1). The help page gives
this template:

```{r eval=FALSE}
pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
```

You hand it `q`, a value on the x-axis, and it returns the area under the curve to
the left of that point. The mean sits dead centre and splits the curve in half, so:

```{r}
pnorm(0)    # half the distribution is below the mean
pnorm(1)    # ~84% is below one standard deviation above the mean
pnorm(-1)   # ~16% is below one standard deviation below it
```

The figures below show each of those as a shaded area.
```{r echo=FALSE}
# pnorm plot 1
# pnorm(0)
x2 <- c(x[1:600], rep(0,601))
y2 <- c(y[1:600], rep(0,601))

p <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(linewidth = 1.3) +
     geom_vline(xintercept = 0, color = "blue", linewidth = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(0)", x = NULL, y = NULL, 
          subtitle = "P is for Probability") +
     geom_area(inherit.aes = F, mapping = aes(x=x2, y=y2), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.50", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 1.2
# 1-pnorm(0)
x9 <- c(rep(0,600), x[601:1201])
y9 <- c(rep(0,600), y[601:1201])

p1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(linewidth = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "1-pnorm(0)", x = NULL, y = NULL, 
          subtitle = "P is for Probability") +
     geom_area(inherit.aes = F, mapping = aes(x=x9, y=y9), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = 2.6, xmax = 2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.50", x = 2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = 2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 2
# pnorm(1)
x3 <- c(x[1:801], rep(1,400))
y3 <- c(y[1:801], rep(0,400))

p2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(1)", subtitle = "P is for probability", 
          x = NULL, y = NULL) +
     geom_area(inherit.aes = F, mapping = aes(x=x3, y=y3), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.84", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 1, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 3
# pnorm(-1)
x4 <- c(x[1:401], rep(-1,800))
y4 <- c(y[1:401], rep(0,800))

p3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(-1)", subtitle = "P is for probability", 
          x = NULL, y = NULL) +
     geom_area(inherit.aes = F, mapping = aes(x=x4, y=y4), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.16", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = -1, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-pnorm
#| fig-cap: "The pnorm function takes a quantile (value on the x-axis) and returns the area under the curve to the left of that value."

print(p)
print(p1)
print(p2)
print(p3)
```

The option `lower.tail = TRUE` tells R to use the area to the *left* of the given
point — the default. To get the area to the **right** instead, either switch to
`lower.tail = FALSE` or subtract from 1:

```{r}
pnorm(1, lower.tail = FALSE)   # area to the RIGHT of 1
1 - pnorm(1)                   # exactly the same thing
```



## dnorm

`dnorm` gives the **height** of the bell curve at a value — the probability
*density*. The curve is tallest at the mean and falls off symmetrically on either
side. Unlike `pnorm`, the result is not a probability you can read as a percentage;
it's the height of the curve, used mainly for *drawing* the distribution or in
likelihood calculations.

```{r}
dnorm(0)    # the peak of the standard normal
dnorm(1)    # lower, one SD out...
dnorm(-1)   # ...and the same height on the other side, by symmetry
```

The figures below mark each of these heights on the curve.

```{r echo=FALSE}
# dnorm plot 1
# dnorm(0)

d1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(0)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.33, ymax = 0.39,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.40", x = -2.7, y = 0.36, size = 6) +
     annotate(geom = "segment", y = 0.4, yend = 0.4, x = -3, xend = 0, 
              color = "tomato", size = 1.25, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# dnorm plot 2
# dnorm(1)

d2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(1)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.17, ymax = 0.23,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.24", x = -2.7, y = 0.2, size = 6) +
     annotate(geom = "segment", y = 0.24, yend = 0.24, x = -3, xend = 1, 
              color = "tomato", size = 1.1, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
           plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# dnorm plot 3
# dnorm(-1)

d3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(-1)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.17, ymax = 0.23,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.24", x = -2.7, y = 0.2, size = 6) +
     annotate(geom = "segment", y = 0.24, yend = 0.24, x = -3, xend = -1, 
              color = "tomato", size = 1.1, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
           plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-dnorm
#| fig-cap: "The `dnorm` function returns the height of the normal distribution at a given point."

print(d1)
print(d2)
print(d3)
```

## qnorm

`qnorm` runs `pnorm` in reverse: hand it a probability and it returns the value
with that much area to its left. Since `pnorm(1)` is about 0.84, `qnorm(0.84)` is
about 1 — the two functions undo each other. This is how you find **percentiles**.

```{r}
qnorm(0.5)     # the median: half the area lies to its left
qnorm(0.25)    # the first quartile
qnorm(0.975)   # the value cutting off the top 2.5% — the famous 1.96
```

The figures below show several quantiles as shaded areas.

```{r echo=FALSE}
# qnorm plot 1
# qnorm(0.5)

q1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.5)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x2, y=y2), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.25, xmax = 0.25, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0", x = 0, y = 0.06, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "50%", x = -1.25, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 2
# qnorm(0.25)
x5 <- c(x[1:466], rep(-0.6745,735))
y5 <- c(y[1:466], rep(0,735))

q2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -0.6745, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.25)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x5, y=y5), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "-0.67", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = -0.6745, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "25%", x = -1.65, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 3
# qnorm(0.1)
x6 <- c(x[1:345], rep(-1.28,856))
y6 <- c(y[1:345], rep(0,856))

q3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1.28, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.1)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x6, y=y6), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "-1.28", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = -1.28, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "10%", x = -1.85, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 4
# qnorm(0.75)
x7 <- c(x[1:735], rep(0.6745,466))
y7 <- c(y[1:735], rep(0,466))

q4 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0.6745, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.75)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x7, y=y7), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.67", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = 0.6745, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "75%", x = -.95, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 5
# qnorm(0.9)
x8 <- c(x[1:856], rep(1.28,345))
y8 <- c(y[1:856], rep(0,345))

q5 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1.28, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.9)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x8, y=y8), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "1.28", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = 1.28, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "90%", x = 0.8, y = 0.025, 
              size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-qnorm
#| fig-cap: "The qnorm function is the inverse of the pnorm function in that it takes a probability and gives the quantile." 

print(q1)
print(q2)
print(q3)
print(q4)
print(q5)
```


## rnorm

`rnorm` draws random samples from the distribution — the workhorse for *simulating*
data (we leaned on it heavily in the t-distribution chapter). Give it a count `n`
and it hands back that many random values:

```{r}
rnorm(5)
```

Run that again and you'll get five *different* numbers. To make a simulation
reproducible — so you and a reader see the same "random" draws — set a seed first:

```{r}
set.seed(42)
rnorm(5)
```

And just like the other functions, `mean` and `sd` shift and scale the draws:

```{r}
rnorm(5, mean = 100, sd = 15)   # e.g. five simulated IQ scores
```

```{r echo=FALSE}
# rnorm plot
# rnorm(5)

r1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "rnorm(n=5)", subtitle = "R is for Random generation", 
          x = NULL, y = NULL, caption = "Output: c(-1.8, -0.8, 0.1, 0.45, 1.1)") +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 1.1, xend = 1.1,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 0.45, xend = 0.45,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = -0.8, xend = -0.8,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 0.1, xend = 0.1,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = -1.8, xend = -1.8,
              color = "blue", linetype = 2) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5)) +
     theme(plot.caption = element_text(face = "bold", size = 14, hjust = 0.5))
```


```{r}
#| label: fig-rnorm
#| fig-cap: "The rnorm function takes a number of samples and returns a vector of random numbers from the normal distribution (with mean=0, sd=1 as defaults)"
print(r1)
```


## Worked example: IQ scores

The normal distribution (also called the Gaussian distribution) is defined by two
numbers: the **mean** (µ), which sets the centre, and the **standard deviation**
(σ), which sets the spread. IQ scores are deliberately built to be normal with a
**mean of 100** and a **standard deviation of 15**, which makes them a perfect
playground for the four functions.

A handy rule of thumb — the "68–95–99.7 rule" — says about 68% of values fall
within one standard deviation of the mean (here, 85 to 115) and about 95% within
two (70 to 130). Let's check the 68% with `pnorm`. The area *between* 85 and 115 is
the area below 115 minus the area below 85:

```{r}
pnorm(115, mean = 100, sd = 15) - pnorm(85, mean = 100, sd = 15)
```

About 0.68 — exactly what the rule promises. Notice the pattern: a "between"
question becomes one `pnorm` minus another.

## Exercises

Use the IQ distribution (mean 100, sd 15) for each. Each solution names *which* of
the four functions fits the question — that choice is the real skill.

1. **Top 10%.** Which IQ score marks the 90th percentile — the cutoff above which
   only the top 10% fall?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   qnorm(0.9, mean = 100, sd = 15)
   ```
   We want a *value* from a *probability*, so this is a `qnorm` job. About 119.
   :::

2. **Above 130.** What fraction of people have an IQ above 130 (two standard
   deviations up)?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   1 - pnorm(130, mean = 100, sd = 15)
   # or, equivalently:
   pnorm(130, mean = 100, sd = 15, lower.tail = FALSE)
   ```
   About 2.3% — consistent with the 95% rule leaving roughly 2.5% in each tail.
   :::

3. **Below 70.** What fraction have an IQ below 70?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   pnorm(70, mean = 100, sd = 15)
   ```
   Also about 2.3%. The curve is symmetric, so the lower tail past 70 mirrors the
   upper tail past 130.
   :::

4. **Simulate a classroom.** Draw 30 random IQ scores and count how many land above
   115. (Set a seed so your answer is reproducible.)

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   set.seed(1)
   scores <- rnorm(30, mean = 100, sd = 15)
   sum(scores > 115)
   ```
   Theory predicts about 16% above 115 — roughly 5 of 30 — but any single small
   sample wobbles around that. Re-run with a different seed to feel the sampling
   variability for yourself.
   :::

## Summary

You now have the four-function toolkit for the normal distribution — and, thanks to
R's naming rule, the blueprint for every other distribution too:

- `pnorm(x)` — area to the **left** of `x` (a probability). Use `lower.tail = FALSE`
  or `1 - pnorm(x)` for the right tail, and one `pnorm` minus another for a range.
- `dnorm(x)` — the **height** of the curve at `x` (a density).
- `qnorm(p)` — the **value** at probability `p` (the inverse of `pnorm`); your tool
  for percentiles.
- `rnorm(n)` — `n` random **draws**; pair it with `set.seed()` for reproducible
  simulations.

Keep the pictures in mind: `pnorm` is an area, `dnorm` is a height, `qnorm` reads
the area axis backwards to a value, and `rnorm` scatters points along the curve.

    
