---
title: "Working with distribution functions"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
```

Heights, blood pressures, measurement errors, log-transformed gene-expression
values — measure enough of almost anything in biology and the values tend to pile
up into the same bell-shaped curve: the **normal distribution**. Once your data
follow that curve, you'll want to ask questions of it. *What fraction of samples
fall below some cutoff? Which value marks the top 5%? Give me twenty random draws
to simulate an experiment.* R answers each of those with a different function.

The catch is remembering which function does which job. This chapter is a
**visual guide**: for every function we pair the R call with a picture of exactly
what it measures on the bell curve, because the difference between a *probability*,
a *density*, and a *quantile* is far easier to see than to memorize. It's genuinely
hard to reason about these functions without a shape to point at — so a shape is
what we'll build for each one.

## What you'll learn

- Use `pnorm()` to turn a value into a probability (the area under the curve to its
  left).
- Use `dnorm()` to get the height (density) of the curve at a value.
- Use `qnorm()` to go the other direction — from a probability back to a value.
- Use `rnorm()` to draw random samples from a normal distribution.
- Combine these functions to answer real questions about normally distributed
  data, such as IQ scores.

::: {.callout-tip}
## One naming rule for *every* distribution

R builds these names from a one-letter prefix plus the distribution: **d** for
density, **p** for probability, **q** for quantile, **r** for random. Learn it once
and it transfers everywhere — `dbinom`/`pbinom`/`qbinom`/`rbinom` for the binomial,
`dt`/`pt`/`qt`/`rt` for the t-distribution we meet in the next chapter,
`dpois`/`ppois`/… for the Poisson. Here we use the `…norm` family for the normal.
:::

```{r echo=FALSE}
# Prep the distribution data
x <- seq(-3, 3, 0.005)
y <- dnorm(x)
normal <- as.data.frame(cbind(x, y))
```

The four functions, at a glance:

| Function | Input | Output |
|----------|-------|--------|
| `pnorm`  | a value `q` | P(X < q): area under the curve to the **left** of `q` |
| `dnorm`  | a value `x` | the **height** of the density curve at `x` |
| `qnorm`  | a probability from 0 to 1 | the value `x` where P(X < x) equals that probability |
| `rnorm`  | a count `n` | `n` random samples from the distribution |

: Functions for the normal distribution {#tbl-norm-funcs}

In one sentence each: `pnorm` gives the area to the left of a point, `dnorm` gives
the height of the curve at a point, `qnorm` is the inverse of `pnorm` (probability
back to value), and `rnorm` generates random samples. The rest of the chapter
unpacks them one at a time.

## pnorm

`pnorm` answers the question "what fraction of the distribution lies to the
**left** of this value?" If you don't supply a mean and standard deviation, R
assumes the *standard normal* (mean 0, standard deviation 1). The help page gives
this template:

```{r eval=FALSE}
pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
```

You hand it `q`, a value on the x-axis, and it returns the area under the curve to
the left of that point. The mean sits dead centre and splits the curve in half, so:

```{r}
pnorm(0)    # half the distribution is below the mean
pnorm(1)    # ~84% is below one standard deviation above the mean
pnorm(-1)   # ~16% is below one standard deviation below it
```

The figures below show each of those as a shaded area.
```{r echo=FALSE}
# pnorm plot 1
# pnorm(0)
x2 <- c(x[1:600], rep(0,601))
y2 <- c(y[1:600], rep(0,601))

p <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(linewidth = 1.3) +
     geom_vline(xintercept = 0, color = "blue", linewidth = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(0)", x = NULL, y = NULL, 
          subtitle = "P is for Probability") +
     geom_area(inherit.aes = F, mapping = aes(x=x2, y=y2), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.50", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 1.2
# 1-pnorm(0)
x9 <- c(rep(0,600), x[601:1201])
y9 <- c(rep(0,600), y[601:1201])

p1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(linewidth = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "1-pnorm(0)", x = NULL, y = NULL, 
          subtitle = "P is for Probability") +
     geom_area(inherit.aes = F, mapping = aes(x=x9, y=y9), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = 2.6, xmax = 2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.50", x = 2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = 2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 2
# pnorm(1)
x3 <- c(x[1:801], rep(1,400))
y3 <- c(y[1:801], rep(0,400))

p2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(1)", subtitle = "P is for probability", 
          x = NULL, y = NULL) +
     geom_area(inherit.aes = F, mapping = aes(x=x3, y=y3), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.84", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 1, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# pnorm plot 3
# pnorm(-1)
x4 <- c(x[1:401], rep(-1,800))
y4 <- c(y[1:401], rep(0,800))

p3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "pnorm(-1)", subtitle = "P is for probability", 
          x = NULL, y = NULL) +
     geom_area(inherit.aes = F, mapping = aes(x=x4, y=y4), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -2.6, xmax = -2, ymin = 0.12, ymax = 0.18,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.16", x = -2.3, y = 0.15, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = -1, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-pnorm
#| fig-cap: "The pnorm function takes a quantile (value on the x-axis) and returns the area under the curve to the left of that value."

print(p)
print(p1)
print(p2)
print(p3)
```

The option `lower.tail = TRUE` tells R to use the area to the *left* of the given
point — the default. To get the area to the **right** instead, either switch to
`lower.tail = FALSE` or subtract from 1:

```{r}
pnorm(1, lower.tail = FALSE)   # area to the RIGHT of 1
1 - pnorm(1)                   # exactly the same thing
```



## The cumulative distribution function

Step back and look at what `pnorm` really *is*. Each call answers one question —
"how much probability lies to the left of this value?" If you ask that question at
*every* point along the x-axis and plot the answers, you trace out a single curve:
the **cumulative distribution function**, or **CDF**. Where the density (`dnorm`)
tells you how *concentrated* probability is at each point, the CDF tells you how
much has *accumulated* by that point. They are two views of the same distribution,
and learning to move between them is most of what "distribution intuition" really
means.

Here is the CDF of the standard normal — `pnorm(x)` evaluated at every `x`:

```{r}
#| label: fig-cdf-curve
#| echo: false
#| fig-cap: "The cumulative distribution function of the standard normal: pnorm(x) plotted at every x. It is the running total of the area under the density curve, climbing from 0 (no probability accumulated yet) to 1 (all of it)."
#| fig-width: 7
#| fig-height: 3.4
cdf <- data.frame(x = x, p = pnorm(x))
ggplot(cdf, aes(x, p)) +
  geom_line(linewidth = 1.3, color = "steelblue") +
  scale_x_continuous(n.breaks = 7) +
  labs(x = "x", y = expression(P(X <= x))) +
  theme_classic(base_size = 14)
```

A few features are worth naming, because they hold for *every* CDF, not just the
normal:

- It **starts near 0 and ends near 1.** A CDF is accumulated probability, and all
  the probability sums to 1, so the curve can only climb from 0 up to 1.
- It **never goes down.** Moving to the right can only *add* area, never remove it,
  so a CDF is always non-decreasing.
- It is **S-shaped (for the normal) because of where the density lives.** Out in
  the left tail almost no probability is accumulating, so the curve is nearly flat;
  through the middle — where the bell is tallest — probability piles up quickly and
  the curve climbs steeply; in the right tail there is little left to add, so it
  levels off toward 1.

That last point is the key link between the two views: **the density is the *slope*
of the CDF.** The curve rises fastest exactly where the density is tallest (the
mean), and flattens wherever the density is near zero (the tails). Read the other
direction, the **CDF is the running total — the accumulated area — of the density.**

### The area is the height

The cleanest way to *feel* the connection is to put the two pictures side by side.
In @fig-cdf-link the same value `q` appears in both rows: the **shaded area** under
the density (top) is exactly the **height** of the CDF (bottom). As `q` slides to
the right, the shading fills in and the point climbs the S-curve. That shaded
area-equals-curve-height is precisely the number `pnorm(q)` returns.

```{r}
#| label: fig-cdf-link
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Two views of the same probability at q = -1, 0, and 1. Top row: the density with the area to the left of q shaded. Bottom row: the CDF, with a point at (q, pnorm(q)) and dashed guides. In every column the shaded area above equals the height below."
#| fig-width: 9
#| fig-height: 5
library(patchwork)
pair <- function(q) {
  area <- subset(normal, x <= q)
  top <- ggplot(normal, aes(x, y)) +
    geom_area(data = area, fill = "tomato", alpha = 0.3) +
    geom_line(linewidth = 1) +
    labs(subtitle = paste0("shaded area = ", sprintf("%.2f", pnorm(q))),
         x = NULL, y = "density") +
    theme_classic(base_size = 12)
  bot <- ggplot(cdf, aes(x, p)) +
    geom_line(linewidth = 1, color = "steelblue") +
    annotate("segment", x = q, xend = q, y = 0, yend = pnorm(q),
             linetype = 2, color = "grey45") +
    annotate("segment", x = -3, xend = q, y = pnorm(q), yend = pnorm(q),
             linetype = 2, color = "grey45") +
    annotate("point", x = q, y = pnorm(q), size = 3, color = "steelblue") +
    labs(subtitle = paste0("pnorm(", q, ") = ", sprintf("%.2f", pnorm(q))),
         x = "x", y = expression(P(X <= x))) +
    theme_classic(base_size = 12)
  top / bot
}
pair(-1) | pair(0) | pair(1)
```

### Everything you already know, on one curve

Once you can see the CDF as a curve, the whole `d`/`p`/`q`/`r` family lines up on
it:

- **`pnorm`** reads the curve **upward**: hand it a value `q` on the x-axis and it
  returns the *height* of the curve there — the probability to the left.
- **`qnorm`** reads the same curve **across**: hand it a probability on the y-axis
  and it returns the *x* where the curve reaches that height. That is why `qnorm`
  is the exact inverse of `pnorm` — it's the same S-curve, read sideways.
- **`dnorm`** is the curve's **steepness** — how fast probability is piling up at
  that point.

So `pnorm` and `qnorm` are not two separate things to memorize; they are the same
curve approached from two directions, and `dnorm` is its slope.

::: {.callout-note}
## This isn't special to the normal
Everything in this section is general. *Every* distribution — binomial, Poisson,
t, exponential, and the rest — has a density (or, for counts, a probability mass)
and a matching CDF, related in exactly the same way: the CDF is the accumulated
area under the density, and the density is the slope of the CDF. That shared
structure is *why* R gives every distribution the same `d`/`p`/`q`/`r` quartet.
Swap `norm` for `pois` or `binom` and the picture you just built still holds.
:::

## dnorm

`dnorm` gives the **height** of the bell curve at a value — the probability
*density*. The curve is tallest at the mean and falls off symmetrically on either
side. Unlike `pnorm`, the result is not a probability you can read as a percentage;
it's the height of the curve, used mainly for *drawing* the distribution or in
likelihood calculations.

```{r}
dnorm(0)    # the peak of the standard normal
dnorm(1)    # lower, one SD out...
dnorm(-1)   # ...and the same height on the other side, by symmetry
```

The figures below mark each of these heights on the curve.

```{r echo=FALSE}
# dnorm plot 1
# dnorm(0)

d1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(0)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.33, ymax = 0.39,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.40", x = -2.7, y = 0.36, size = 6) +
     annotate(geom = "segment", y = 0.4, yend = 0.4, x = -3, xend = 0, 
              color = "tomato", size = 1.25, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# dnorm plot 2
# dnorm(1)

d2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(1)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.17, ymax = 0.23,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.24", x = -2.7, y = 0.2, size = 6) +
     annotate(geom = "segment", y = 0.24, yend = 0.24, x = -3, xend = 1, 
              color = "tomato", size = 1.1, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
           plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# dnorm plot 3
# dnorm(-1)

d3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "dnorm(-1)", subtitle = "D is for Distribution", 
          x = NULL, y = NULL) +
     annotate(geom = "rect", xmin = -3, xmax = -2.4, ymin = 0.17, ymax = 0.23,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.24", x = -2.7, y = 0.2, size = 6) +
     annotate(geom = "segment", y = 0.24, yend = 0.24, x = -3, xend = -1, 
              color = "tomato", size = 1.1, linetype = 6) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
           plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-dnorm
#| fig-cap: "The `dnorm` function returns the height of the normal distribution at a given point."

print(d1)
print(d2)
print(d3)
```

## qnorm

`qnorm` runs `pnorm` in reverse: hand it a probability and it returns the value
with that much area to its left. Since `pnorm(1)` is about 0.84, `qnorm(0.84)` is
about 1 — the two functions undo each other. This is how you find **percentiles**.

```{r}
qnorm(0.5)     # the median: half the area lies to its left
qnorm(0.25)    # the first quartile
qnorm(0.975)   # the value cutting off the top 2.5% — the famous 1.96
```

The figures below show several quantiles as shaded areas.

```{r echo=FALSE}
# qnorm plot 1
# qnorm(0.5)

q1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.5)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x2, y=y2), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.25, xmax = 0.25, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0", x = 0, y = 0.06, size = 6) +
     annotate(geom = "segment", x = -2.7, xend = 0, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "50%", x = -1.25, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 2
# qnorm(0.25)
x5 <- c(x[1:466], rep(-0.6745,735))
y5 <- c(y[1:466], rep(0,735))

q2 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -0.6745, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.25)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x5, y=y5), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "-0.67", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = -0.6745, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "25%", x = -1.65, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 3
# qnorm(0.1)
x6 <- c(x[1:345], rep(-1.28,856))
y6 <- c(y[1:345], rep(0,856))

q3 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = -1.28, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.1)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x6, y=y6), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "-1.28", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = -1.28, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "10%", x = -1.85, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 4
# qnorm(0.75)
x7 <- c(x[1:735], rep(0.6745,466))
y7 <- c(y[1:735], rep(0,466))

q4 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 0.6745, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.75)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x7, y=y7), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "0.67", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = 0.6745, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "75%", x = -.95, y = 0.025, size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
# qnorm plot 5
# qnorm(0.9)
x8 <- c(x[1:856], rep(1.28,345))
y8 <- c(y[1:856], rep(0,345))

q5 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     geom_vline(xintercept = 1.28, color = "blue", size = 1.3, linetype = 2) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "qnorm(0.9)", x = NULL, y = NULL, 
          subtitle = "Q is for Quantile") +
     geom_area(inherit.aes = F, mapping = aes(x=x8, y=y8), fill = "tomato", 
               alpha = 0.2) +
     annotate(geom = "rect", xmin = -0.3, xmax = 0.3, ymin = 0.03, ymax = 0.09,
              alpha = 1, fill = "white", color = "black") +
     annotate(geom = "text", label = "1.28", x = 0, y = 0.06, size = 5) +
     annotate(geom = "segment", x = -2.7, xend = 1.28, y = 0.01, yend = 0.01, 
              size = 1.1, arrow = arrow(length = unit(3, "mm"))) +
     annotate(geom = "text", label = "90%", x = 0.8, y = 0.025, 
              size = 4) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5))
```

```{r echo=FALSE}
#| label: fig-qnorm
#| fig-cap: "The qnorm function is the inverse of the pnorm function in that it takes a probability and gives the quantile." 

print(q1)
print(q2)
print(q3)
print(q4)
print(q5)
```


## rnorm

`rnorm` draws random samples from the distribution — the workhorse for *simulating*
data (we leaned on it heavily in the t-distribution chapter). Give it a count `n`
and it hands back that many random values:

```{r}
rnorm(5)
```

Run that again and you'll get five *different* numbers. To make a simulation
reproducible — so you and a reader see the same "random" draws — set a seed first:

```{r}
set.seed(42)
rnorm(5)
```

And just like the other functions, `mean` and `sd` shift and scale the draws:

```{r}
rnorm(5, mean = 100, sd = 15)   # e.g. five simulated IQ scores
```

```{r echo=FALSE}
# rnorm plot
# rnorm(5)

r1 <- ggplot(mapping = aes(x, y), data = normal) +
     geom_line(size = 1.3) +
     scale_x_continuous(n.breaks = 7) +
     labs(title = "rnorm(n=5)", subtitle = "R is for Random generation", 
          x = NULL, y = NULL, caption = "Output: c(-1.8, -0.8, 0.1, 0.45, 1.1)") +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 1.1, xend = 1.1,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 0.45, xend = 0.45,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = -0.8, xend = -0.8,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = 0.1, xend = 0.1,
              color = "blue", linetype = 2) +
     annotate(geom = "segment", y = 0, yend = 0.4, x = -1.8, xend = -1.8,
              color = "blue", linetype = 2) +
     theme_classic(base_size = 14, base_family = "Courier") +
     theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
     theme(plot.subtitle = element_text(face = "italic", hjust = 0.5)) +
     theme(plot.caption = element_text(face = "bold", size = 14, hjust = 0.5))
```


```{r}
#| label: fig-rnorm
#| fig-cap: "The rnorm function takes a number of samples and returns a vector of random numbers from the normal distribution (with mean=0, sd=1 as defaults)"
print(r1)
```


## Worked example: IQ scores

The normal distribution (also called the Gaussian distribution) is defined by two
numbers: the **mean** (µ), which sets the centre, and the **standard deviation**
(σ), which sets the spread. IQ scores are deliberately built to be normal with a
**mean of 100** and a **standard deviation of 15**, which makes them a perfect
playground for the four functions.

A handy rule of thumb — the "68–95–99.7 rule" — says about 68% of values fall
within one standard deviation of the mean (here, 85 to 115) and about 95% within
two (70 to 130). Let's check the 68% with `pnorm`. The area *between* 85 and 115 is
the area below 115 minus the area below 85:

```{r}
pnorm(115, mean = 100, sd = 15) - pnorm(85, mean = 100, sd = 15)
```

About 0.68 — exactly what the rule promises. Notice the pattern: a "between"
question becomes one `pnorm` minus another.

## Exercises

Use the IQ distribution (mean 100, sd 15) for each. Each solution names *which* of
the four functions fits the question — that choice is the real skill.

1. **Top 10%.** Which IQ score marks the 90th percentile — the cutoff above which
   only the top 10% fall?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   qnorm(0.9, mean = 100, sd = 15)
   ```
   We want a *value* from a *probability*, so this is a `qnorm` job. About 119.
   :::

2. **Above 130.** What fraction of people have an IQ above 130 (two standard
   deviations up)?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   1 - pnorm(130, mean = 100, sd = 15)
   # or, equivalently:
   pnorm(130, mean = 100, sd = 15, lower.tail = FALSE)
   ```
   About 2.3% — consistent with the 95% rule leaving roughly 2.5% in each tail.
   :::

3. **Below 70.** What fraction have an IQ below 70?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   pnorm(70, mean = 100, sd = 15)
   ```
   Also about 2.3%. The curve is symmetric, so the lower tail past 70 mirrors the
   upper tail past 130.
   :::

4. **Simulate a classroom.** Draw 30 random IQ scores and count how many land above
   115. (Set a seed so your answer is reproducible.)

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   set.seed(1)
   scores <- rnorm(30, mean = 100, sd = 15)
   sum(scores > 115)
   ```
   Theory predicts about 16% above 115 — roughly 5 of 30 — but any single small
   sample wobbles around that. Re-run with a different seed to feel the sampling
   variability for yourself.
   :::

## Summary

You now have the four-function toolkit for the normal distribution — and, thanks to
R's naming rule, the blueprint for every other distribution too:

- `pnorm(x)` — area to the **left** of `x` (a probability). Use `lower.tail = FALSE`
  or `1 - pnorm(x)` for the right tail, and one `pnorm` minus another for a range.
- `dnorm(x)` — the **height** of the curve at `x` (a density).
- `qnorm(p)` — the **value** at probability `p` (the inverse of `pnorm`); your tool
  for percentiles.
- `rnorm(n)` — `n` random **draws**; pair it with `set.seed()` for reproducible
  simulations.

Keep the pictures in mind: `pnorm` is an area, `dnorm` is a height, `qnorm` reads
the area axis backwards to a value, and `rnorm` scatters points along the curve.

    
