---
title: "Packages, simulation, and the law of large numbers"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r setup, include=FALSE}
set.seed(42)
```

In the [previous chapter](sampling_and_functions.qmd) we built `random_sequence()`, a tiny
generative model that spits out random DNA. We claimed each base is equally likely.
But is it? And what does a *collection* of these sequences actually look like? When
you can't answer a question by staring at code, two general-purpose tools come to the
rescue, and you'll use them together for the rest of your career:

- **Simulation** — generate the data many times over and look at the results in
  aggregate, and
- **Visualization** — draw a picture so the pattern jumps out.

Simulation we can do with base R. Visualization, in the style this book uses, needs a
**package** — so that's where we start.

## What you'll learn

- Install and load R **packages**, and tell those two steps apart.
- Reduce a whole dataset to a single number — a **statistic** — and write a function
  that computes one.
- Repeat a random experiment many times with `replicate()`.
- Visualize the spread of a statistic with a histogram (using **ggplot2**).
- State and *see* the **law of large numbers**: more data means a statistic settles
  toward its expected value.

## Packages

R on its own is powerful, but most of what makes it indispensable lives in
**packages**: bundles of functions, data, and documentation that other people have
written and shared. Rather than coding a histogram routine from scratch, you load a
package that already has one — well-tested, documented, and used by thousands.

### Installing vs. loading

There are two distinct steps, and conflating them is the single most common beginner
stumble.

**Installing** downloads a package onto your computer. You do it **once** (per machine,
per major update). Most packages come from CRAN, the central R package repository:

```{r eval=FALSE}
install.packages("ggplot2")
```

For bioinformatics packages you'll instead use **Bioconductor**, whose installer,
`BiocManager`, can also install regular CRAN packages. You install `BiocManager` itself
just once:

```{r eval=FALSE}
install.packages("BiocManager")
BiocManager::install("ggplot2")
```

**Loading** makes an installed package's functions available in your *current* R
session. You do it **every** session, with `library()`:

```{r}
library(ggplot2)
```

After that line, `ggplot2`'s functions are in scope and ready to use.

::: {.callout-tip}
## Install once, load every time
The classic analogy: **installing** a package is like screwing a light bulb into a
socket — you do it once and it stays. **Loading** it with `library()` is flicking the
switch — you do it every time you want the light, in every new R session. R forgets
its loaded packages when you quit, so `library()` calls live at the top of your
scripts and documents.
:::

## A statistic: one number that summarizes many

Our generative model claims A, C, G, and T are equally likely. A natural way to check
is **GC content**: the fraction of a sequence's bases that are G or C. It's a real and
useful quantity — GC content varies systematically across organisms and across regions
of a genome — and it's a perfect first example of a **statistic**.

A *statistic* is just a single number computed from data that summarizes some aspect
of it. The mean of a set of measurements is a statistic; so is GC content. Statistics
take a pile of numbers you can't hold in your head and boil it down to something you
can.

Let's write a function for it. We can reuse the `%in%` idea and the trick that the
mean of a logical vector is the fraction of `TRUE`s (see the
[Vectors chapter](vectors.qmd)):

```{r}
gc_content <- function(sequence) {
  bases <- strsplit(sequence, "")[[1]]   # split the string into single letters
  mean(bases %in% c("G", "C"))           # fraction of bases that are G or C
}
```

We also need our generator from last chapter. Each chapter stands on its own, so
let's redefine both pieces here:

```{r}
nucleotides <- c("A", "C", "G", "T")

random_sequence <- function(n = 100, prob = NULL) {
  bases <- sample(nucleotides, size = n, replace = TRUE, prob = prob)
  paste(bases, collapse = "")
}
```

Now compute the GC content of a single random sequence:

```{r}
seq1 <- random_sequence(n = 100)
gc_content(seq1)
```

You'll get something *near* 0.5, but probably not exactly — randomness wobbles. Run it
on another sequence and the number changes:

```{r}
gc_content(random_sequence(n = 100))
gc_content(random_sequence(n = 100))
```

A single statistic from a single sequence doesn't tell us much, precisely because it
jumps around. To see the real behavior of our model, we need *many* sequences.

## Repeating an experiment with `replicate()`

The `replicate()` function runs an expression many times and collects the results into
a vector. Read it as "do this `n` times and give me back all the answers." Here we
generate 1000 sequences and record the GC content of each:

```{r}
gc_values <- replicate(1000, gc_content(random_sequence(n = 100)))
length(gc_values)
head(gc_values)
```

`gc_values` is now a vector of 1000 statistics — one experiment's worth of evidence
about how our model behaves. Summaries help, but a picture helps more.

## Visualizing the spread

A **histogram** bins the values and draws a bar for how many fell in each bin,
revealing the shape of the distribution. With `ggplot2` we describe the plot in
layers: the data, an *aesthetic* mapping (which variable goes on the x-axis), and a
*geometry* (here, a histogram).

```{r}
#| label: fig-gc-hist
#| fig-cap: "Distribution of GC content across 1000 simulated 100-base sequences. The values cluster around 0.5, the expected GC content when all four bases are equally likely."
library(ggplot2)

ggplot(data.frame(gc = gc_values), aes(x = gc)) +
  geom_histogram(binwidth = 0.02, fill = "steelblue", color = "white") +
  labs(x = "GC content", y = "Number of sequences")
```

The values pile up around 0.5 — just as they should when every base is equally
likely — and fan out symmetrically on either side. That spread *is* the randomness in
our model made visible. (If `qplot()` is what you learned years ago, note that it's now
deprecated; `ggplot() + geom_histogram()` is the current idiom, and the
[ggplot2 chapter](ggplot2/intro.qmd) builds it up properly.)

## The law of large numbers

Here is the payoff. Each sequence's GC content is itself an average — the fraction of
G/C over the bases in that sequence. What happens to that average as we put **more
data** into each sequence, i.e. make the sequences longer?

Let's compute GC content for sequences of three different lengths, many times each, and
compare the spreads:

```{r}
#| label: fig-lln
#| fig-cap: "GC content for sequences of length 25, 100, and 1000 (1000 sequences each). As sequence length grows, the distribution of the statistic tightens around its expected value of 0.5 — the law of large numbers in action."
gc_by_length <- rbind(
  data.frame(length = "25",   gc = replicate(1000, gc_content(random_sequence(25)))),
  data.frame(length = "100",  gc = replicate(1000, gc_content(random_sequence(100)))),
  data.frame(length = "1000", gc = replicate(1000, gc_content(random_sequence(1000))))
)
gc_by_length$length <- factor(gc_by_length$length, levels = c("25", "100", "1000"))

ggplot(gc_by_length, aes(x = gc)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "white") +
  facet_wrap(~ length, ncol = 1) +
  labs(x = "GC content", y = "Number of sequences")
```

The center stays put at 0.5, but the **spread shrinks dramatically** as the sequences
get longer. A 25-base sequence can easily land at 0.3 or 0.7 by chance; a 1000-base
sequence almost never strays far from 0.5. More data, less wobble.

That is the **law of large numbers**: as you average over more and more independent
observations, the average converges toward the true expected value, and the chance of
a large deviation shrinks toward zero. Our expected GC content is 0.5 (G and C are two
of four equally likely bases), and longer sequences hug that value ever more tightly.

::: {.callout-important}
## What the law does and doesn't promise
The law of large numbers is about the **average**, not the individual draws. Single
bases never stop being random — extend a sequence and each *new* base is still a
coin-flip for G/C. What stabilizes is the *fraction* over the whole sequence. And the
convergence relies on our independence assumption from the
[last chapter](sampling_and_functions.qmd); in real DNA, where neighboring bases are correlated, the
*expected* GC content is still meaningful, but the simple "spread shrinks like this"
arithmetic no longer applies cleanly.
:::

## Changing the model: GC-rich and GC-poor genomes

Expected GC content of 0.5 was a consequence of our equal-probability assumption.
Real organisms span a wide range — the *Plasmodium falciparum* genome is about 80% AT,
while some *Streptomyces* exceed 70% GC. Our `prob` argument lets us model that. Weight
G and C four times as heavily as A and T and the whole distribution slides toward high
GC:

```{r}
#| label: fig-gc-rich
#| fig-cap: "GC content of 1000 simulated GC-rich sequences (G and C weighted 4x relative to A and T). The distribution now centers near 0.8 rather than 0.5 — the expected value tracks the probabilities you put into the model."
gc_rich <- replicate(
  1000,
  gc_content(random_sequence(n = 100, prob = c(1, 4, 4, 1)))  # A, C, G, T
)

ggplot(data.frame(gc = gc_rich), aes(x = gc)) +
  geom_histogram(binwidth = 0.02, fill = "darkorange", color = "white") +
  labs(x = "GC content", y = "Number of sequences")
```

The peak moves to roughly 0.8, exactly where the weights predict (the combined
probability of G or C is 8/10). The expected value of a statistic is set by the model's
assumptions — change the assumptions and the statistic follows.

## Exercises

1.  Write a one-line expression (no new function) that computes the mean GC content
    over 500 sequences of length 200.

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    mean(replicate(500, gc_content(random_sequence(200))))
    ```
    `replicate()` gathers 500 statistics into a vector and `mean()` reduces that vector
    to one number — a statistic computed from other statistics.
    :::

2.  Without running it, predict whether the histogram for length-2000 sequences would
    be wider or narrower than the length-1000 one in @fig-lln. Why?

    ::: {.callout-note collapse="true"}
    ## Solution
    Narrower. The law of large numbers says more data per sequence means less spread
    in the average, so doubling the length tightens the distribution further around 0.5.
    :::

3.  Use `prob` to build a strongly AT-rich generator (like *Plasmodium*), simulate 1000
    sequences of length 100, and report the approximate center of their GC-content
    distribution.

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    at_rich <- replicate(
      1000,
      gc_content(random_sequence(n = 100, prob = c(4, 1, 1, 4)))  # A, C, G, T
    )
    mean(at_rich)
    ```
    Weighting A and T four times as heavily puts the combined G/C probability at 2/10,
    so GC content centers near 0.2.
    :::

4.  `library()` errors with "there is no package called 'ggplot2'" on a fresh machine.
    What step was skipped, and what's the one-line fix?

    ::: {.callout-note collapse="true"}
    ## Solution
    The package was never **installed** on that machine. Run
    `install.packages("ggplot2")` (or `BiocManager::install("ggplot2")`) once, then
    `library(ggplot2)` will work in every session afterward.
    :::

## Summary

You now have the simulation-and-visualization loop that underlies a huge amount of
practical statistics:

- **Packages** extend R. **Install** them once (`install.packages()` /
  `BiocManager::install()`) and **load** them every session (`library()`).
- A **statistic** is a single number summarizing data; GC content is ours, and a few
  lines of R turn it into a reusable function.
- **`replicate()`** runs a random experiment many times, and a **histogram** shows how
  the resulting statistic is distributed.
- The **law of large numbers** says a statistic computed from more data clusters more
  tightly around its expected value — visible as the shrinking spread in @fig-lln — and
  that expected value is determined by the assumptions baked into your model.

These same two tools — repeat it many times, then look at the distribution — return in
the statistics and machine-learning chapters, where the "experiment" is a real analysis
rather than a string of random bases.
