---
title: "Random sampling and writing functions"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r setup, include=FALSE}
# Seed the random-number generator so every example below renders identically.
# Remove this line (or change the number) and you'll get a different—but equally
# valid—set of random sequences each time.
set.seed(42)
```

By now you can talk to R at the console, store values in objects, and call
functions that other people wrote (see the [R mechanics chapter](r_intro_mechanics.qmd)).
In this chapter we put those pieces together to build something small but real: a
little machine that **generates random DNA sequences**.

That goal is a good excuse to learn two ideas you'll use constantly. The first is
**random sampling** — asking R to draw values at random, which is the engine behind
every simulation in this book. The second is **writing your own functions** — packaging
a few lines of code under a name so you can reuse them with a single call. Every
later chapter assumes you can do this, so this is the place to get it under your fingers.

There's a bigger idea hiding in here, too. By the end you'll have written a tiny
**generative model** of DNA: a set of rules that can produce sequence data that never
existed before. It is a crude model — deliberately so — but it is a genuine cousin,
in spirit, of the DNA language models now used in genomics research. Those models
learn fantastically complex rules about how bases depend on one another; ours will
use one almost laughably simple rule. Seeing the simple version first makes the
fancy version far less mysterious.

## What you'll learn

- Draw values at random with `sample()`, and read its arguments.
- Explain **sampling with and without replacement**, and why DNA needs *with*.
- Recognize the **modeling assumptions** baked into a simulation — and where they
  break down in real biology.
- Write your own **function**: give it a name, a body, arguments, and defaults.
- See how a few simple rules add up to a (very basic) **generative model**.

## The four letters of DNA

DNA is written in an alphabet of four letters — the nucleotide bases adenine,
cytosine, guanine, and thymine. We can hold that alphabet in a **character vector**:

```{r}
nucleotides <- c("A", "C", "G", "T")
nucleotides
```

That's it — a four-element vector, each element a one-letter string. If `c()`,
quoting, and the idea of a vector feel shaky, the [Vectors chapter](vectors.qmd)
covers them in depth; here we only need this one small vector to get going.

## Drawing a base at random

To "read" a base off this alphabet at random, we use the `sample()` function.
Like every function, we call it by writing its name followed by parentheses, with
the data it should work on inside. Give `sample()` a vector and it returns one of
the elements, chosen at random:

```{r}
sample(nucleotides, size = 1)
```

Run that again and you'll (usually) get a different letter — that's the point. The
`size` argument says how many elements to draw. Ask for a few:

```{r}
sample(nucleotides, size = 3)
```

::: {.callout-tip}
## Arguments have names
`size` is the *name* of an argument, and writing `size = 3` makes the call easy to
read. You can look up the arguments a function accepts with `help(sample)` or, more
briefly, `?sample`. Getting in the habit of reading help pages is one of the highest-
return skills in R.
:::

## With or without replacement

Now try to draw a whole sequence — say, ten bases:

```{r error=TRUE}
sample(nucleotides, size = 10)
```

R refuses, and the error message tells you why: by default `sample()` draws
**without replacement**. Picture the four letters as tiles in a bag. Drawing without
replacement means each tile, once pulled, is set aside and can't be drawn again — so
from a bag of four tiles you can pull at most four, and you'd just get the same four
letters in a shuffled order. That is nothing like DNA, where the same base appears
over and over.

What we want is sampling **with replacement**: pull a tile, write down its letter,
then *drop it back in the bag* before the next draw. Every draw faces the full bag of
four, so letters can — and do — repeat. We switch this on with `replace = TRUE`:

```{r}
sample(nucleotides, size = 10, replace = TRUE)
```

Now we can draw sequences of any length we like.

::: {.callout-important}
## The assumption hiding in `replace = TRUE`
Sampling with replacement, the way we're doing it, makes every position **independent**
of every other: knowing that position 5 is a `G` tells you nothing about position 6.
That is a *modeling assumption*, and in real DNA it is plainly false. Bases in a
coding sequence come in three-letter codons; promoters and binding sites have
recurring motifs; whole regions are GC-rich or AT-rich. Real genomes are full of
local structure that our independent-draws model ignores entirely.

That doesn't make the model useless — simple models are often the right starting
point, and a deliberately wrong baseline is the thing more realistic models get
measured against. But always keep the assumption in view, because the conclusions
you draw from a simulation are only as good as the assumptions you fed into it.
:::

## From letters to a sequence

A draw like the one above is a *vector* of single letters. A biologist usually wants
the sequence as one string, `"ACGT…"`. We glue the letters together with `paste()`,
using `collapse = ""` to join them with nothing in between:

```{r}
bases <- sample(nucleotides, size = 10, replace = TRUE)
bases
paste(bases, collapse = "")
```

`paste()` is one of the workhorse string functions; it and its relatives get a fuller
treatment in the [Vectors chapter](vectors.qmd). For now it's just the last step in
turning random draws into something that looks like DNA.

## Writing your own function

We now have a recipe — sample with replacement, then paste — that we'll want to run
again and again. Retyping it every time is tedious and error-prone. The fix is to
wrap the recipe in a **function** of our own.

You already *call* functions all the time (`sample()`, `paste()`, `mean()`). Writing
one just means bottling up some code under a new name. Every R function has three
parts:

- a **name** (so you can call it later),
- a **body** (the code that runs), and
- a set of **arguments** (the inputs you can vary each time).

You build one with the `function` keyword, put the body between braces `{ }`, and
save the result to a name with `<-`:

```{r}
random_sequence <- function(n) {
  bases <- sample(nucleotides, size = n, replace = TRUE)
  paste(bases, collapse = "")
}
```

Read that as: "`random_sequence` is a function of one argument, `n`; it draws `n`
bases with replacement and pastes them into a single string." A function returns the
value of its **last line**, so `random_sequence()` hands back the pasted string.

Call it the same way you call any function — name, then parentheses, with a value for
`n`:

```{r}
random_sequence(n = 10)
random_sequence(n = 30)
```

Each call re-runs the body, so each call gives a fresh sequence. We've turned three
lines of fiddly code into one reusable verb.

::: {.callout-tip}
## A function is just another object
`random_sequence` is an object, exactly like `nucleotides` is. Type its name *without*
parentheses and R shows you the code it contains; type it *with* parentheses and R
runs that code. The parentheses are the trigger.
:::

### Default arguments

Right now you must supply `n` every time, or you get an error. Often it's convenient
to give an argument a **default** — a value R uses when you don't specify one. Set the
default in the function definition with `=`:

```{r}
random_sequence <- function(n = 20) {
  bases <- sample(nucleotides, size = n, replace = TRUE)
  paste(bases, collapse = "")
}
```

Now a bare call produces a 20-base sequence, but you can still override the default
when you want a different length:

```{r}
random_sequence()
random_sequence(n = 50)
```

### A second argument: base composition

So far every base is equally likely — each of A, C, G, T turns up about a quarter of
the time. Real genomes are lopsided: some organisms are strikingly GC-poor (the
malaria parasite *Plasmodium falciparum*) and others GC-rich (many *Streptomyces*).
The `sample()` function can model that through its `prob` argument, which takes a
vector of relative probabilities, one per element of the alphabet. Let's expose that
as an argument of our own function:

```{r}
random_sequence <- function(n = 20, prob = NULL) {
  bases <- sample(nucleotides, size = n, replace = TRUE, prob = prob)
  paste(bases, collapse = "")
}
```

The default `prob = NULL` tells `sample()` to fall back on equal probabilities, so the
old behavior is unchanged. But now we can ask for, say, a GC-rich sequence by making
G and C more likely than A and T (remember the order is A, C, G, T):

```{r}
# A, C, G, T weighted 1 : 4 : 4 : 1  ->  mostly G and C
random_sequence(n = 50, prob = c(1, 4, 4, 1))
```

Count the letters and you should see far more `G`s and `C`s than `A`s and `T`s.

## A (very simple) generative model

Step back and look at what we built. `random_sequence()` is a small **generative
model** of DNA: a set of rules that can manufacture new sequence data. Its rules are
just two:

1. each base is drawn from the four-letter alphabet, and
2. the draws are independent, each following a fixed probability distribution
   (uniform by default, or whatever `prob` you supply).

That is about the simplest generative model of DNA imaginable. The large DNA
**language models** used in modern genomics are generative models too, and in spirit
they do the same job — produce, or assign probabilities to, sequence. The difference
is in rule 2. Where our model treats every position as independent, a language model
makes each base depend on the entire context around it, capturing codons, motifs,
regulatory grammar, and long-range structure that ours throws away. They have
millions or billions of learned parameters; ours has at most four. Same idea, wildly
different complexity — and you now understand the simple end of that spectrum from the
inside.

In the [next chapter](packages_and_simulation.qmd) we'll put this model to work:
generate *many* sequences, summarize each with a single number, and watch a beautiful
piece of statistics — the law of large numbers — emerge.

## Exercises

Each solution names the key idea and a one-line *why*, because the reasoning transfers
further than the answer.

1.  Use `sample()` directly (no custom function) to draw a single random codon — three
    bases — as a character vector.

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    sample(nucleotides, size = 3, replace = TRUE)
    ```
    `size = 3` asks for three draws, and `replace = TRUE` lets a base repeat — a codon
    like `"TTT"` is perfectly legal, so we *must* sample with replacement.
    :::

2.  Explain in one sentence why `sample(nucleotides, size = 6)` (no `replace`) throws
    an error, but `sample(nucleotides, size = 6, replace = TRUE)` does not.

    ::: {.callout-note collapse="true"}
    ## Solution
    Without replacement there are only four tiles to draw from, so you can't pull six;
    with replacement each tile goes back in the bag after every draw, so the supply
    never runs out.
    :::

3.  Write a function `random_codon()` that returns a single three-base string (e.g.
    `"GAT"`). Reuse the pattern from `random_sequence()`.

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    random_codon <- function() {
      bases <- sample(nucleotides, size = 3, replace = TRUE)
      paste(bases, collapse = "")
    }
    random_codon()
    ```
    Same three parts as before — name, body, (here empty) arguments — just with `size`
    fixed at 3 and no inputs to vary.
    :::

4.  Our model assumes every base position is independent. Name one real biological
    feature of DNA that violates that assumption, and say in a sentence why it matters.

    ::: {.callout-note collapse="true"}
    ## Solution
    Many answers work: codons (bases come in functional triplets), sequence motifs in
    promoters and binding sites, CpG islands, or GC-rich/AT-rich regions. Any of these
    means the base at one position carries information about its neighbors — exactly
    the dependence our independent-draws model ignores, which is why sequences it
    generates won't contain genuine genes or regulatory signals.
    :::

5.  Modify a copy of `random_sequence()` so that it returns the *vector* of bases
    instead of a single pasted string. Which line do you remove?

    ::: {.callout-note collapse="true"}
    ## Solution
    ```{r}
    random_bases <- function(n = 20, prob = NULL) {
      sample(nucleotides, size = n, replace = TRUE, prob = prob)
    }
    random_bases(n = 8)
    ```
    Drop the `paste()` line; a function returns its last expression, so making the
    `sample()` call the last line returns the raw vector.
    :::

## Summary

You can now generate random data and package code for reuse — two skills the rest of
the book leans on constantly:

- **`sample()`** draws values at random from a vector; `size` sets how many and
  `replace = TRUE` allows repeats. Building DNA requires sampling *with* replacement,
  which models every position as **independent** — a convenient fiction that real
  genomic structure violates.
- **You write a function** with `function(args) { body }`, save it with `<-`, and it
  returns the value of its last line. Arguments can have **defaults**, and extra
  arguments (like `prob`) let one function cover many cases.
- Strung together, a couple of simple rules make a **generative model** — here, a
  bare-bones one that is, in spirit, the simplest relative of the DNA language models
  used in research.

Next we'll generate sequences by the thousands, reduce each to a single summary
number — a **statistic** — and meet the **law of large numbers**.
