---
title: "Ranges Exercises"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
format:
  html:
    code-fold: true
    code-summary: "Show the answer"
    code-tools: true
---

```{r init, include=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE, message=FALSE, R.options=list(max.print=50))
```

The two ranges chapters — *Genomic ranges and features* and *Genomic ranges with
real ChIP-seq peaks* — walked you through the `GenomicRanges` toolkit on small
hand-built objects and a single real peak set. This chapter is their hands-on
companion. Here you'll point those same tools at *real* ENCODE data and answer
the kinds of questions a genomicist actually asks: How many peaks did this
experiment call? How wide are they? Do two replicates agree? Are open-chromatin
regions enriched near gene promoters? Is one set of histone marks enriched at CpG
islands?

The data comes from the [ENCODE project](https://www.encodeproject.org/) via the
`AnnotationHub` package. ENCODE published these datasets as BED-style files;
AnnotationHub has packaged them into ready-to-use `GRanges` objects, so you can
load one with a single line instead of downloading and parsing a file. (If you
ever have a peak file of your own — peak calls, enhancer regions, and so on — you
can read it into a `GRanges` the same way the ChIP-seq chapter does, with
`rtracklayer::import()`.)

Each exercise below is a short prompt followed by a folded solution. Try the
prompt yourself first; then click **Show the answer** to reveal the code *and* its
output, with a sentence or two on what to notice.

## What you'll learn

- Query `AnnotationHub` to find and load a published dataset as a `GRanges`.
- Inspect a `GRanges` — its class, its metadata columns, and its per-chromosome
  peak counts.
- Measure and summarize peak widths, and compute how much of the genome a peak set
  covers.
- Read and switch the `seqlevelsStyle` of an object (the `chr1` vs `1` naming
  convention) and recognize the silent bug a mismatch causes.
- Compute overlaps between two range sets with `%over%`, and reshape ranges with
  `resize()`, `flank()`, `promoters()`, and `reduce()`.
- Quantify how enriched one set of ranges is inside another with intersect/union.

## Exercise 1: explore a DNase peak set

DNase-seq finds **open chromatin** — stretches of DNA that are accessible to the
DNase enzyme, which is where regulatory machinery tends to sit. In this exercise
you'll load one DNase hypersensitivity peak set from the K562 cell line and get
to know it as a `GRanges` object.

**Find and load the peak set.** Use `AnnotationHub` to find the record
`goldenpath/hg19/encodeDCC/wgEncodeUwDnase/wgEncodeUwDnaseK562PkRep1.narrowPeak.gz`
and load it into a variable named `dnase`.

```{r dnase}
library(AnnotationHub)
# GenomicRanges gives us the GRanges accessors and overlap/resize operators;
# GenomeInfoDb provides the seqlevels()/seqlevelsStyle() machinery used below.
library(GenomicRanges)
library(GenomeInfoDb)
ah <- AnnotationHub()
query(ah, "goldenpath/hg19/encodeDCC/wgEncodeUwDnase/wgEncodeUwDnaseK562PkRep1.narrowPeak.gz")
# The query above matches exactly one record, so we grab it with [[1]].
dnase <- query(ah, "goldenpath/hg19/encodeDCC/wgEncodeUwDnase/wgEncodeUwDnaseK562PkRep1.narrowPeak.gz")[[1]]
```

The `query()` call returns a one-row hub listing, and `[[1]]` downloads that one
record and hands you the object itself. AnnotationHub caches the download, so a
second `[[1]]` is instant.

**What kind of object did you get?** Print it and ask for its class.

```{r}
dnase
class(dnase)
```

It's a `GRanges` — the same class you built by hand in the ranges chapter, only
this one was pulled from a public archive. Printing it shows the seqnames, ranges,
strand, and a block of metadata columns.

**What metadata is attached?** The per-range metadata lives in the `mcols`.

```{r}
head(mcols(dnase))
```

These columns are the extra fields from the original narrowPeak file — things like
the signal value and score for each peak. They ride along with every range so you
never lose the link between a location and its data.

**How many peaks are on each chromosome?** `seqnames()` returns the chromosome of
every peak; `table()` tallies them.

```{r}
# Show the busiest 25 chromosomes; the full list includes many empty contigs.
sort(table(seqnames(dnase)), decreasing = TRUE)[1:25]
```

Each count is the number of DNase peaks called on that chromosome. Bigger
chromosomes generally carry more peaks, but accessibility varies by chromosome too.

**How wide are the peaks?** Summarize the `width()` of every range.

```{r}
summary(width(dnase))
```

The summary gives you the min, median, mean, and max peak width in base pairs. A
tight distribution around a couple hundred bp is typical for a DNase peak caller.

**What sequence names were used — `chr1` or just `1`?** Look at the `seqlevels`
and the `seqlevelsStyle`, then switch the style to Ensembl and look again.

```{r}
head(seqlevels(dnase), 25)   # first 25 of many contigs
length(seqlevels(dnase))     # how many sequence names in all
seqlevelsStyle(dnase)
seqlevelsStyle(dnase) <- "ensembl"
seqlevelsStyle(dnase)
head(seqlevels(dnase), 25)   # same chromosomes, now in Ensembl style
```

The object started in **UCSC** style (`chr1`, `chr2`, …). Assigning a new
`seqlevelsStyle` rewrites every sequence name in place — after the switch the same
chromosomes are now named `1`, `2`, … in **Ensembl** style. Nothing about the
ranges changed, only the labels.

::: {.callout-warning}
## The `chr` gotcha that silently returns zero overlaps

`seqlevelsStyle` matters the moment you compare two objects. Overlap
operations match on the *exact* sequence name, so a peak on `chr1` will **never**
overlap a region on `1` — R reports zero overlaps and raises no error. If an
overlap count comes back suspiciously at zero, the first thing to check is whether
both objects share the same `seqlevelsStyle`. We switched to Ensembl here to
practice; in Exercise 3 we'll switch back before comparing against UCSC-style gene
models.
:::

**How much of the genome do the peaks cover?** Assuming the peaks don't overlap,
sum their widths and divide by the total genome length.

```{r}
sum(width(dnase))
sum(seqlengths(dnase))
sum(width(dnase)) / sum(seqlengths(dnase))
```

The first number is the total base pairs of open chromatin called; the second is
the size of the genome. Their ratio — a small fraction of a percent — tells you
that accessible chromatin is a tiny slice of the whole genome, which is exactly
why these regions are interesting.

## Exercise 2: compare two replicates

A good experiment is reproducible. ENCODE ran the DNase assay twice on K562, so we
have a second replicate. In this exercise you'll load it and ask how well the two
replicates agree.

**Load the second replicate** (`...wgEncodeUwDnaseK562PkRep2.narrowPeak.gz`) as
`dnase2`.

```{r}
query(ah, "goldenpath/hg19/encodeDCC/wgEncodeUwDnase/wgEncodeUwDnaseK562PkRep2.narrowPeak.gz")
# Again, a single matching record — grab it with [[1]].
dnase2 <- query(ah, "goldenpath/hg19/encodeDCC/wgEncodeUwDnase/wgEncodeUwDnaseK562PkRep2.narrowPeak.gz")[[1]]
```

**How many peaks are in each replicate?** Compare the lengths.

```{r}
length(dnase)
length(dnase2)
```

The two replicates call a broadly similar number of peaks. A wildly different
count between replicates would be a red flag that something went wrong in one of
the experiments.

**How wide are the `dnase2` peaks?**

```{r}
summary(width(dnase2))
```

The width distribution looks much like the first replicate's — another reassuring
sign of reproducibility.

**What proportion of the genome does `dnase2` cover?** Same calculation as before.

```{r}
sum(width(dnase2)) / sum(seqlengths(dnase2))
```

A comparable coverage fraction to replicate 1, as you'd hope for two runs of the
same assay.

**How many `dnase` peaks overlap a `dnase2` peak?** The `%over%` operator returns
a logical vector — `TRUE` for each `dnase` peak that hits at least one `dnase2`
peak — so summing it counts the overlapping peaks.

```{r}
sum(dnase %over% dnase2)
```

A high overlap count is the quantitative version of "the replicates agree": the
same regions of open chromatin show up both times. This is the kind of number you
would report when arguing that a peak set is trustworthy.

**Expand every peak by 50 bp on each end.** Suppose your peak caller was too
strict and you want peaks 100 bp wider. Use `resize()` with `fix="center"` so the
expansion is symmetric, and call the result `dnase_wide`.

```{r}
w <- width(dnase)
dnase_wide <- resize(dnase, width = w + 100, fix = "center")
head(width(dnase_wide))             # first few widened peaks
all(width(dnase_wide) == w + 100)   # every peak is exactly 100 bp wider
```

`resize()` sets each range to a new width; `fix="center"` keeps the midpoint fixed
so the peak grows 50 bp in each direction. The `all(...)` check confirms every
peak ended up exactly 100 bp larger.

## Exercise 3: DNase sites relative to gene promoters

Open chromatin should cluster near where genes are regulated — especially at
**promoters**, the regions just upstream of a gene's start. In this exercise
you'll build promoter regions from a gene model and ask whether the DNase sites
land there more than chance would predict.

You'll need the `TxDb.Hsapiens.UCSC.hg19.knownGene` transcript database. If it
isn't already installed, install it once (this is a one-time setup step, not part
of the analysis):

```r
# one-time setup, not part of the analysis
BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene")
```

**Load the genes from the TxDb.** Attach the package, grab the `TxDb` object, and
pull out the genes. What class is the result, and how many genes are there?

```{r}
library("TxDb.Hsapiens.UCSC.hg19.knownGene")
kg <- TxDb.Hsapiens.UCSC.hg19.knownGene
gx <- genes(kg)
class(gx)
length(gx)
```

`genes()` returns a `GRanges` — one range per gene, in UCSC (`chr1`) style to match
the genome it was built on. The length tells you how many genes the annotation
contains.

**Get promoter regions with `flank()`.** Read `?flank`. The promoter sits *upstream*
of a gene, so flanking each gene by 2 kb on its upstream side gives a first
approximation.

```{r}
head(flank(gx, 2000))
```

`flank()` returns a new range immediately upstream of each gene — strand-aware, so
"upstream" means the correct direction for genes on either strand. These are
2 kb-wide candidate promoter regions.

**Do the same thing with `promoters()`.** The TxDb knows where transcripts start,
so `promoters()` builds promoter regions directly (see `?promoters` for the default
window).

```{r}
proms <- promoters(kg)
```

`promoters()` is the purpose-built tool: it returns a promoter window around each
transcript's start site without you having to reason about strand and flanking
yourself.

**Do these promoters overlap each other?** Genes are dense, and nearby
transcripts share promoter neighborhoods, so many of these windows will overlap.
`countOverlaps()` of the set against itself counts how many promoters each one
touches.

```{r}
summary(countOverlaps(proms))
```

Every promoter overlaps at least itself (count ≥ 1), and the larger counts show
that promoters pile up on top of one another — overlapping transcripts and
alternative start sites. That redundancy is a problem if we want clean,
non-double-counted regions.

**Collapse overlapping promoters with `reduce()`.** `reduce()` merges every group
of overlapping ranges into a single contiguous region, giving us a tidy set of
promoter regions to test against.

```{r}
# reduce() takes all overlapping ranges and collapses each cluster
# into one region spanning the whole cluster.
prom_regions <- reduce(proms)

# After reducing, each region overlaps only itself.
summary(countOverlaps(prom_regions))
```

After `reduce()`, the self-overlap count is 1 everywhere — confirming the regions
no longer overlap each other and can be treated as a clean, non-redundant set.

**Count DNase sites that overlap a promoter region.** Here is where the
`seqlevelsStyle` warning above comes due: we switched `dnase` to Ensembl (`1`) in
Exercise 1, but the gene models are UCSC (`chr1`). Watch what happens.

```{r}
# dnase is still in Ensembl style from Exercise 1, while
# prom_regions is UCSC style — so this first count is misleadingly zero.
sum(dnase %over% prom_regions)

# Switch dnase back to UCSC so the sequence names match, then recount.
seqlevelsStyle(dnase) <- "UCSC"
sum(dnase %over% prom_regions)
sum(dnase2 %over% prom_regions)
```

The first count is **zero** — not because the data disagree, but because `chr1`
and `1` never match. After switching `dnase` back to UCSC style, the same overlap
query returns a large, sensible count. This is the silent bug from the warning,
caught and fixed. (`dnase2` was never reassigned, so it was UCSC-style all along.)

**Is that overlap surprising?** If DNase sites and promoters were placed
*independently* on the genome, how many overlaps would you expect by chance?
Compare that expectation to what you observed.

```{r}
# Use one consistent genome size. Some contigs have unknown (NA) lengths,
# so drop those when summing.
genome_bp <- sum(as.numeric(seqlengths(gx)), na.rm = TRUE)
prop_proms <- sum(width(prom_regions)) / genome_bp
# Under independence, the expected number of DNase peaks landing in a promoter
# is (fraction of the genome that is promoter) times the number of peaks.
prop_proms * length(dnase)
```

The expected-by-chance number is far smaller than the count you actually observed.
That gap is the biology: DNase hypersensitive sites are **not** scattered at
random — they concentrate at promoters, exactly where regulatory machinery opens
up the chromatin. The overlap count is meaningful precisely because it dwarfs the
null expectation.

## Exercise 4: enrichment of histone marks at CpG islands

Histone modifications mark functional regions of the genome: `H3K4me3` flags active
promoters, `H3K4me1` flags enhancers, `H3K27me3` flags repressed regions, and so
on. In this exercise you'll ask how much each mark **overlaps CpG islands** — and
build a small reusable enrichment score along the way.

The data again comes from `AnnotationHub`. ENCODE published these histone ChIP-seq
peak calls as BED files; the hub serves them as `GRanges`. Each mark's query
matches several records, so to keep things moving we've already picked the right
hub IDs and you can load each mark directly.

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
```

Here are the `GRanges` for four histone marks plus one extra `H3K4me3` replicate:

```{r}
h3k4me1 <- ah[["AH25832"]]
h3k4me3 <- ah[["AH25833"]]
h3k9ac <- ah[["AH25834"]]
h3k27me3 <- ah[["AH25835"]]
h3k4me3_2 <- ah[["AH27284"]]
```

Each variable holds the peak calls from one ChIP-seq experiment pulling down that
histone mark. Now grab the hg19 CpG islands the same way:

```{r}
cpg <- query(ah, c("cpg", "UCSC", "hg19"))[[1]]
```

**How many `H3K4me1` peaks overlap a CpG island?** Start with the simple `%over%`
count, as in Exercise 2.

```{r}
sum(h3k4me1 %over% cpg)
```

That's the number of `H3K4me1` peaks touching at least one CpG island — a quick
first look at the association.

**How many *bases* overlap?** A peak count ignores how much of each peak overlaps.
`intersect()` returns the overlapping regions themselves, so summing their widths
gives the total overlapping base pairs.

```{r}
# intersect() returns only the overlapping stretches as a new range set;
# summing their widths gives the total bases of overlap.
sum(width(intersect(h3k4me1, cpg)))
```

This is a finer measure than the peak count: it tells you how many base pairs the
two sets actually share, not just how many peaks touch.

**Turn that into an enrichment score.** Broad marks cover more bases and so overlap
CpG islands more just by being big. To correct for that, normalize the shared bases
by the *total* bases covered by either set (their `union`). The result is a
Jaccard-style ratio between 0 and 1: the fraction of the combined footprint that
both sets share.

```{r}
sum(width(union(h3k4me1, cpg)))
# Normalize the overlap by the combined coverage of both sets.
sum(width(intersect(h3k4me1, cpg))) / sum(width(union(h3k4me1, cpg)))
```

A score near 0 means the two sets barely co-occur; a score near 1 means they cover
almost exactly the same ground. Because the denominator grows with peak breadth,
this score lets you compare a broad mark and a narrow mark on equal footing.

**Wrap it in a function** so you can score any pair of range sets.

```{r}
range_enrichment_score <- function(r1, r2) {
  i <- sum(width(intersect(r1, r2)))
  u <- sum(width(union(r1, r2)))
  return(i / u)
}
```

The function is just the intersect-over-union calculation with names: `i` for the
shared bases, `u` for the combined footprint. Give it a try on `H3K4me1` and the
CpG islands:

```{r}
range_enrichment_score(h3k4me1, cpg)
```

You get back the same normalized score as before. Now you can call
`range_enrichment_score()` on any two `GRanges` — try the other histone marks
against `cpg` and see which one is most enriched at CpG islands.

## Summary

You've now run the core range-analysis workflow end to end on real ENCODE data.
You can:

- **Find and load** a published dataset from `AnnotationHub` as a `GRanges`, and
  inspect its class, metadata, and per-chromosome peak counts.
- **Measure** peak widths and compute the fraction of the genome a peak set covers.
- **Read and switch `seqlevelsStyle`**, and — just as important — recognize the
  silent zero-overlap bug a `chr1`-vs-`1` mismatch causes.
- **Compare range sets** with `%over%`, and **reshape** ranges with `resize()`,
  `flank()`, `promoters()`, and `reduce()`.
- **Quantify enrichment** of one set inside another with an intersect-over-union
  score.

For the underlying concepts — how `GRanges` objects are built, the full catalog of
set operations and overlap methods, and the coordinate-system details — return to
the *Genomic ranges and features* chapter. For the real-data import workflow with
`rtracklayer`, see *Genomic ranges with real ChIP-seq peaks*. This chapter was the
place to put those tools to work.
