---
title: "Ranges Exercises"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r init, include=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE, message=FALSE, R.options=list(max.print=50))
```

The two ranges chapters — *Genomic ranges and features* and *Genomic ranges with
real ChIP-seq peaks* — walked you through the `GenomicRanges` toolkit, first on
small hand-built objects and then on a real CTCF peak set. This chapter is their
hands-on companion: a set of short problems that ask you to *use* those tools
rather than read about them.

The exercises come in two halves. The first half (Exercises 1–3) stays entirely
on small `GRanges` you build by hand, so you can run them anywhere with nothing
but `GenomicRanges` loaded — no downloads, no annotation packages. These drill the
core mechanics: construction and accessors, the intra- and inter-range
operations, and the `chr1`-vs-`1` naming trap that quietly breaks overlaps. The
second half (Exercises 4–5) turns you loose on the **real** ENCODE CTCF peaks from
the ChIP-seq chapter and a human gene model, to answer the kind of question a
genomicist actually asks: which peaks sit in a promoter, and which gene is each
peak closest to?

Each exercise is a short prompt. Try it yourself first; then expand the
**Solution** box to reveal the code, its output, and a sentence or two on what to
notice.

## What you'll learn

- Build a `GRanges` from scratch and pull it apart with the coordinate accessors
  (`seqnames()`, `start()`, `width()`, `strand()`) and `mcols()`.
- Reshape ranges with the intra-range operations (`flank()`, `shift()`,
  `resize()`) and summarize a set with the inter-range operations (`reduce()`,
  `disjoin()`, `gaps()`, `coverage()`).
- Relate two range sets with `findOverlaps()`, `%over%`, `subsetByOverlaps()`,
  `countOverlaps()`, `nearest()`, and `distanceToNearest()`.
- Recognize the silent zero-overlap bug a `chr1`-vs-`1` `seqlevelsStyle` mismatch
  causes, and fix it.
- Apply the whole toolkit to a real ENCODE peak set and a `TxDb` gene model.

```{r setup-lib}
library(GenomicRanges)
```

## Exercise 1: build a GRanges and take it apart

Before analysing anyone else's data, it pays to be fluent in *constructing* a
`GRanges` and reading every piece back out. In this exercise you'll build a small
catalogue of imaginary gene features by hand and practice the accessors.

**Build the object.** Construct a `GRanges` named `feat` with five ranges on
`chr1`, with these starts and ends, strands, and a `gene` metadata column:

| name | start | end  | strand | gene  |
|------|-------|------|--------|-------|
| f1   | 100   | 250  | +      | AAA   |
| f2   | 300   | 470  | +      | BBB   |
| f3   | 600   | 760  | -      | CCC   |
| f4   | 720   | 900  | -      | DDD   |
| f5   | 1000  | 1180 | +      | EEE   |

::: {.callout-note collapse="true"}
## Solution

```{r ex1-build}
feat <- GRanges(
  seqnames = "chr1",
  ranges   = IRanges(start = c(100, 300, 600, 720, 1000),
                     end   = c(250, 470, 760, 900, 1180)),
  strand   = c("+", "+", "-", "-", "+"),
  gene     = c("AAA", "BBB", "CCC", "DDD", "EEE"))
names(feat) <- c("f1", "f2", "f3", "f4", "f5")
feat
```

The `GRanges()` constructor takes the coordinate pieces by name; any extra named
argument (`gene` here) becomes a metadata column on the right of the `|`. A single
`seqnames = "chr1"` is recycled to all five ranges.
:::

**What's on the left of the `|`, and what's on the right?** Pull out the
`seqnames`, the `strand`, the `width` of each range, and then the metadata.

::: {.callout-note collapse="true"}
## Solution

```{r ex1-accessors}
seqnames(feat)
strand(feat)
width(feat)        # end - start + 1, because GRanges are 1-based and inclusive
mcols(feat)        # everything on the right of the |
mcols(feat)$gene   # one metadata column
```

The coordinate parts each have their own accessor; everything you attached lives
in `mcols()`. Note `width(feat)[1]` is 151, not 150 — both endpoints count.
:::

**Subset two ways.** Keep only the genes on the `+` strand, and — separately —
keep only the ranges wider than 160 bp.

::: {.callout-note collapse="true"}
## Solution

```{r ex1-subset}
feat[strand(feat) == "+"]
feat[width(feat) > 160]
```

A `GRanges` subsets like a vector: any logical vector the same length as the
object selects ranges. Here `strand(feat) == "+"` and `width(feat) > 160` are each
such a vector.
:::

## Exercise 2: reshape and summarize one range set

This exercise stays with `feat` and works through the single-set operations: the
*intra-range* ones that reshape each range on its own, and the *inter-range* ones
that look at the whole set together.

**Promoters with `flank()`.** Grab the 100 bp immediately *upstream* of each
feature. Then look carefully at `f3`/`f4` (the `-`-strand genes) and explain where
their flanks landed.

::: {.callout-note collapse="true"}
## Solution

```{r ex2-flank}
proms <- flank(feat, 100)
proms
```

`flank()` is strand-aware. For the `+` genes the upstream flank sits at *lower*
coordinates (just before the start); for the `-` genes `f3` and `f4`, "upstream"
runs the other way, so their flanks sit at *higher* coordinates, just past their
ends. Every flank is 100 bp wide — only the side changes.
:::

**Standardize the widths with `resize()`.** Make every feature a 200 bp window
centered on its midpoint, and confirm the widths.

::: {.callout-note collapse="true"}
## Solution

```{r ex2-resize}
win <- resize(feat, width = 200, fix = "center")
width(win)
all(width(win) == 200)
```

`fix = "center"` holds each range's midpoint still and grows or shrinks it to the
requested width — the standard way to build comparable fixed-width windows.
:::

**Collapse overlaps with `reduce()`.** Features `f3` (600–760) and `f4` (720–900)
overlap. Merge any overlapping or touching ranges into a non-redundant set, then
report how many ranges survive. (Try it once with the strands as-is, and once with
`ignore.strand = TRUE`, and explain the difference.)

::: {.callout-note collapse="true"}
## Solution

```{r ex2-reduce}
reduce(feat)
length(reduce(feat))
reduce(feat, ignore.strand = TRUE)
```

`reduce()` merges by chromosome *and strand* by default, so `f3` and `f4` — both
on `-` — collapse into one range, leaving four. Because none of the ranges sit on
opposite strands in the same place, `ignore.strand = TRUE` gives the same merge
here, but in general it would also fuse ranges that overlap across strands.
:::

**Per-base coverage.** Compute the `coverage()` of `feat` and find the single
highest depth anywhere on `chr1`. Where does it occur, and why?

::: {.callout-note collapse="true"}
## Solution

```{r ex2-coverage}
cov <- coverage(feat)
cov                       # an RleList, one element per chromosome
max(cov$chr1)             # the tallest pile-up
```

`coverage()` counts, base by base, how many ranges cover each position. The
maximum is 2, in the 720–760 stretch where `f3` and `f4` overlap; everywhere else
at most one feature is present, so the depth is 1 (or 0 in the gaps).
:::

## Exercise 3: the `chr1`-vs-`1` overlap trap

Overlap operations match on the *exact* sequence name, so a range on `chr1` will
never overlap a range on `1` — and R reports zero overlaps with no error or
warning. This exercise reproduces that silent bug deliberately on two tiny
hand-built objects, then fixes it, so you recognize it when it bites you for real.

**Build two overlapping sets in different naming styles.** Build `ucsc` with two
ranges on `"chr1"`, and `ens` with two ranges on `"1"` that clearly overlap them by
coordinate.

::: {.callout-note collapse="true"}
## Solution

```{r ex3-build}
library(GenomeInfoDb)
ucsc <- GRanges("chr1", IRanges(c(100, 500), c(300, 700)))
ens  <- GRanges("1",    IRanges(c(250, 650), c(400, 800)))
ucsc
ens
seqlevelsStyle(ucsc)
seqlevelsStyle(ens)
```

By coordinate, `ucsc` range 1 (100–300) overlaps `ens` range 1 (250–400), and
`ucsc` range 2 (500–700) overlaps `ens` range 2 (650–800). But the two objects use
different naming conventions: `ucsc` is **UCSC** style (`chr1`) and `ens` is
**Ensembl** style (`1`).
:::

**Count the overlaps. What do you get, and is it right?**

::: {.callout-note collapse="true"}
## Solution

```{r ex3-bad}
sum(ucsc %over% ens)
countOverlaps(ucsc, ens)
```

Zero — even though we built the ranges to overlap. Nothing errors; the names
`chr1` and `1` simply never match, so every overlap test fails silently. This is
the trap: an overlap count of 0 that *should* be positive almost always means a
`seqlevelsStyle` mismatch, not a biological result.
:::

**Fix it and re-count.** Put both objects in the same naming style, then count
again.

::: {.callout-note collapse="true"}
## Solution

```{r ex3-fix}
seqlevelsStyle(ens) <- "UCSC"   # rewrite "1" -> "chr1" in place
ens                             # same ranges, new sequence names
sum(ucsc %over% ens)
countOverlaps(ucsc, ens)
```

Assigning a new `seqlevelsStyle` rewrites every sequence name without touching the
coordinates. Once both objects are UCSC style the names match, and the overlap
count is the 2 we expected all along.
:::

::: {.callout-warning}
## Check `seqlevelsStyle` before every cross-object operation

Whenever two `GRanges` come from different sources — one from UCSC, one from
Ensembl, one from a peak caller, one from an annotation package — confirm they
share a `seqlevelsStyle` before you compare them. The mismatch produces no error,
just wrong (usually zero) answers. `seqlevelsStyle(x) <- "UCSC"` (or `"Ensembl"`)
harmonizes them.
:::

## Exercise 4: real CTCF peaks meet a gene model

Now we leave the hand-built toys behind. In the ChIP-seq chapter you imported a
real set of ENCODE CTCF peaks; here you'll load the same file and ask how those
peaks relate to human genes. CTCF often binds near gene boundaries, so promoters
are a natural place to look.

The peak file is the one the book keeps in its repository (ENCODE accession
`ENCFF960ZGP`), read straight from its URL. The gene model is the hg38 UCSC
knownGene `TxDb`, which is already declared as a dependency of the book.

```{r ex4-load}
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

bed_url <- "https://raw.githubusercontent.com/seandavi/RBiocBook/main/data/ctcf_peaks_ENCFF960ZGP.bed.gz"
ctcf <- import(bed_url, format = "narrowPeak")

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```

**Sanity-check the naming styles before you compare anything.** This is the lesson
of Exercise 3, applied for real: confirm the peaks and the `TxDb` use the same
`seqlevelsStyle`.

::: {.callout-note collapse="true"}
## Solution

```{r ex4-style}
seqlevelsStyle(ctcf)
seqlevelsStyle(genes(txdb))
```

Both are **UCSC** style (`chr1`, `chr2`, …) — the peak file came from ENCODE in
UCSC naming and the `TxDb` was built on the UCSC genome — so overlaps between them
are safe. If either had come back `"Ensembl"`, you would harmonize first.
:::

**Build a clean set of promoter regions.** Pull the transcripts from the `TxDb`,
take a 2 kb promoter window around each transcription start site with
`promoters()`, then collapse the overlapping windows into a non-redundant set with
`reduce()`. How many promoter regions remain?

::: {.callout-note collapse="true"}
## Solution

```{r ex4-promoters}
tx    <- transcripts(txdb)
proms <- promoters(tx, upstream = 2000, downstream = 200)
prom_regions <- reduce(proms)
length(proms)          # one window per transcript (many overlap)
length(prom_regions)   # non-redundant promoter regions after reduce()
```

`promoters()` builds a strand-aware window around each transcript's start. Because
many transcripts share a start neighbourhood, those windows overlap heavily;
`reduce()` merges each overlapping cluster into one region so nothing is
double-counted downstream.
:::

**How many CTCF peaks fall in a promoter region?** Use `%over%` for the count, and
`subsetByOverlaps()` to pull out the peaks themselves.

::: {.callout-note collapse="true"}
## Solution

```{r ex4-overlap}
sum(ctcf %over% prom_regions)
prom_peaks <- subsetByOverlaps(ctcf, prom_regions)
length(prom_peaks)
```

`%over%` returns a logical vector — `TRUE` for each peak that hits at least one
promoter region — so `sum()` counts them. `subsetByOverlaps()` returns those same
peaks as a `GRanges`, ready for further work.
:::

**Is that overlap more than chance?** If CTCF peaks were scattered over the genome
*independently* of promoters, how many would you expect to land in a promoter by
chance? Compare that expectation to what you observed.

::: {.callout-note collapse="true"}
## Solution

```{r ex4-chance}
# Fraction of the genome covered by promoter regions. Some contigs have unknown
# (NA) lengths, so drop those when summing the genome size.
genome_bp  <- sum(as.numeric(seqlengths(txdb)), na.rm = TRUE)
prom_frac  <- sum(as.numeric(width(prom_regions))) / genome_bp

expected <- prom_frac * length(ctcf)   # peaks expected in promoters by chance
observed <- sum(ctcf %over% prom_regions)
c(expected = expected, observed = observed, ratio = observed / expected)
```

Promoter regions are a small fraction of the genome, so the chance expectation is
modest — yet the observed count is several-fold larger. That gap is the biology:
CTCF binding is **not** randomly placed with respect to promoters. The overlap is
meaningful precisely because it dwarfs the null expectation.
:::

## Exercise 5: which gene is each peak closest to?

Counting overlaps answers "is this peak *in* a feature?" But many peaks fall
*between* genes, and there the useful question is "which gene is this peak nearest,
and how far away?" That's a job for `nearest()` and `distanceToNearest()`.

**Annotate each peak with its nearest gene.** Get the per-gene ranges from the
`TxDb`, then for every CTCF peak find the index of the nearest gene with
`nearest()`. Attach the matched gene id to the peaks as a new metadata column.

::: {.callout-note collapse="true"}
## Solution

```{r ex5-nearest}
gn <- genes(txdb)
idx <- nearest(ctcf, gn)            # index into gn for each peak (NA if none)
# Keep only peaks that found a nearest gene (e.g. drop peaks on contigs gn lacks).
ok  <- !is.na(idx)
ctcf_ann <- ctcf[ok]
ctcf_ann$nearest_gene <- names(gn)[idx[ok]]
head(ctcf_ann$nearest_gene)
```

`nearest()` returns, for each peak, the index of the single closest gene —
overlapping, upstream, or downstream, whichever is nearest. Indexing `names(gn)`
with that vector turns the indices into gene ids, which we store on the peaks.
:::

**How far is each peak from its nearest gene?** Use `distanceToNearest()` and
summarize the distances. How many peaks sit *inside* a gene (distance 0)?

::: {.callout-note collapse="true"}
## Solution

```{r ex5-distance}
d2n  <- distanceToNearest(ctcf, gn)
dist <- mcols(d2n)$distance
summary(dist)
sum(dist == 0)            # peaks overlapping a gene
mean(dist == 0)           # ...as a fraction of all matched peaks
```

`distanceToNearest()` returns a `Hits` object with a `distance` column: 0 means the
peak overlaps its nearest gene, and any positive number is the base-pair gap to the
closest gene edge. The summary shows the spread; the count and fraction of zeros
tell you how much CTCF binding falls within genes versus in the intergenic space
between them.
:::

## Summary

You've now exercised the whole range-analysis toolkit, from hand-built objects to
real ENCODE data. You can:

- **Build a `GRanges`** from scratch and pull it apart with the coordinate
  accessors and `mcols()`, and subset it like a vector.
- **Reshape** a set with `flank()`, `resize()`, and `reduce()`, and **summarize**
  it with `coverage()` — reading strand-awareness and overlap depth correctly.
- **Recognize and fix** the silent zero-overlap bug a `chr1`-vs-`1`
  `seqlevelsStyle` mismatch causes.
- **Relate two real range sets** with `%over%`, `subsetByOverlaps()`,
  `nearest()`, and `distanceToNearest()`, and judge an overlap against a
  chance expectation.

For the underlying concepts — how `GRanges` objects are built, the full catalog of
set operations and overlap methods, and the coordinate-system details — return to
the *Genomic ranges and features* chapter. For the real-data import workflow with
`rtracklayer`, see *Genomic ranges with real ChIP-seq peaks*. This chapter was the
place to put those tools to work.
</content>
</invoke>
