---
title: "Reading and writing data files"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

Almost no analysis starts inside R. Your data lives somewhere else first — an
expression matrix exported from an instrument, a sample sheet a collaborator
typed into Excel, a public dataset posted on a government website. Before you can
do anything useful, you have to get that tabular data *into* R, and once you've
cleaned or summarized it, you'll often need to get it back *out* to share with
someone who does not use R.

This chapter is about that round trip. You'll write a small table of genes and
their expression values to disk, read it back, pull a real dataset straight from
a URL, and do the same with Excel files. None of it is hard once you know which
function to reach for — and knowing that is most of the battle.

## What you'll learn

- Write a data frame to a CSV file with `write_csv()` and read it back with
  `read_csv()`.
- Read a CSV directly from a URL, without downloading it by hand.
- Read and write Excel files with `read_excel()` and `write_xlsx()`.
- Recognize when other formats (Google Sheets, JSON, databases) need a different
  package.

This chapter assumes you've already organized your analysis into an RStudio
Project (see @sec-rstudio-projects), so that the relative paths below resolve from
your project folder.

## CSV files

A CSV ("comma-separated values") file is the lingua franca of tabular data:
plain text, one row per line, columns separated by commas. Nearly every tool —
Excel, Python, databases, sequencing pipelines — can read and write CSV, which is
exactly why it's so useful for moving data around.

Base R reads and writes CSV with `read.csv()` and `write.csv()`. We'll use the
`readr` package instead, which provides `read_csv()` and `write_csv()`. They are
faster, give cleaner output, and guess column types more sensibly. Install it
once:

```{r eval=FALSE}
install.packages("readr")
```

Then load it at the start of your session:

```{r}
library(readr)
```

### Writing a CSV file

We need something to write, so let's build a small biological data frame: a
handful of genes, each with a symbol and a measured expression value.

```{r}
expr <- data.frame(
  gene_id    = c("YAL001C", "YAL002W", "YAL003W", "YAL005C", "YAL007C"),
  symbol     = c("TFC3", "VPS8", "EFB1", "SSA1", "ERP2"),
  expression = c(8.2, 5.1, 11.7, 9.4, 6.8)
)
expr
```

Each row is one gene; the columns mix text (`gene_id`, `symbol`) and numbers
(`expression`), which is the everyday shape of real expression data. Now write it
to disk with `write_csv()`:

```{r}
write_csv(expr, "data.csv")
```

That call created a file named `data.csv` in your working directory. You can
confirm it landed where you expect:

```{r}
# Where am I working right now?
getwd()
# Did the file get created?
dir(pattern = "data.csv")
```

If you wanted the file somewhere else, you'd pass a fuller path (for example
`"outputs/tables/data.csv"`). Inside an RStudio Project, that path is relative to
the project folder.

### Reading a CSV file

Reading the file back is the mirror image — `read_csv()`:

```{r}
expr_in <- read_csv("data.csv")
```

Notice the message `readr` prints: it tells you how it guessed each column's
type (text vs. number). Let's look at what came back:

```{r}
expr_in
```

The genes and their expression values round-tripped intact. The
`gene_id` and `symbol` columns came back as character, `expression` as a number
(`dbl`) — `readr` inferred all of that automatically.

### Reading a CSV straight from the web

`read_csv()` is not limited to files on your own disk. Hand it a URL and it will
fetch and parse the file in one step. This matters because a great deal of useful
data lives on public websites, and reading it directly — rather than downloading
by hand — keeps your analysis reproducible: anyone who runs your script pulls the
same data from the same place.

As an example, the `palmerpenguins` project publishes its penguin body
measurements as a public CSV on GitHub:

```{r}
penguins <- read_csv("https://raw.githubusercontent.com/allisonhorst/palmerpenguins/main/inst/extdata/penguins.csv")
penguins
```

The dataset — bill and flipper measurements for three penguin species — is
documented [here](https://allisonhorst.github.io/palmerpenguins/). The point is
not the penguins themselves: it's that `read_csv()` detected the file's structure
over the network and handed you a data frame, exactly as it did for your local
file. Any public CSV, biological or otherwise, comes in the same way.

## Excel files

Excel spreadsheets (`.xlsx`) are everywhere in biology — sample sheets, plate
layouts, hand-curated gene lists. Unlike CSV, an Excel file can hold multiple
sheets, formulas, and formatting, so it needs a dedicated package. Reading is
handled by `readxl`; writing by `writexl`.

### Reading an Excel file

Install and load `readxl`:

```{r eval=FALSE}
install.packages("readxl")
```

```{r}
library(readxl)
```

We don't have an Excel file lying around, and — unlike `read_csv()` — `readxl`
can't read straight from a URL. So we first *download* the file, then read it from
disk. Here's a small public example spreadsheet:

```{r}
download.file(
  "https://www.w3resource.com/python-exercises/pandas/excel/SaleData.xlsx",
  "SaleData.xlsx",
  mode = "wb"
)
```

The `mode = "wb"` ("write binary") matters for non-text files like `.xlsx`;
without it, the download can be corrupted on some platforms. Confirm it arrived:

```{r}
dir(pattern = "SaleData.xlsx")
```

Now read it with `read_excel()`:

```{r}
sales <- read_excel("SaleData.xlsx")
sales
```

`read_excel()` has options for choosing a specific sheet (`sheet =`), a cell range
(`range =`), or how column types are guessed — handy when a spreadsheet has more
than one tab or some messy header rows.

### Writing an Excel file

To go the other direction, use `write_xlsx()` from `writexl`. Install and load it:

```{r eval=FALSE}
install.packages("writexl")
```

```{r}
library(writexl)
```

Then write our gene expression data frame to an `.xlsx` file:

```{r}
write_xlsx(expr, "data.xlsx")
```

And confirm it was created:

```{r}
getwd()
dir(pattern = "data.xlsx")
```

If you open `data.xlsx` in Excel, you'll see the same five genes and their
expression values you started with.

## Additional formats

CSV and Excel cover most of what you'll meet, but R reads and writes many other
formats too. When you hit one, reach for the matching package:

- **Google Sheets** — `googlesheets4` reads and writes sheets directly from your
  Google account.
- **JSON** — `jsonlite` converts between R objects and JSON, common for data
  pulled from web APIs.
- **Databases** — `DBI` together with a backend like `RSQLite` lets you query a
  database and pull results straight into a data frame.

The pattern is always the same: find the package for the format, then look for its
`read_*` and `write_*` functions.

## Exercises

Use the `expr` data frame from this chapter (five yeast genes with a `symbol` and
an `expression` value) unless a problem says otherwise.

1. **Round-trip a CSV.** Write `expr` to a file called `genes.csv`, then read it
   into a new object called `genes_in`. Confirm the file exists with `dir()`.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   write_csv(expr, "genes.csv")
   genes_in <- read_csv("genes.csv")
   dir(pattern = "genes.csv")
   ```
   `write_csv()` saves it; `read_csv()` brings it back. The file name is just a
   string, so you can call it anything you like.
   :::

2. **Read a column back.** From `genes_in`, print just the `symbol` column.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   genes_in$symbol
   ```
   The `$` operator pulls one column out of a data frame as a vector — the same
   trick you used in the data frames chapter.
   :::

3. **Save to Excel.** Write `expr` to `genes.xlsx` and confirm it was created.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   write_xlsx(expr, "genes.xlsx")
   dir(pattern = "genes.xlsx")
   ```
   `write_xlsx()` from `writexl` writes a `.xlsx` file with no need for Excel to
   be installed.
   :::

4. **Why a relative path?** A colleague sends you a script whose first line is
   `read.csv("/Users/dana/project/data/counts.csv")`. It fails on your machine.
   Why, and what would you change it to?

   ::: {.callout-note collapse="true"}
   ## Solution
   That absolute path points to a folder on Dana's computer that doesn't exist on
   yours. Inside an RStudio Project, a relative path like
   `read.csv("data/counts.csv")` works for anyone who opens the project, because
   the working directory is the project folder.
   :::

## Summary

You can now move tabular data in and out of R in both directions:

- **Organize first.** As covered in @sec-rstudio-projects, an RStudio Project
  gives you a stable working directory so relative paths (`data/file.csv`) stay
  portable; absolute paths do not.
- **CSV** — `write_csv()` saves a data frame, `read_csv()` reads it back, and
  `read_csv()` will even read straight from a URL for reproducible access to
  public data.
- **Excel** — `read_excel()` reads `.xlsx` (download first; it can't read a URL),
  and `write_xlsx()` writes one.
- **Other formats** — Google Sheets, JSON, and databases each have a package
  (`googlesheets4`, `jsonlite`, `DBI`/`RSQLite`) that follows the same
  `read_*`/`write_*` pattern.

With data flowing freely in and out, you're ready to spend your time on the part
that matters: the analysis.
