---
title: "Organizing, saving, and loading your work"
author:
  - "Sean Davis"
  - "Lori Kern"
---

Every analysis you do leaves a trail: scripts, data files, figures, and the
objects you build up in your R session along the way. Where you put those files,
and whether you can get your work *back* the next time you sit down, is the
difference between an analysis you can rerun a year from now and one you can
barely reconstruct tomorrow. This chapter is about keeping that trail tidy.

Think of it the way you'd think of a **lab notebook**. A good notebook isn't just
a place to scribble numbers — it's a record organized so that someone else, or
*you* six months from now, can open it and follow exactly what was done, in what
order, with which samples. The same discipline pays off at the keyboard. Loose
files scattered across your Desktop, a Downloads folder, and three different
project directories are the digital equivalent of notes on the backs of
envelopes: fine in the moment, useless when you need to reconstruct what you did.

The reader you should organize for is almost always **future you**. The
collaborator who can't find the raw data, the reviewer who asks you to rerun the
analysis a year later, the version of yourself coming back to a half-finished
project after a conference — all of them are best served by work that is
*self-contained* and *portable*. If a colleague can clone your project folder,
open one file, and run your scripts with no edits, then sharing your analysis
costs you nothing. If instead your code is laced with paths like
`/Users/yourname/Desktop/...`, every hand-off becomes a debugging session. Good
organization is what turns a private analysis into something you can share — and
trust — without ceremony.

There's one place the notebook analogy breaks down, and it's worth naming. A
paper notebook is a *permanent, append-only* record: you never erase a page, and
its value is precisely that it captures what you did, mistakes and all. The
objects in an R session are the opposite — *transient* working state that
vanishes when you quit, and saving them is a deliberate act. That difference is
exactly why this chapter has two halves. First we make your *files* tidy and
portable with **RStudio Projects** — the durable, notebook-like record of an
analysis. Then we look at how to **save and load** the transient parts — your
workspace, individual objects, and your command history — so you can pick up
exactly where you left off.

## What you'll learn

- Organize an analysis in an RStudio Project so your file paths stay portable.
- Lay out a project folder that separates raw data, scripts, and outputs.
- Use relative paths (and the `here` package) instead of fragile absolute paths.
- Save and restore your whole workspace with `save.image()` and `load()`.
- Save and reload *selected* objects with `save()` and `load()`.
- Save and replay your command history with `savehistory()` and `loadhistory()`.

## Organizing your work with RStudio Projects {#sec-rstudio-projects}

Before you read or write a single file, it helps to decide *where* those files
live. An RStudio Project is a folder that holds everything for one analysis —
scripts, data, and outputs — together in one place. When you create a project,
RStudio drops a `.Rproj` file into the folder, and that file anchors your
workspace.

Working in a project buys you a few things:

- **A consistent working directory.** Open the project and your working
  directory is automatically the project folder. No more guessing where R thinks
  it is.
- **Everything in one place.** Scripts, data, and outputs travel together.
- **Reproducibility.** A collaborator can open your project and run your code
  without rewriting any paths.
- **Version control.** Projects work cleanly with Git and GitHub.

### Creating a project

You can make a new project from **File > New Project...**, or from the project
dropdown in the top-right corner of RStudio. You'll be offered three choices:

- **New Directory** — start a fresh project folder.
- **Existing Directory** — turn a folder you already have into a project.
- **Version Control** — clone a repository from GitHub.

### A sensible project layout

A good project has a predictable shape — one that *you* define and then stick to.
A common starting point:

```
my_analysis_project/
├── my_analysis_project.Rproj
├── data/
│   ├── raw/
│   └── processed/
├── scripts/
├── notebooks/
├── outputs/
│   ├── figures/
│   └── tables/
├── README.md
└── .gitignore
```

This keeps raw data separate from processed data, scripts in one place, and
outputs where you can find them later.

### Why this fixes file paths

The biggest practical payoff is path management. When you open a project,
RStudio sets the working directory to the project folder, so you can refer to
files *relative* to that folder instead of spelling out the full path from the
root of your hard drive.

```{r eval=FALSE}
# An absolute path is tied to one machine and one user:
expr <- read.csv("/Users/username/Documents/my_analysis/data/dataset.csv")

# A relative path works for anyone who opens the project:
expr <- read.csv("data/dataset.csv")
```

::: {.callout-warning}
## Absolute paths break reproducibility

An absolute path like `/Users/username/Documents/...` only exists on *your*
computer. Email that script to a colleague and the very first line fails for
them. Relative paths (`data/dataset.csv`), anchored by a project, are the habit
that keeps your code portable.
:::

::: {.callout-tip}
## The `here` package for robust paths

In larger projects — especially ones with notebooks tucked into subfolders — the
`here` package makes paths bulletproof. `here()` always resolves from the project
root (the folder with the `.Rproj` file), no matter which subfolder your script
sits in:

```{r eval=FALSE}
install.packages("here")
library(here)

expr <- read.csv(here("data", "dataset.csv"))
```
:::

When you share a project folder (or push it to GitHub), a collaborator can clone
it, open the `.Rproj` file, and run your scripts with no path edits at all. That
seamless hand-off is what makes an analysis credible and repeatable.

## Saving and loading your workspace

A project keeps your *files* organized. But what about the objects you build up
during a session — the data frame you cleaned, the model you fit, the vector of
gene names you assembled by hand? Those live in R's memory, and they vanish when
you quit unless you save them.

The whole collection of objects in your current session is called your
**workspace** (or the *global environment*). When you quit R with `q()` and
answer "yes" to *Save workspace image?*, R writes your entire workspace to a file
named `.RData` in the working directory. That is exactly equivalent to running:

```{r, eval=FALSE}
save.image()
```

The next time you start R *in the same directory*, it finds that `.RData` file
and reloads every object automatically — you're back where you left off.

You can also save the workspace to a file you name yourself:

```{r, eval=FALSE}
save.image(file = "analysis_checkpoint.RData")
```

A named file behaves differently from the default `.RData`: R will **not** load
it automatically on startup. You reload it by hand whenever you want it back:

```{r, eval=FALSE}
load("analysis_checkpoint.RData")
```

::: {.callout-tip}
## Prefer named files, and a fresh session

Letting R silently restore `.RData` every time feels convenient, but it quietly
brings along *every* object from your last session — including mistakes. Many
experienced users turn that behavior off and instead save named checkpoints with
`save.image("descriptive_name.RData")`, so each saved state is deliberate and
self-documenting.
:::

## Saving and loading selected objects

Saving the *entire* workspace is often more than you want. Usually only a couple
of objects are worth keeping — the cleaned dataset, the final result — while the
rest is scratch work. The `save()` function lets you pick exactly which objects
to write, and `load()` brings them back.

```{r, eval=FALSE}
ages <- 1:4
months <- c("may", "june", "july", "august")
flags <- c(TRUE, FALSE, TRUE)

# Save just two of the three objects:
save(months, ages, file = "subset.RData")
```

In a later session, `load()` drops `months` and `ages` straight back into your
workspace — `flags` is not in the file, so it does not come along:

```{r, eval=FALSE}
load("subset.RData")
```

::: {.callout-note}
## `save()` versus `saveRDS()`

`save()` stores objects *together with their names*, so `load()` recreates them
under those same names. A close cousin, `saveRDS()`, stores a *single* object
with no name attached, and you assign it yourself on the way back in:
`my_data <- readRDS("file.rds")`. Reach for `saveRDS()` when you want one object
and full control over what it's called.
:::

## Saving and loading your command history

R also remembers the *commands* you type, not just the objects they create. When
you save your session on quit, your command history is saved too. You can manage
it by hand with `savehistory()` and `loadhistory()`:

```{r, eval=FALSE}
savehistory(file = "analysis.Rhistory")   # write the commands you've run
loadhistory(file = "analysis.Rhistory")   # replay them into a new session
```

Saving the history is handy for turning an exploratory session into a clean
script later: you can open the `.Rhistory` file, keep the lines that mattered,
and drop the dead ends.

## Exercises

1. **Spot the fragile path.** A collaborator sends you a script that begins with
   `expr <- read.csv("C:/Users/jordan/Desktop/data/counts.csv")`. It runs fine on
   their laptop but errors on yours. Why, and how would you rewrite the line so it
   works for both of you?

   ::: {.callout-note collapse="true"}
   ## Solution
   That path exists only on Jordan's machine, so R can't find the file on yours.
   Put the data inside an RStudio Project (say, in a `data/` folder) and use a
   relative path instead:

   ```{r eval=FALSE}
   expr <- read.csv("data/counts.csv")
   ```

   Now anyone who opens the project gets the same working directory, and the path
   resolves the same way for everyone.
   :::

2. **Save just what matters.** You have three objects in your session —
   `counts`, `samples`, and a throwaway `tmp`. Write the one line that saves only
   `counts` and `samples` to a file called `results.RData`, then the line that
   loads them back in a future session.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r eval=FALSE}
   save(counts, samples, file = "results.RData")
   # ... later, in a fresh session ...
   load("results.RData")
   ```
   `tmp` is never named in the `save()` call, so it isn't written — and `load()`
   restores `counts` and `samples` under their original names.
   :::

3. **One object, your name.** You want to hand a single cleaned data frame,
   `brfss_clean`, to a colleague who will read it under a name of *their* choosing.
   Which function fits better, `save()` or `saveRDS()`, and why?

   ::: {.callout-note collapse="true"}
   ## Solution
   `saveRDS()`. It stores the object without a name, so your colleague assigns it
   themselves on the way in:

   ```{r eval=FALSE}
   saveRDS(brfss_clean, file = "brfss_clean.rds")
   # colleague's session:
   my_data <- readRDS("brfss_clean.rds")
   ```
   `save()` would have forced the object back into their workspace as
   `brfss_clean`, whether they wanted that name or not.
   :::

## Summary

You now have the habits that keep an analysis reproducible from one session to
the next:

- **Organize** each analysis as an RStudio Project so your working directory is
  predictable and your paths are *relative*, not tied to one machine.
- **Checkpoint** a whole session with `save.image()` / `load()`, preferring named
  files over the silent `.RData` restore.
- **Save selectively** with `save()` (objects plus their names) or `saveRDS()`
  (one object, you name it on the way back).
- **Keep your history** with `savehistory()` / `loadhistory()` to turn a messy
  session into a tidy script.

With your work organized and safely saved, let's explore how to read and write
the data files that will live within these project structures.
