---
title: "The t-statistic and t-distribution"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

Suppose you measure a gene's expression in five treated samples and five
controls, and the treated mean comes out higher. Is that a real effect, or just
the kind of gap you'd expect by chance from five noisy measurements? With only a
handful of samples you can't lean on the normal distribution as-is, because you
had to *estimate* the variability from the same tiny sample you're testing. The
**t-distribution** is the honest accounting for that extra uncertainty — and in
this chapter we'll *build* it by simulation so you can see exactly where it comes
from, then use R's `t.test()` to put it to work.

The distribution that the t-statistic follows was described in a famous 1908
paper [@student_probable_1908] by "Student" — a pseudonym for
[William Sealy Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), who
worked at the Guinness brewery and published anonymously because his employer
treated statistical methods as a trade secret.

## What you'll learn

- Explain why estimating the standard deviation from a small sample forces us off
  the normal distribution and onto the t-distribution.
- Read a Z-score and a t-score and say what each one measures.
- Use simulation to show that t-based p-values are calibrated (uniform under the
  null hypothesis) when the normal-based ones are not.
- Run one-sample, two-sample, and formula-based tests with `t.test()`.
- Estimate statistical power with `power.t.test()` and by simulation.

## Background

A **t-test** is a [statistical hypothesis test](https://en.wikipedia.org/wiki/Statistical_hypothesis_testing)
you reach for when you want to compare means and you only have a sample to work
from. It comes in three everyday flavors:

- **One-sample** — compare a sample mean to a single hypothesized or target value.
- **Two-sample** — compare the means of two groups.
- **Paired** — compare two measurements on the same units (e.g., before-and-after
  on the same patients).

A t-test combines three ingredients — the **t-statistic**, the
**t-distribution**, and the **degrees of freedom** — to decide whether a
difference is bigger than chance would comfortably produce. (To compare three or
more means at once you'd reach instead for an analysis of variance, or ANOVA.)

We'll get to `t.test()`, but the point of this chapter is to first see *why* the
t-distribution exists. That story starts with something you may already have met:
the Z-score.

## The Z-score and probability

Before we talk about t-scores, let's review the Z-score, its relationship to the
normal distribution, and how it turns into a probability.

The Z-score is defined as:

$$Z = \frac{x - \mu}{\sigma}$$ {#eq-zscore}

where $\mu$ is the population mean that $x$ is drawn from and $\sigma$ is the
population standard deviation — **taken as known**, not estimated from the data.
In words: a Z-score says how many standard deviations a value sits from the mean.

The probability of observing a Z-score of $z$ or smaller can be calculated with
`pnorm(z, mu, sigma)`.

For example, suppose our "population" is known and truly has a mean of 0 and a
standard deviation of 1 (the *standard normal*). For any observation drawn from
that population, we can assign a probability of seeing a value that extreme by
random chance — *under the assumption that the null hypothesis is **TRUE***.

```{r}
zscore <- seq(-5, 5, 1)
```

For each value of `zscore`, let's calculate the probability and collect the
results in a `data.frame`. We'll name it `pdat` (for "probability data") so it
doesn't collide with the `df` we use later for degrees of freedom.

```{r}
pdat <- data.frame(
    zscore = zscore,
    pval   = pnorm(zscore, 0, 1)
)
pdat
```

Why is the p-value for something 5 standard deviations *above* the mean
(`zscore = 5`) nearly 1 here? Because `pnorm` reports the *lower* tail by default
— the probability of a value *at or below* `z`. A value 5 standard deviations
above the mean has almost the entire distribution below it, so that probability
is nearly 1.

Let's plot probability against Z-score:

```{r}
plot(pdat$zscore, pdat$pval, type = 'b')
```

This curve is the *cumulative distribution function* (CDF). Read it like a lookup
table: pick a Z-score on the x-axis and read off the probability of seeing a
value that small. Because we built our example to follow the standard normal, this
is also the CDF of the standard normal distribution.

::: {.callout-tip}
## What a p-value is (and isn't)

A p-value is the probability of a result *at least this extreme* **if the null
hypothesis were true**. It is *not* the probability that the null hypothesis is
true, and a small p-value is not a measure of effect size. For an accessible,
authoritative take on the common misreadings, see the American Statistical
Association's statement on p-values [@doi:10.1080/00031305.2016.1154108].
:::

### Small diversion: a two-sided pnorm

`pnorm` returns a *one-sided* probability using the lower tail by default. Often
we don't care which direction a value deviates — only that it's far from the mean
— so we want a *two-sided* p-value. We can build one in three steps:

1. Take the absolute value of `x`, so direction no longer matters.
2. Call `pnorm` with `lower.tail = FALSE`, so larger `|x|` gives smaller
   probabilities.
3. Multiply by 2 to count *both* tails.

```{r}
twosidedpnorm <- function(x, mu = 0, sd = 1) {
    2 * pnorm(abs(x), mu, sd, lower.tail = FALSE)
}
```

Now we can ask how surprising it is to land 2 or 3 standard deviations from the
mean in either direction:

```{r}
twosidedpnorm(2)
twosidedpnorm(3)
```

## The t-distribution

Everything above leaned on one big assumption: that we *know* the standard
deviation $\sigma$. Recall the Z-score:

$$Z = \frac{x - \mu}{\sigma}$$

and the *population* standard deviation:

$$\sigma = \sqrt{\frac{1}{N}\sum_{i=1}^{N}(x_i - \mu)^2}$$ {#eq-pop-sd}

When we genuinely know $\sigma$, the Z-score is the right tool. But usually we
only have a *sample*, not the whole population. Then we swap the Z-score for the
**t-score**:

$$t = \frac{x - \bar{x}}{s}$$ {#eq-tscore}

This looks almost identical to the Z-score, with two changes: we use the *sample*
mean $\bar{x}$ in place of $\mu$, and we *estimate* the standard deviation as $s$
from the data:

$$s = \sqrt{\frac{1}{N-1}\sum_{i=1}^{N}(x_i - \bar{x})^2}$$ {#eq-sample-sd}

In words: the sample standard deviation $s$ is the typical distance of a data
point from the sample mean. The only change from the population formula is the
divisor — $N-1$ instead of $N$ — which corrects for the fact that we used the
data twice: once to estimate the mean, and again to measure the spread around it.

::: {.callout-important}
## The whole point of the t-distribution

Because we *estimate* the standard deviation from a small sample, our test
statistic is more uncertain than a Z-score. The t-distribution captures that by
having **fatter tails** for small samples — so it takes a more extreme result to
reach significance. As the sample grows, the estimate sharpens and the
t-distribution slides back toward the normal.
:::

We can *see* those fatter tails. Let's plot the t-distribution densities for 3, 6,
10, and 20 degrees of freedom (smaller degrees of freedom means a smaller sample)
alongside the normal:

```{r fig-t-vs-z, message=FALSE, fig.cap="t-distributions for various degrees of freedom. The tails are fatter for smaller degrees of freedom — the visible cost of estimating the standard deviation from the data."}
library(dplyr)
library(ggplot2)
t_values <- seq(-6, 6, 0.01)
tdens <- data.frame(
    value  = t_values,
    t_3    = dt(t_values, 3),
    t_6    = dt(t_values, 6),
    t_10   = dt(t_values, 10),
    t_20   = dt(t_values, 20),
    Normal = dnorm(t_values)
) |>
    tidyr::pivot_longer(-value, names_to = "Distribution", values_to = "density")
ggplot(tdens, aes(x = value, y = density, color = Distribution)) +
    geom_line()
```

The `dt` and `dnorm` functions give the *density* (the height of the curve) at
each point. Notice how the `t_3` curve sits lowest in the middle and highest in
the tails — that extra tail weight is the uncertainty from a tiny sample, drawn to
scale.

We can also look at the cumulative version of each curve:

```{r}
tcdf <- tdens |>
    group_by(Distribution) |>
    arrange(value) |>
    mutate(cdf = cumsum(density))
ggplot(tcdf, aes(x = value, y = cdf, color = Distribution)) +
    geom_line()
```

### p-values based on Z vs t

When we have a *sample* and want to test how far its mean sits from a population
mean, we scale by the standard error (the standard deviation of the sample mean),
$\sigma/\sqrt{n}$:

$$z = \frac{x - \mu}{\sigma/\sqrt{n}}$$

Let's compare the p-value we'd get from the normal distribution against the one
from the t-distribution, for a single small sample.

```{r}
set.seed(5432)
samp <- rnorm(5, mean = 0.5)
z <- sqrt(length(samp)) * mean(samp)  # simplifying assumption: sigma = 1, mu = 0
```

The p-value if we *pretend* we know the standard deviation:

```{r}
pnorm(z, lower.tail = FALSE)
```

In reality we don't know it, so we estimate it from the data with `sd()` and use
the t-distribution (with `length(samp) - 1` degrees of freedom):

```{r}
ts <- sqrt(length(samp)) * mean(samp) / sd(samp)
pnorm(ts, lower.tail = FALSE)                      # wrong: treats the estimate as known
pt(ts, df = length(samp) - 1, lower.tail = FALSE)  # right: accounts for the estimate
```

The two numbers differ, and that difference is the whole story: using the normal
distribution on an *estimated* standard deviation quietly understates our
uncertainty. The next section makes that mistake visible at scale.

### Experiment: does the choice of distribution actually matter?

Here's a claim worth testing for yourself rather than taking on faith: *if you
estimate the standard deviation but compute p-values from the normal
distribution, your p-values will be wrong.* Let's find out how wrong.

There's a tidy way to check. If a test is working correctly, then **under the
null hypothesis its p-values should be uniformly distributed between 0 and 1** —
every value equally likely. So we'll simulate many samples where the null is
*true*, compute p-values three ways, and look at the histograms. A flat histogram
means the test is honest; a lopsided one means it isn't.

The plan:

1. Draw many samples of size `n` from the standard normal distribution.
2. Score each with the **Z**-statistic (pretending we know $\sigma$).
3. Score each with the **t**-statistic (estimating $\sigma$ from the data), then
   compute p-values *two* ways — from the normal, and from the t-distribution.

First, a function that draws a sample and returns its Z-score:

```{r}
zf <- function(n) {
    samp <- rnorm(n)
    sqrt(length(samp)) * mean(samp) / 1  # simplifying assumption: sigma = 1, mu = 0
}
```

Give it a try:

```{r}
zf(5)
```

Run 10,000 replicates and look at the p-value histogram. Here we *do* know the
population standard deviation (we're sampling from the standard normal), so the
normal distribution is the correct choice — and the histogram should be flat:

```{r}
z10k <- replicate(10000, zf(5))
hist(pnorm(z10k))
```

Flat, as promised. Now the t-score function — same idea, but we *estimate* the
standard deviation with `sd()` instead of assuming it:

```{r}
tf <- function(n) {
    samp <- rnorm(n)
    # use the sample standard deviation, since we "don't know" the population one
    sqrt(length(samp)) * mean(samp) / sd(samp)
}
```

If we feed those t-scores to the **normal** distribution — the common mistake —
the histogram is no longer flat:

```{r}
t10k <- replicate(10000, tf(5))
hist(pnorm(t10k))
```

The pile-ups near 0 and 1 mean this test rejects the null too often: it's
overconfident, exactly because it treats an *estimated* standard deviation as if
it were known. Now feed the same t-scores to the **t-distribution** (with
`n - 1 = 4` degrees of freedom):

```{r}
hist(pt(t10k, 4))
```

Flat again. The t-distribution restores honest, uniform p-values — *that* is the
payoff for all the math above.

We can also compare the two sets of scores directly with a Q-Q plot.

A **Q-Q plot** plots the quantiles of two distributions against each other. If
the two distributions are identical, the points fall on a straight line; where
they differ, the points peel away from it.

Here we compare our simulated z-scores against our simulated t-scores. Because the
t-scores carry the extra uncertainty from estimating the standard deviation, we
expect them to disagree with the z-scores in the tails.

```{r}
qqplot(z10k, t10k)
abline(0, 1)
```

The points bow away from the line at both ends — the t-distribution's fatter
tails, made visible. What would happen if you raised the sample size from 5 to
50? Try it: the bow should shrink as the t-distribution approaches the normal.

## Checkpoint: t vs normal

Before we start using `t.test()`, let's pin down the one idea this chapter has
been circling.

The t-distribution is a *family* of distributions indexed by the **degrees of
freedom**, which track the sample size. It has heavier tails than the normal for
small samples and approaches the normal as the sample grows. Those heavier tails
make it more conservative — it takes a more extreme result to reach significance —
which is the correct response to the uncertainty introduced by *estimating* the
standard deviation from the data. That correction matters most exactly when you
have the least data.

## t.test {#sec-ttest}

Now the practical part. R's `t.test()` packages everything above into one
function — you hand it data, it returns a t-statistic, degrees of freedom, a
p-value, and a confidence interval.

### One-sample

A one-sample test compares a sample mean to a single value (zero, by default).
We'll simulate a sample whose true mean is 1, so we *know* the answer the test
should be reaching for.

```{r}
x <- rnorm(20, 1)
# small sample: just the first 5 values
t.test(x[1:5])
```

We rigged this so the null hypothesis (mean = 0) is **false** — the true mean is
1. But with only 5 values the evidence is thin, so the p-value is modest. Add more
data and the effect comes through clearly:

```{r}
t.test(x[1:20])
```

### Two-sample

A two-sample test compares the means of two groups:

```{r}
x <- rnorm(10, 0.5)
y <- rnorm(10, -0.5)
t.test(x, y)
```

### From a data frame

Often your data and group labels live in two columns of a data frame rather than
in two separate vectors:

```{r}
tdat <- data.frame(
    value = c(x, y),
    group = as.factor(rep(c('g1', 'g2'), each = 10))
)
tdat
```

R lets you run the same test with **formula notation**:

```{r}
t.test(value ~ group, data = tdat)
```

Read `value ~ group` as "*`value` is a function of `group`*" — in practice, do a
t-test of the `value`s in `g1` against those in `g2`.

### Equivalence to a linear model

A two-sample t-test (with equal variances assumed) is really a linear model in
disguise. Run the test:

```{r}
t.test(value ~ group, data = tdat, var.equal = TRUE)
```

and compare it to fitting a line:

```{r}
res <- lm(value ~ group, data = tdat)
summary(res)
```

The `groupg2` coefficient is the difference in means, and its p-value matches the
t-test's. Same question, same answer, two front doors.

## Power calculations

The **power** of a test is the probability that it *correctly* rejects the null
hypothesis when the alternative is true — in other words, the probability of *not*
missing a real effect (not making a Type II error). Power rises with the
significance level (`alpha`), the sample size, and the effect size. It's what you
compute *before* an experiment to ask "how many samples do I need?"

`power.t.test()` ties these quantities together. From `help("power.t.test")`, the
main arguments are:

- `n` — sample size
- `delta` — effect size (difference in means)
- `sd` — standard deviation
- `sig.level` — significance level
- `power` — power

You supply any four and it solves for the fifth. For example, the power of a
one-sample test with `n = 5`, `sd = 1`, and an effect size of 1:

```{r}
power.t.test(n = 5, delta = 1, sd = 1, sig.level = 0.05)
```

That prints a full summary. To grab just the power value, use `$`:

```{r}
power.t.test(n = 5, delta = 1, sd = 1,
             sig.level = 0.05, type = 'one.sample')$power
```

::: {.callout-tip}
When a function returns a result that doesn't look "computable" — like the summary
from `power.t.test` — you can pull out a single piece with `$`. How do you know
what's available to extract? Use `names()` for the labels, or `str()` for the full
structure:

```{r}
names(power.t.test(n = 5, delta = 1, sd = 1,
             sig.level = 0.05, type = 'one.sample'))
# or
str(power.t.test(n = 5, delta = 1, sd = 1,
             sig.level = 0.05, type = 'one.sample'))
```
:::

Just as often you'll flip the question around: *how big a sample* do I need to hit
a target power? Solve for `n` by supplying `power` instead:

```{r}
power.t.test(delta = 1, sd = 1, sig.level = 0.05, power = 0.8, type = "one.sample")
```

`power.t.test` is convenient and fast. But as we saw with the t-distribution
itself, sometimes the distribution of a test statistic is **not** easily
calculated in closed form. In those cases we can estimate power by *simulation* —
run the experiment thousands of times and count how often it rejects the null:

```{r}
sim_t_test_pval <- function(n = 5, delta = 1, sd = 1, sig.level = 0.05) {
    x <- rnorm(n, delta, sd)
    t.test(x)$p.value <= sig.level
}
pow <- mean(replicate(1000, sim_t_test_pval()))
pow
```

Step by step: `sim_t_test_pval` simulates one sample of size `n` from a normal
distribution with mean `delta` and standard deviation `sd`, runs a one-sample
t-test, and returns `TRUE` when the p-value clears the significance threshold.
`replicate` repeats that 1000 times, and `mean` turns the resulting `TRUE`/`FALSE`
vector into the *proportion* of significant results — our estimate of power.

The two approaches should land in the same place:

```{r}
power.t.test(n = 5, delta = 1, sd = 1, sig.level = 0.05, type = 'one.sample')$power
mean(replicate(1000, sim_t_test_pval(n = 5, delta = 1, sd = 1, sig.level = 0.05)))
```

Close — and the simulation route keeps working in situations where no neat
formula exists.

## Exercises

::: {.callout-note}
Several exercises reuse objects (`samp`, `t10k`, `tdat`) built earlier in the
chapter. Run the chapter's chunks first, or recreate the objects in your session.
:::

1. **One-sided vs two-sided.** Use `twosidedpnorm()` to find the two-sided
   p-value for a Z-score of 1.96. Does the answer match the "95%" rule of thumb
   you may have heard?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   twosidedpnorm(1.96)
   ```
   It's almost exactly 0.05 — which is *why* 1.96 is the classic cutoff for a
   two-sided test at the 5% level.
   :::

2. **Degrees of freedom matter.** Compute the two-sided p-value for `t = 2.5`
   from the t-distribution with 4 degrees of freedom and again with 30. Which is
   larger, and does that fit the "fatter tails" story?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   2 * pt(2.5, df = 4,  lower.tail = FALSE)
   2 * pt(2.5, df = 30, lower.tail = FALSE)
   ```
   The 4-df p-value is larger. Fewer degrees of freedom means fatter tails, so the
   same t-statistic is *less* surprising — it takes more extreme evidence to reach
   significance with a small sample.
   :::

3. **A paired test.** Imagine the same 8 patients measured before and after a
   treatment. Simulate the pairs below and run a paired t-test with
   `t.test(after, before, paired = TRUE)`. How does the result compare to treating
   the two groups as independent (`paired = FALSE`)?

   ```{r}
   set.seed(11)
   before <- rnorm(8, 10, 2)
   after  <- before + rnorm(8, 1, 0.5)  # a real, consistent bump per patient
   ```

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   t.test(after, before, paired = TRUE)
   t.test(after, before, paired = FALSE)
   ```
   The paired test has a much smaller p-value. By matching each patient to
   themselves it removes the patient-to-patient variability, leaving only the
   change due to treatment — so it detects the effect that the unpaired test can
   barely see.
   :::

4. **Power and sample size.** Using `power.t.test`, find the sample size needed to
   detect an effect of `delta = 0.5` (with `sd = 1`) at 80% power for a one-sample
   test. Then halve the effect size to `0.25` and recompute. Roughly how does the
   required `n` change?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   power.t.test(delta = 0.5,  sd = 1, sig.level = 0.05, power = 0.8, type = "one.sample")$n
   power.t.test(delta = 0.25, sd = 1, sig.level = 0.05, power = 0.8, type = "one.sample")$n
   ```
   Halving the effect size roughly *quadruples* the required sample size — small
   effects are expensive to detect. This is the single most useful intuition to
   carry out of a power calculation.
   :::

## Summary

You can now:

- Explain why estimating the standard deviation from a small sample pushes us from
  the normal distribution onto the t-distribution, and what "fatter tails" buys
  you.
- Distinguish a Z-score from a t-score and turn either into a p-value with `pnorm`
  and `pt`.
- Use simulation to *check* that a test is honest — uniform p-values under the
  null — rather than taking it on trust.
- Run one-sample, two-sample, paired, and formula-based tests with `t.test()`, and
  recognize the two-sample t-test as a linear model.
- Estimate power and required sample size with `power.t.test()` and by simulation.

The thread tying it all together: when you estimate uncertainty from your data
instead of knowing it, you have to pay for that estimate — and the t-distribution
is how statistics keeps the books honest.

## Resources

- The [pwr package](https://cran.r-project.org/web/packages/pwr/vignettes/pwr-vignette.html)
  for a fuller toolkit of power calculations.
- The American Statistical Association's statement on p-values
  [@doi:10.1080/00031305.2016.1154108] for what a p-value does and doesn't tell
  you.
