In this chapter, we will explore the concept of probability distribution, with
a particular focus on the normal distribution, using the example of IQ
(intelligence quotient) scores. We will demonstrate the use of R functions to
work with the normal distribution, and provide examples and interpretations of
the results.

## Normal Distribution, visualized

The normal distribution, also known as the Gaussian distribution, is a
continuous probability distribution characterized by its bell-shaped curve. It
is defined by two parameters: the mean (µ) and the standard deviation (σ). The
mean represents the central tendency of the distribution, while the standard
deviation represents the dispersion or spread of the data.

We've all seen a picture of the normal distribution. But what does the normal
bell-shaped curve actually represent? 

```{r fig-normal-distribution, fig.cap="The Normal Distribution. When we make a picture that looks like this, we are looking at the probability density function, or sometimes just the `density`.", echo=FALSE}
z = seq(-4, 4, by = 0.01)
plot(z, dnorm(z), type = "l", xlab = "z", ylab = "Density")
```

- What is the mean of _this_ normal distribution?
- What is the standard deviation of _this_ normal distribution?
- What are the values of the density function at z = 0, z = 1, and z = 2?
- What is another term for the values on the x-axis? (Hint: it starts with a "z".)

The density of the normal distribution is highest at the mean, and decreases
symmetrically in both directions. The area under the curve represents the
probability of observing a value within a certain range. For example, the area
under the curve between z = -1 and z = 1 represents the probability of
observing a value between -1 and 1. This area is equal to 0.6827, or 68.27%.

```{r}
# Prep the distribution data
x <- seq(-4, 4, 0.005)
y <- dnorm(x)
normal <- as.data.frame(cbind(x,y))
```

## IQ Scores

The IQ scores are an excellent example of the normal distribution, as they are
designed to follow this distribution pattern. The mean IQ score is set at 100,
and the standard deviation is set at 15. This means that the majority of the
population (about 68%) have an IQ score between 85 and 115, while 95% of the
population have an IQ score between 70 and 130.

## R Functions for Normal Distribution

There are four main functions in R to work with the normal distribution (or any distribution, for that matter): `pnorm`,
`qnorm`, `dnorm`, and `rnorm`. We will describe each function and provide examples of
their usage and interpretation.

pnorm: This function calculates the cumulative distribution function (CDF) for the normal distribution. It gives the probability of observing a value less than or equal to a specified value (x).
Example: What is the probability of having an IQ score of 120 or lower?
pnorm(120, mean = 100, sd = 15)  
 
Interpretation: The output value represents the probability of having an IQ score of 120 or lower. In this case, the probability is approximately 0.91, meaning that about 91% of the population have an IQ score of 120 or lower.
qnorm: This function calculates the quantiles of the normal distribution. It returns the value (x) corresponding to a specified probability (p).

Example: What is the IQ score at the 90th percentile?
qnorm(0.9, mean = 100, sd = 15)  
 
Interpretation: The output value represents the IQ score at the 90th percentile. In this case, the IQ score is approximately 119. This means that 90% of the population have an IQ score of 119 or lower.
dnorm: This function calculates the probability density function (PDF) for the normal distribution. It gives the probability density (height of the curve) at a specified value (x).

Example: What is the probability density of having an IQ score of 110?
dnorm(110, mean = 100, sd = 15)  
 
Interpretation: The output value represents the probability density of having an IQ score of 110. In this case, the density is approximately 0.024. Note that this value does not represent a probability, but rather the height of the curve at the specified value.
rnorm: This function generates random numbers from a normal distribution with specified mean and standard deviation.

Example: Generate 10 random IQ scores.
rnorm(10, mean = 100, sd = 15)  
 
Interpretation: The output will be a vector of 10 random IQ scores generated from a normal distribution with a mean of 100 and a standard deviation of 15. These values can be used for simulation purposes or to create a sample dataset for further analysis.

In conclusion, the normal distribution is a fundamental concept in probability and statistics, and the IQ scores provide a practical example of its application. The R functions pnorm, qnorm, dnorm, and rnorm offer powerful tools to work with the normal distribution, allowing for the calculation of probabilities, quantiles, densities, and the generation of random samples.
