---
title: "Worked example: predicting age from DNA methylation"
---

How old is a tissue sample? Not the donor's age on a birthday cake, but the
biological age written into the DNA itself. It turns out you can read it off the
**DNA-methylation** pattern: at certain sites the methylation level drifts
steadily with age, and a model trained on those sites becomes an **epigenetic
clock**. This example builds one, following the work of [Naue et al.
(2017)](https://www.sciencedirect.com/science/article/pii/S1872497317301643).

This is the second of three worked examples. Unlike the
[cancer-classification example](ml_cancer.qmd), where the target was a category,
here the target — a person's age — is a *number*, so this is a **regression**
problem. We use the same `mlr3` framework throughout; if you need the refresher on
tasks, learners, measures, and the train/test split, see
[The mlr3verse](mlr3verse_intro.qmd) and the
[cancer example](ml_cancer.qmd) for the general workflow. Here we keep our eyes on
the biology and on how regression differs from classification.

The three core ideas from the cancer chapter all return, but each takes on a new
shade once the goal is an epigenetic clock rather than a cancer label:

- **The split is about *validity*, not just accuracy.** A clock that predicts age
  beautifully on the very people it was trained on is worthless as a biomarker — of
  course it fits them, it learned them. The whole point of an epigenetic clock is to
  estimate the age of *someone new*, so the held-out test set is not a nicety here;
  it is the only thing that tells you whether you have built a real biomarker or an
  elaborate lookup table.
- **Feature selection is forced on you.** A methylation array measures hundreds of
  thousands of CpG sites, but a study has only dozens or hundreds of people. Far
  more features than samples — the *p* ≫ *n* regime — is the normal state of affairs
  in genomics, and it means selecting or limiting features is not an optional tidy-up
  but a requirement (see the warning below).
- **The yardstick changes.** Classification used accuracy and a confusion matrix.
  Regression has no "correct/incorrect" — instead we measure how far off the
  predictions are, with **RMSE** (typical error, in years) and **R²** (the fraction
  of the variation in age the model explains).

::: {.callout-warning}
## More features than samples: the *p* ≫ *n* problem

When the number of features *p* dwarfs the number of samples *n*, a flexible model
can find some combination of CpG sites that fits the training ages *perfectly* by
chance alone — pure memorization dressed up as signal. This is overfitting in its
most dangerous form, because the training score looks fantastic. The defences are
the same ones you already know: limit the features (keep only the CpG sites most
associated with age), prefer simpler models, and *always* judge on the held-out
test set. The next chapter shows a second defence — letting the model shrink its
own coefficients.
:::

## What you'll learn

- Build a **regression** task from a real methylation dataset with
  `as_task_regr()`.
- Fit and compare three regression learners — **linear regression**, a
  **regression tree**, and a **random forest** — through the same `mlr3` calls.
- Read **regression diagnostic plots** and a **truth-versus-prediction** scatter.
- Interpret **R²** as the fraction of variation explained (and **RMSE** as the
  typical error in years), and spot overfitting when the test R² falls below the
  training R².
- Explain why having far more CpG features than people (the *p* ≫ *n* problem)
  makes feature selection and an honest test set non-negotiable for a biomarker.

## Setup

```{r message=FALSE}
library(mlr3verse)
library(ranger)       # for random forests
library(data.table)
library(ggplot2)
set.seed(789)
```

## The data

The biology behind it: within each individual, the level of
[DNA methylation](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3174260/) at
certain sites changes steadily with age. The
[CpG sites](https://en.wikipedia.org/wiki/CpG_site) whose methylation correlates
most strongly with age make excellent **biomarkers** — and excellent features for
a regression model. The data come from
[this Galaxy tutorial](https://training.galaxyproject.org/training-material/topics/statistics/tutorials/regression_machinelearning/tutorial.html).^[The
two files are published on Zenodo at
<https://zenodo.org/records/2545213> (`test_rows_labels.csv` and
`train_rows.csv`). We read copies bundled with the book under `data/` so the
chapter renders without a network download, but you can also read them straight
from those URLs with `fread()`.]

```{r message=FALSE, warning=FALSE}
# The data are vendored under data/ (originally from the Galaxy regression
# tutorial, Zenodo record 2545213) so the chapter renders without a network
# download. fread() auto-detects the tab separator.
meth_age <- rbind(
    fread('../data/methylation_age_test.csv'),
    fread('../data/methylation_age_train.csv')
)
```

A quick look at the data:

```{r}
head(meth_age)
```

Each row is a person, each `cg…` column is the methylation level at one CpG site,
and `Age` is what we want to predict. Because the target is a *number*, this is a
**regression** task — so we use `as_task_regr()`.

```{r}
task <- as_task_regr(meth_age, target = 'Age')
```

We split into training and test sets exactly as in the
[cancer example](ml_cancer.qmd), seeding for reproducibility.

```{r}
set.seed(7)
train_set <- sample(task$row_ids, 0.67 * task$nrow)
test_set <- setdiff(task$row_ids, train_set)
```

## Linear regression

We start with the simplest regression model: a straight-line fit relating
methylation to age.

```{r}
learner <- lrn("regr.lm")
```

### Train

```{r}
learner$train(task, row_ids = train_set)
```

After fitting a linear model, the standard **diagnostic plots**
(@fig-regression-diagnostic) help check whether the model's assumptions hold.

```{r fig-regression-diagnostic, fig.cap="Regression diagnostic plots. Top left: residuals vs. fitted values (look for no pattern). Top right: normal Q-Q plot (points should follow the line). Bottom left: scale-location. Bottom right: residuals vs. leverage."}
par(mfrow = c(2, 2))
plot(learner$model)
```

### Predict

```{r}
pred_train <- learner$predict(task, row_ids = train_set)
pred_test <- learner$predict(task, row_ids = test_set)
```

### Assess

Printing a regression prediction shows the predicted (`response`) and true
(`truth`) values side by side.

```{r}
pred_train
```

A scatter plot of truth versus prediction makes the fit visible: a perfect model
would put every point on the diagonal.

```{r}
ggplot(pred_train, aes(x = truth, y = response)) +
    geom_point() +
    geom_smooth(method = 'lm')
```

We summarize the fit with **R²** (the fraction of variation in age the model
explains; 1.0 is perfect, 0 is no better than guessing the mean).

```{r}
measures <- msrs(c('regr.rsq'))
pred_train$score(measures)
```

```{r}
pred_test
pred_test$score(measures)
```

The training R² is high, but the **test R²** is the honest measure of how well
the clock would predict the age of a new person. A drop from training to test is,
again, the signature of overfitting.

## Regression tree

We can use a tree for regression too. Instead of voting on a class, each leaf of
the tree reports a predicted *number*.

```{r}
learner <- lrn("regr.rpart", keep_model = TRUE)
learner$train(task, row_ids = train_set)
learner$model
```

Notice something odd: a single regression tree can only output a handful of
**discrete** age values — one per leaf. It cannot predict, say, 43.7 if no leaf
sits there.

### Predict

```{r}
pred_train <- learner$predict(task, row_ids = train_set)
pred_test <- learner$predict(task, row_ids = test_set)
```

### Assess

```{r}
pred_train
```

The discreteness is obvious in the scatter plot: predictions stack up into
horizontal bands, one per leaf.

```{r}
ggplot(pred_train, aes(x = truth, y = response)) +
    geom_point() +
    geom_smooth(method = 'lm')
```

```{r}
measures <- msrs(c('regr.rsq'))
pred_train$score(measures)
```

```{r}
pred_test
pred_test$score(measures)
```

The R² differs noticeably from the linear model — the single tree's blocky
predictions are a poor fit for a continuous quantity like age.

## Random forest

A random forest is also tree-based, but averaging over many trees smooths away
the blocky, discrete behaviour of a single tree, giving genuinely continuous
predictions.

```{r}
set.seed(789)
learner <- lrn("regr.ranger", mtry = 2, min.node.size = 20)
learner$train(task, row_ids = train_set)
learner$model
```

### Predict

```{r}
pred_train <- learner$predict(task, row_ids = train_set)
pred_test <- learner$predict(task, row_ids = test_set)
```

### Assess

```{r}
pred_train
```

```{r}
ggplot(pred_train, aes(x = truth, y = response)) +
    geom_point() +
    geom_smooth(method = 'lm')
```

The points now scatter smoothly around the diagonal — no more horizontal bands.

```{r}
measures <- msrs(c('regr.rsq'))
pred_train$score(measures)
```

```{r}
pred_test
pred_test$score(measures)
```

Compare the test R² across the three regression models. The forest typically
predicts age more accurately than either the linear model or the single tree, and
without the single tree's discrete artifacts.

## Summary

You have now built an epigenetic clock — a complete **regression** analysis on
real methylation data, through the same `mlr3` workflow you used for cancer
classification:

- You turned the methylation table into an `mlr3` regression **task** with
  `as_task_regr()` and split it into training and test sets.
- You fit a **linear regression**, a **regression tree**, and a **random forest**,
  and judged each with **R²** and a truth-versus-prediction scatter.
- The single tree's predictions came in discrete bands — a poor fit for a
  continuous target — while the random forest's smooth predictions typically beat
  both the straight line and the single tree.

The change from the cancer example was only the *kind* of target: a number instead
of a category, so `as_task_regr()` instead of `as_task_classif()` and R² instead
of accuracy. The discipline of trusting only the test-set score carries over
unchanged.

## Exercises

Each exercise reuses objects you already built, so the solutions are quick to run.

1. **Tune the forest.** The random forest used `mtry = 2` and
   `min.node.size = 20`. Refit it with the defaults (drop both arguments) and
   compare the test R². Did the hand-chosen settings help?

   ::: {.callout-note collapse="true"}
   ## Solution

   ```{r eval=FALSE}
   rf_default <- lrn("regr.ranger")
   rf_default$train(task, row_ids = train_set)
   rf_default$predict(task, row_ids = test_set)$score(msrs(c('regr.rsq')))
   ```

   Changing `mtry` and `min.node.size` is a one-line edit to the `lrn()` call;
   everything else is identical. On this dataset the defaults are usually close,
   which is part of why random forests are a low-fuss default — but the honest
   comparison is always the held-out test R².
   :::

2. **Add a second measure.** Alongside R², score the test predictions with
   root-mean-squared error, `msr("regr.rmse")`, which reports the typical
   prediction error in *years*. Which number is easier to explain to a biologist?

   ::: {.callout-note collapse="true"}
   ## Solution

   ```{r eval=FALSE}
   pred_test$score(msrs(c('regr.rsq', 'regr.rmse')))
   ```

   R² is unitless and says what *fraction* of the variation in age the model
   explains; RMSE is in the target's own units (years) and says how far off a
   typical prediction is. For an audience that thinks in years, RMSE is often the
   more intuitive headline number.
   :::
