---
title: "The mlr3verse framework"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

In the [previous chapter](models.qmd) you ran a whole tour of supervised
algorithms — linear regression, k-nearest neighbours, penalized regression,
decision trees, random forests — on the Palmer penguins. Each one came from a
different R package with its own habits: `lm()` wants a formula, `glmnet()` wants a
matrix, `knn()` wants the training and test sets handed in separately. Swapping one
algorithm for another meant rewriting the surrounding plumbing every time.

That friction is exactly what the **mlr3verse** removes. It wraps every algorithm in
one consistent interface, so the question "what happens if I try a random forest
instead of a tree?" becomes a one-line change instead of a rewrite. This chapter is a
guided tour of the four object types that interface is built from. We'll meet them on
the same penguins you already know — predicting a bird's **species** from its
morphological measurements — so you can focus on the framework rather than on a new
dataset.

## What you'll learn

- Explain the four core mlr3 object types — **Task**, **Learner**, **Measure**,
  **Resampling** — and what each one is responsible for.
- Build a classification **Task** from a data frame and read off its target and
  features with `task$target_names` and `task$feature_names`.
- Discover available algorithms in the **learner registry** with `mlr_learners$keys()`
  and create one with `lrn()`.
- Inspect a learner's tunable **hyperparameters** with `learner$param_set`.
- Choose a **Measure** with `msr()` and a **Resampling** strategy with `rsmp()`.

## Why a framework at all?

The R ecosystem offers dozens of machine-learning packages, each with its own
interface and conventions. That diversity is powerful, but it also means you spend
time relearning a new API for every algorithm and rewriting the same data-preparation
code over and over. The mlr3verse fixes this by giving you one consistent,
object-oriented framework that all the algorithms plug into.

The design follows a simple idea: **separate the moving parts**. Instead of one giant
function that does everything from preprocessing to evaluation, mlr3 splits the
workflow into four kinds of object, each with one job.

- **Tasks** hold the data and define the problem (what are we predicting?).
- **Learners** are the algorithms (how do we learn the rule?).
- **Measures** score performance (how good is the result?).
- **Resamplings** control how data is split for honest validation (did it generalize?).

::: {.callout-important}
## The big idea: interchangeable parts

Think of mlr3's four object types as the standardized, swappable components of a
machine. A Task is the workpiece, a Learner is a tool head, a Measure is the gauge you
check the result with, and a Resampling is the jig that holds everything for a fair
test. Because every component shares the same fittings, you can pop out the
decision-tree tool head and drop in a random forest without rebuilding the machine.
That single design choice — common fittings everywhere — is what makes "try another
algorithm" a one-line edit instead of a rewrite. Learn the four sockets once and every
algorithm in the ecosystem clicks into them the same way.
:::

The framework is built on R6 classes, which is why you'll see the `object$method()`
syntax (a dollar sign and parentheses) throughout, rather than the
`function(object)` style you may be used to. It also takes reproducibility seriously:
every random step can be pinned down with `set.seed()`, which we'll do whenever a step
involves chance.

## The mlr3verse ecosystem

`mlr3verse` is a meta-package: loading it pulls in a family of focused packages that
build on the core. You rarely need to load them individually.

| Package | Purpose |
|---|---|
| `mlr3` | Core framework: tasks, learners, measures, resampling |
| `mlr3learners` | Interfaces to popular algorithms (random forest, glmnet, xgboost, …) |
| `mlr3pipelines` | Preprocessing and model-composition pipelines |
| `mlr3tuning` | Hyperparameter optimization |
| `mlr3measures` | Additional evaluation metrics |
| `mlr3viz` | Plotting utilities for models and results |

For this tour we only need the core ideas, all reachable through `mlr3verse`. Let's
load it along with the penguins.

```{r setup-data}
library(mlr3verse)
library(palmerpenguins)
library(data.table)

data(penguins)
penguins <- na.omit(penguins)   # drop the few rows with missing measurements
```

We dropped the rows with missing values up front so the examples stay simple; in a
real analysis you might instead impute them.

## Tasks: the problem and the data

A **Task** bundles your data together with metadata about the problem: which column is
the **target** (the answer you want to predict) and which columns are the **features**
(the inputs you predict from). A *classification* task has a categorical target; a
*regression* task has a continuous one. The Task also figures out feature types for you
and offers helpers for filtering rows and selecting columns.

Our biological question is the same one from the previous chapter: can we predict a
penguin's **species** from its body measurements? That makes `species` the target.

```{r task-example}
task <- as_task_classif(penguins, target = "species")

task$nrow            # number of penguins (rows)
task$ncol            # number of columns, including the target
task$feature_names   # the predictor columns
task$target_names    # the column we're predicting
```

Reading that output top to bottom: the task holds `r nrow(penguins)` penguins, the
`feature_names` are the morphological and contextual measurements we predict *from*
(bill length and depth, flipper length, body mass, island, sex, year), and
`target_names` confirms that `species` is what we're predicting. Notice we never told
mlr3 which columns were features — it took every column except the target
automatically. That bookkeeping is the Task's job, so you don't have to track it
yourself.

## Learners: the algorithms

A **Learner** is an algorithm wrapped in the standard mlr3 interface. Every learner
declares what it can do — which task types it supports, whether it tolerates missing
values, and which hyperparameters it exposes — and provides the same `train()` and
`predict()` methods regardless of the maths underneath.

You discover learners through the **registry**, `mlr_learners`. Querying it for
classification learners answers the practical question "what algorithms can I even try
on this task?"

```{r learner-exploration}
classif_learners <- mlr_learners$keys("classif")
head(classif_learners, 10)
```

Each entry is the id of an algorithm you can use on a classification task — for
example `classif.rpart` (a decision tree) or `classif.ranger` (a random forest). This
is the menu the framework offers for our species-prediction problem.

::: {.callout-note}
## Learner naming convention

mlr3 names learners as `[task_type].[algorithm]`. So `classif.rpart` is a decision
tree for classification, while `regr.lm` is linear regression for a regression task.
Once you know this scheme you can guess a learner's name before you look it up.
:::

Let's create the decision-tree learner and inspect its hyperparameters — the dials you
can turn to change how it learns.

```{r learner-params}
rpart_learner <- lrn("classif.rpart")
rpart_learner$param_set$ids()   # the hyperparameters you can tune
```

These are `rpart`'s tuning dials. A few you'll meet again: `cp` is the complexity
penalty that controls how aggressively the tree is pruned (larger means a simpler
tree), `maxdepth` caps how deep the tree may grow, and `minsplit` sets the smallest
node the algorithm is willing to split. You don't have to set any of them — the learner
ships with sensible defaults — but this is where you'd reach to fight overfitting.

## Measures: scoring the result

A **Measure** is the yardstick you use to judge predictions. The choice matters,
because different measures reward different things, and the measure you optimize is the
behaviour you'll get. For classification you'll commonly reach for accuracy or its
mirror image, classification error.

```{r measures-example}
acc_measure <- msr("classif.acc")   # accuracy: fraction predicted correctly
ce_measure  <- msr("classif.ce")    # classification error: fraction wrong (1 - accuracy)

acc_measure
```

Printing a measure tells you its key facts at a glance: its id (`classif.acc`), the
range of values it can take (`[0, 1]` for accuracy), and `Minimize: FALSE` — meaning
for *this* measure bigger is better, so the framework knows to maximize it when tuning.
Classification error would show `Minimize: TRUE`, because there you want the number as
small as possible.

## Resamplings: honest validation

The cardinal rule from the previous chapters: never judge a model on the same data it
trained on. A **Resampling** object encodes *how* you split the data to keep yourself
honest. The common strategies are a single **holdout** split, **k-fold
cross-validation** (split into k parts, train on k−1 and test on the held-out part,
rotate), and the **bootstrap** (resample with replacement many times).

```{r resampling-example}
set.seed(42)
holdout   <- rsmp("holdout", ratio = 0.8)      # 80% train, 20% test
cv5       <- rsmp("cv", folds = 5)             # 5-fold cross-validation
bootstrap <- rsmp("bootstrap", repeats = 30)   # 30 bootstrap resamples

cv5$param_set
```

We seeded the chunk because holdout and bootstrap make random splits, and we want the
same split every time the book is rebuilt. The printed `param_set` for `cv5` lists the
strategy's own settings: `folds` is the number of cross-validation folds (we set it to
5), and `repeats` is how many times the whole k-fold procedure is repeated (1 by
default). Together these control the trade-off between a reliable performance estimate
and how much computation it costs.

## How the four fit together

Each object owns one part of the workflow, and they snap together in a predictable way:
a **Learner** trains on a **Task** to produce a fitted model, that model makes
predictions, a **Measure** scores those predictions, and a **Resampling** strategy
wraps the whole thing so the score reflects performance on data the model has never
seen. Because all learners share the same interface, swapping `classif.rpart` for
`classif.ranger` changes one string and nothing else.

::: {.callout-important}
## Why this design pays off

- **Consistency** — every algorithm uses the same `train()`/`predict()` interface, so
  there's one set of habits to learn, not one per package.
- **Flexibility** — mix and match tasks, learners, measures, and resamplings to run
  experiments by editing a single component.
- **Reproducibility** — randomness is controllable, so results are repeatable.
- **Best practices baked in** — the design nudges you toward train/test splitting and
  cross-validation rather than leaving them as an afterthought.
:::

You've now met all four mlr3 building blocks on a problem you already understand. In the
[next chapter](machine_learning_mlr3.qmd) these pieces come together on real biological
data — classifying cancer types from gene expression, predicting age from DNA
methylation — where you'll see a Learner actually train on a Task and have its
predictions scored by a Measure under a Resampling. The concepts are the same; only the
data gets bigger and more interesting.

## Summary

You can now:

- Name the four core mlr3 objects and the single job each one does — **Task** (data +
  problem), **Learner** (algorithm), **Measure** (score), **Resampling** (validation
  strategy).
- Build a classification Task with `as_task_classif()` and inspect it with
  `task$feature_names` and `task$target_names`.
- List available algorithms with `mlr_learners$keys()`, create one with `lrn()`, and
  see its tunable dials with `learner$param_set$ids()`.
- Pick a Measure with `msr()` and a Resampling strategy with `rsmp()`, and read what
  their printouts are telling you.

The payoff is the interchangeable-parts design: learn the four sockets once, and every
algorithm in the mlr3 ecosystem plugs into them the same way.

## Exercises

These exercises reuse the penguins. Recall that in the previous chapter we also framed
a **regression** problem — predicting a penguin's `body_mass_g` from its other
measurements. Let's set that up in mlr3.

1. **Build a regression task.** Create a regression task that predicts `body_mass_g`,
   using `as_task_regr()`. Confirm the target and list the features.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   task_mass <- as_task_regr(penguins, target = "body_mass_g")
   task_mass$target_names
   task_mass$feature_names
   ```
   `as_task_regr()` is the regression counterpart of `as_task_classif()`. The target is
   now the continuous `body_mass_g`, and `species` has joined the feature list — any
   column that isn't the target becomes a feature.
   :::

2. **Find regression learners.** Query the registry for learners that work on a
   regression task, and show the first ten.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   regr_learners <- mlr_learners$keys("regr")
   head(regr_learners, 10)
   ```
   These are the `regr.*` algorithms — for example `regr.lm` (linear regression) and
   `regr.rpart` (a regression tree). The naming convention tells you each one's task
   type at a glance.
   :::

3. **Set up 10-fold cross-validation.** Create a resampling object for 10-fold
   cross-validation and confirm the fold count from its parameter set.

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   cv10 <- rsmp("cv", folds = 10)
   cv10$param_set$values
   ```
   `rsmp("cv", folds = 10)` builds the strategy; `param_set$values` reports the settings
   you actually chose, confirming `folds` is 10. More folds give a more stable estimate
   of performance at the cost of more computation.
   :::

4. **Pick a regression measure.** Create the root-mean-squared-error measure with
   `msr("regr.rmse")` and print it. Does the framework want it minimized or maximized?

   ::: {.callout-note collapse="true"}
   ## Solution
   ```{r}
   rmse_measure <- msr("regr.rmse")
   rmse_measure
   ```
   The printout shows `Minimize: TRUE` — RMSE is an error, so smaller is better, and the
   framework will minimize it when comparing models. Contrast this with accuracy, which
   it maximizes.
   :::
