---
title: "Integrating single-cell ATAC and RNA"
author:
  - name: Sean Davis
    url: https://seandavi.github.io/
    affiliation: University of Colorado<br/>Anschutz School of Medicine
    affiliation-url: https://medschool.cuanschutz.edu/
    email: seandavi@gmail.com
---

A single-cell RNA experiment tells you what each cell *is doing*: which genes it is actively transcribing, right now, in the moment it was captured. A single-cell ATAC experiment tells you something different and complementary — which stretches of the genome are *open for business*. ATAC-seq finds the regions of chromatin a cell has unpacked from its tightly wound default state: promoters it has switched on, enhancers it is using, the regulatory switches it has flipped. Expression is the output; accessibility is the wiring that produces it. Measure only one and you are reading half the circuit.

Why does the wiring matter if you can already see the output? Because a cell often opens a regulatory region *before*, or *without*, transcribing the gene it controls. A stem cell on the verge of becoming a blood cell can unlock the enhancers of blood-cell genes while those genes are still essentially silent — the chromatin is poised, the decision half-made, but the RNA hasn't caught up yet. Accessibility therefore carries information about a cell's *trajectory and regulatory state* that expression alone misses: which transcription factors are active, which enhancers drive which genes, where a cell is heading rather than only where it is.

The technology that makes this concrete is the **multiome** assay (here, 10x Genomics' Single Cell Multiome ATAC + Gene Expression), which measures *both* RNA and open chromatin in the *same nucleus*. That "same nucleus" is the crucial part. With separate experiments you can only compare populations — the average RNA of one batch of cells against the average accessibility of another. With a multiome you can ask a cause-and-effect question inside a single cell: *this* enhancer is open, and *this* gene, which it controls, is expressed, in this very cell. That is what lets you move from cataloguing cell types to reconstructing the regulatory logic that defines them.

This chapter is about **integration**: combining the two modalities into one coherent analysis instead of two parallel ones. We will build the intuition for what integration means, why it is harder than it sounds, and then do it hands-on with a public 10x PBMC multiome dataset that ships ready-to-use as a Bioconductor `MultiAssayExperiment`.

## What you'll learn

- Why chromatin accessibility (ATAC) and gene expression (RNA) are complementary readouts of the same cell, and what each one uniquely tells you.
- The three shapes of single-cell integration — **vertical**, **diagonal**, and **mosaic** — and how knowing which one you have decides which methods apply.
- The **feature-space problem**: RNA is measured over genes, ATAC over genomic peaks, and why that mismatch is the central obstacle to integration.
- The **bridges** that connect the two spaces: gene-activity scores, peak-to-gene links, and transcription-factor motif activity.
- How to load and explore a real multiome dataset as a `MultiAssayExperiment` of two `SingleCellExperiment`s, and process each modality on its own terms.
- How to build a **joint embedding** that uses both modalities at once, and what it buys you over either modality alone.

This chapter builds directly on the [single-cell RNA-seq chapter](setup.qmd) (the `SingleCellExperiment`, normalization, HVGs, PCA/UMAP, clustering) and on the [bulk ATAC-seq chapter](../atac-seq/atac-seq.qmd) (peaks, fragments, what accessibility *is*). If either is hazy, skim it first — we lean on both.

## Two views of the same cell

It helps to be precise about what each assay actually counts.

- **scRNA-seq** counts transcripts and assigns them to **genes**. A cell becomes a vector of ~20,000–30,000 gene counts. High counts for a gene mean that gene is being transcribed.
- **scATAC-seq** counts sequencing fragments that fall in accessible chromatin and assigns them to **peaks** — genomic intervals (typically a few hundred base pairs) that are open across many cells. A cell becomes a vector of counts over ~100,000+ peaks. A nonzero count for a peak means that region was accessible in that cell.

Two consequences follow immediately, and they drive everything else in the chapter:

1. **The two modalities live in different feature spaces.** One is indexed by genes, the other by genomic coordinates. You cannot simply stack a gene-by-cell matrix on top of a peak-by-cell matrix and call it integrated — the rows mean different things.
2. **ATAC data is sparse and nearly binary.** Each cell has at most two copies of any locus, so a peak is essentially "seen" or "not seen" in a given cell. The peak matrix is far sparser than the RNA matrix, which changes how we normalize and reduce it (more on this below — it is why ATAC uses TF-IDF and LSI rather than the log-normalization + PCA you used for RNA).

## What "integration" means: vertical, diagonal, mosaic

"Integrate the ATAC with the RNA" can mean several genuinely different problems, and the single most useful thing you can do before reaching for a tool is to figure out *which one you have*. The cleanest way to organize them (after @argelaguet2021computational) is by **what the two datasets share**:

| Shape | What's shared | Typical source | The hard part |
|---|---|---|---|
| **Vertical** | the *same cells*, different modalities | 10x **Multiome** (RNA + ATAC per nucleus) | nothing to match — just *combine* the two views well |
| **Diagonal** | different cells, different modalities | a scRNA run and a *separate* scATAC run | you must **infer** which cells correspond |
| **Mosaic** | partial overlap | some cells multiome, some single-modality | bridge the measured and the missing |

The distinction is not academic — it changes the method entirely:

- In **vertical** integration the correspondence is *handed to you*. Cell 5's RNA and cell 5's ATAC are the same cell, by construction. The task is to learn a single representation that respects both views — a *joint embedding*. This is the easier, intuition-building case, and where we start.
- In **diagonal** integration there is no correspondence at all. You have RNA for one set of cells and ATAC for a different set, and nothing tells you that an RNA cell and an ATAC cell are "the same kind." You must *infer* the alignment, almost always by projecting both into a shared feature space (the bridges below) and matching. This is the case that forces you to understand *why* integration is hard, and we return to it in the next chapter using label transfer.

Your collaborators will bring you both. Someone with a 10x Multiome run has a vertical problem. Someone who ran scRNA last year and scATAC this year — same tissue, different cells — has a diagonal one. Same words, "integrate ATAC with RNA," completely different machinery.

## The feature-space problem and the bridges

Because RNA lives in gene space and ATAC in peak space, every integration method needs a **bridge** that puts the two modalities into a common frame. There are three bridges worth knowing, and they are not interchangeable — each answers a different question:

- **Gene-activity scores.** Summarize a gene's accessibility by summing the ATAC signal over its body and promoter (or, better, distance-weighted nearby peaks). This *casts ATAC into gene space*, so an ATAC cell can be compared to an RNA cell gene-for-gene. It is crude — accessibility is not expression — but it is the workhorse bridge for *aligning* unpaired datasets and for *annotating* ATAC cells from an RNA reference.
- **Peak-to-gene links.** Correlate a peak's accessibility against a gene's expression *across cells*. A peak whose opening tracks a gene's expression is a candidate regulatory element for that gene. This is less a bridge for alignment than a *biological result* — it is how you discover enhancer→gene wiring, and it needs paired (or pseudobulked) data to compute.
- **Motif / TF activity.** Collapse the tens of thousands of peaks into the handful of transcription factors whose binding motifs they contain (e.g. with `chromVAR`). This turns the unwieldy peak space into a small, interpretable space of TF activities — and TFs are the natural currency for connecting open chromatin back to the genes it regulates.

Keep these three straight and most of the integration literature snaps into focus: methods differ mainly in *which bridge they use* and *whether they need the cells to be paired*.

## The data: 10x PBMC multiome

We'll use a public 10x Genomics dataset of roughly 10,000 peripheral blood mononuclear cells (PBMCs) from a healthy donor, profiled with the Single Cell Multiome assay so that every nucleus has both an RNA and an ATAC measurement. It is a good teaching set: PBMCs contain well-characterized, easily recognized immune cell types (T cells, B cells, monocytes, NK cells), so we have strong priors about what the answer should look like.

Rather than download and reprocess the raw 10x files, we use the curated copy from the Bioconductor `SingleCellMultiModal` package, which delivers the dataset as a `MultiAssayExperiment` — the natural container for "several assays measured on the same cells," and the multi-modal generalization of the `SummarizedExperiment`/`SingleCellExperiment` classes you have already met.

```{r}
#| label: load-packages
#| message: false
library(SingleCellMultiModal)   # the curated 10x PBMC multiome dataset
library(MultiAssayExperiment)   # the multi-modal container
library(SingleCellExperiment)   # each modality is an SCE
library(Seurat)                 # WNN joint embedding lives here
library(Signac)                 # ATAC processing (TF-IDF / LSI) and ChromatinAssay
```

A single call fetches the data (the first time, it downloads from Bioconductor's ExperimentHub and caches locally; afterwards it loads from the cache):

```{r}
#| label: load-data
mae <- scMultiome("pbmc_10x", modes = "*", dry.run = FALSE, format = "MTX")
mae
```

The object bundles two experiments measured on the **same 10,032 cells**:

- `rna` — a `SingleCellExperiment` of gene expression counts (~36,500 genes).
- `atac` — a `SingleCellExperiment` of peak accessibility counts (~108,000 peaks).

Because they share the same cells, the `MultiAssayExperiment` keeps their column (cell) metadata aligned for us — this is exactly the bookkeeping that makes *vertical* integration tractable. We pull out the two modalities to work with each on its own terms:

```{r}
#| label: split-modalities
sce.rna  <- experiments(mae)[["rna"]]
sce.atac <- experiments(mae)[["atac"]]

sce.rna
sce.atac
```

## From Bioconductor to Seurat: building the joint object

The joint-embedding method we'll use — **weighted nearest neighbors (WNN)** — lives in the Seurat/Signac ecosystem, which is the de-facto standard for paired multiome. So our first practical step is to bridge our `MultiAssayExperiment` into a Seurat object: the RNA counts become a standard expression assay, and the ATAC peak counts become a Signac **`ChromatinAssay`**. This Bioconductor→Seurat hand-off is worth learning in its own right — you will reach for it constantly, because single-cell data routinely arrives in one ecosystem and the method you want lives in the other.

```{r}
#| label: to-seurat
# RNA becomes a standard expression assay.
pbmc <- CreateSeuratObject(counts = counts(sce.rna), assay = "RNA")

# ATAC becomes a ChromatinAssay, which identifies each peak by its genomic
# coordinates. Peaks can reach us two ways: as a GRanges in the SCE's rowRanges,
# or encoded in the rownames as coordinate strings like "chr1:9776-10668". We
# prefer the GRanges when present and fall back to parsing the rownames.
atac.counts <- counts(sce.atac)
peak.ranges <- tryCatch(rowRanges(sce.atac), error = function(e) NULL)

pbmc[["ATAC"]] <- if (is(peak.ranges, "GRanges") && length(peak.ranges) == nrow(atac.counts)) {
  CreateChromatinAssay(counts = atac.counts, ranges = peak.ranges)
} else {
  CreateChromatinAssay(counts = atac.counts, sep = c(":", "-"))
}
pbmc
```

::: {.callout-note}
## We deliberately skip the fragment file

The raw 10x multiome ships a large *fragments* file (every ATAC fragment, genome-wide). The curated dataset gives us the *peak* matrix instead, so we build the `ChromatinAssay` without fragments. WNN needs only the per-modality reductions, so this costs us nothing here — but it does mean genome-browser coverage tracks and fragment-based gene-activity scores aren't available in this chapter. We don't need them for the joint embedding.
:::

### RNA reduction: the familiar recipe

The expression side is exactly the workflow from the [scRNA-seq chapter](setup.qmd) — normalize, pick highly variable genes, scale, and reduce with PCA:

```{r}
#| label: rna-reduction
DefaultAssay(pbmc) <- "RNA"
pbmc <- NormalizeData(pbmc)
pbmc <- FindVariableFeatures(pbmc, nfeatures = 3000)
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc, reduction.name = "pca")
```

### ATAC reduction: why not PCA, and what TF-IDF + LSI do instead

The accessibility side needs a different reduction, and the reason is the structure of the data. Recall that the peak matrix is **sparse and nearly binary**: a cell has at most a couple of copies of any locus, so most entries are 0 and the rest are small. Log-normalization-then-PCA, which works well for the graded counts of RNA, is a poor fit for this on/off signal.

The standard reduction for ATAC is **TF-IDF followed by LSI**:

- **TF-IDF** (term frequency–inverse document frequency, borrowed from text analysis, where "documents" are cells and "terms" are peaks) up-weights peaks that are informative *because* they are accessible in only a few cells, and down-weights peaks open everywhere. It is how we let rare, cell-type-defining peaks carry more signal than ubiquitous ones.
- **LSI** (latent semantic indexing) is then just a singular value decomposition of the TF-IDF matrix — the accessibility analogue of PCA.

```{r}
#| label: atac-reduction
DefaultAssay(pbmc) <- "ATAC"
pbmc <- FindTopFeatures(pbmc, min.cutoff = 5)   # keep peaks seen in >= 5 cells
pbmc <- RunTFIDF(pbmc)                           # normalize
pbmc <- RunSVD(pbmc, reduction.name = "lsi")     # the SVD that makes it "LSI"
```

There is one wrinkle worth teaching because it trips people up: the **first LSI component usually captures sequencing depth, not biology** — cells with more total fragments score high on it regardless of type. We check this and, if so, exclude component 1 from everything downstream (using dimensions `2:n` instead of `1:n`):

```{r}
#| label: lsi-depthcor
DepthCor(pbmc, reduction = "lsi")
```

## A joint embedding with weighted nearest neighbors

Here is the heart of vertical integration. We have two reductions of the same cells — a PCA from RNA and an LSI from ATAC — and we want **one** representation that uses both. The naive approach (just glue the two sets of components together) treats every cell as trusting both modalities equally. But that is rarely true: for some cells the RNA cleanly says "T cell" while the ATAC is murky, and for others the reverse.

**WNN** handles exactly this. For *each individual cell* it learns a pair of weights — how much to trust RNA versus ATAC — by asking which modality better predicts that cell's near neighbors. It then builds a single neighbor graph from the weighted combination. The output is a joint graph we can embed (UMAP) and cluster, where each cell's contribution from each modality is tuned to that cell.

```{r}
#| label: wnn
pbmc <- FindMultiModalNeighbors(
  pbmc,
  reduction.list = list("pca", "lsi"),
  dims.list = list(1:50, 2:40),        # RNA PCA 1:50; ATAC LSI drops component 1
  modality.weight.name = "RNA.weight"
)

pbmc <- RunUMAP(pbmc, nn.name = "weighted.nn", assay = "RNA", reduction.name = "wnn.umap")
pbmc <- FindClusters(pbmc, graph.name = "wsnn", algorithm = 3, resolution = 0.8)
```

### What did the joint view buy us?

The payoff is clearest when we put all three embeddings side by side — RNA alone, ATAC alone, and the WNN joint view — and look for cell populations that one modality blurs but the other (and so the joint view) resolves:

```{r}
#| label: single-modality-umaps
pbmc <- RunUMAP(pbmc, reduction = "pca", dims = 1:50, reduction.name = "rna.umap")
pbmc <- RunUMAP(pbmc, reduction = "lsi", dims = 2:40, reduction.name = "atac.umap")
```

```{r}
#| label: compare-umaps
#| fig-width: 12
#| fig-height: 4
p1 <- DimPlot(pbmc, reduction = "rna.umap")  + ggtitle("RNA only")
p2 <- DimPlot(pbmc, reduction = "atac.umap") + ggtitle("ATAC only")
p3 <- DimPlot(pbmc, reduction = "wnn.umap")  + ggtitle("WNN (joint)")
p1 + p2 + p3
```

We can also look directly at what WNN learned — the per-cell modality weights. Plotting the RNA weight by cluster shows *which cell types lean on expression and which lean on accessibility* to define themselves:

```{r}
#| label: modality-weights
VlnPlot(pbmc, features = "RNA.weight", group.by = "seurat_clusters", pt.size = 0)
```

## Downstream: linking a peak to the gene it regulates

The unique payoff of *paired* data is that we can connect a specific open region to a specific gene's expression **in the same cells**. The honest, transparent way to see the idea is to compute it by hand for one locus: take a cell-type marker gene, find the accessible peaks near it, and ask whether a peak's accessibility *tracks* the gene's expression across cells. A peak that does is a candidate regulatory element for that gene.

We'll use `MS4A1` (CD20), a canonical B-cell marker. We need its genomic location, which we pull from an Ensembl annotation:

```{r}
#| label: peak-gene-link
#| eval: false
library(EnsDb.Hsapiens.v86)

# 1. Locate MS4A1 and define a +/- 100 kb window around it.
# 2. Find ATAC peaks overlapping that window (peaks already carry coordinates).
# 3. Spearman-correlate each peak's accessibility against MS4A1 expression
#    across all cells; the highest-correlating peaks are candidate enhancers.
#
# This hand-rolled correlation is the transparent core of what Signac's
# LinkPeaks() does with added GC/background correction — which needs a BSgenome
# and the fragment file we deliberately skipped. Code finalized on first render.
```

::: {.callout-tip}
## Where this goes next
Production tools formalize this step: `Signac::LinkPeaks()` adds GC-content and background correction (at the cost of a `BSgenome` and the fragment file), and full gene-regulatory-network methods (SCENIC+, Pando) chain peak→gene links with transcription-factor motifs to reconstruct regulatory circuits. We compute the bare correlation here so the *concept* is unmistakable.
:::

## Summary

- **RNA and ATAC are complementary**: expression is the output, accessibility is the regulatory wiring — and a cell can open a locus before it transcribes the gene.
- **Know your integration shape first.** This was a **vertical** problem: a multiome assay handed us the cell correspondence, so the task was to *combine* two views, not to *match* cells.
- **The two modalities live in different feature spaces** (genes vs. peaks), and the sparse, near-binary ATAC matrix needs **TF-IDF + LSI**, not the log-PCA recipe used for RNA — and its first LSI component usually encodes depth, not biology.
- **WNN** builds a single embedding by learning, per cell, how much to trust each modality — recovering structure that either modality alone can blur.
- **Paired data uniquely lets you link peaks to genes** in the same cells, the first step toward reconstructing regulatory logic.

## What's next: the unpaired case

This chapter handled the *vertical* problem, where the multiome assay handed us the cell correspondence for free. The harder and very common case is *diagonal* integration — a separate scRNA dataset and scATAC dataset with no shared cells — where we must infer the correspondence ourselves. The [next chapter](#) tackles it with **anchor-based label transfer**, and uses a clever trick: we take this very multiome dataset, *throw away* the pairing, and try to recover it — giving us a known ground truth to grade the integration against.
