---
title: "Evidence of use — *The RBioc Book*"
subtitle: "Reproducible provenance for the figures cited in `paper.md`"
author: Sean Davis
date: 2026-06-23
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

This document is the reproducible source for the *Evidence of use* numbers reported in
the JOSE paper (`paper.md`). It summarizes (1) formal teaching offerings and (2) the
online readership recorded by Google Analytics 4 (GA4). It is **not** part of the book
render (the `paper/` directory is listed in `.quartoignore`) and is **not** submitted to
JOSE; it exists so the headline figures are traceable to their source data.

## Formal teaching offerings

The book has been taught, in whole or in part, in the following formal courses.

```{r}
#| label: teaching
library(tibble)
library(knitr)

teaching <- tribble(
  ~Offering,                                  ~Year, ~`Approx. participants`,
  "Cold Spring Harbor Laboratory (CSHL)",     2025L, 25L,
  "Cold Spring Harbor Laboratory (CSHL)",     2026L, 25L,
  "BigCARE",                                  2025L, 25L,
  "BigCARE",                                  2026L, 25L
)

kable(teaching, caption = "Formal offerings of *The RBioc Book*.")

cat(sprintf(
  "Total: %d offerings across %d programs over %d years; ~%d trainees.\n",
  nrow(teaching),
  length(unique(sub(" .*", "", teaching$Offering))),
  length(unique(teaching$Year)),
  sum(teaching$`Approx. participants`)
))
```

## Online readership (GA4)

Source: GA4 export for the property `seandavi.github.io`, page-path report filtered to
`/RBiocBook/*`, for **2024-05-26 → 2026-06-22**. The raw export is archived alongside
this file at `data/ga4_page_views_2024-05-26_to_2026-06-22.csv`.

```{r}
#| label: ga-load
library(readr)
library(dplyr)

ga <- read_csv(
  "data/ga4_page_views_2024-05-26_to_2026-06-22.csv",
  comment = "#",
  show_col_types = FALSE
) |>
  rename(
    page       = `Page path and screen class`,
    views      = Views,
    users      = `Active users`,
    engagement = `Average engagement time per active user`
  )

glimpse(ga)
```

```{r}
#| label: ga-headline
total_views   <- sum(ga$views)
n_pages       <- nrow(ga)
landing_users <- ga |> filter(page == "/RBiocBook/") |> pull(users)

# Per-page average engagement is reported in seconds. Highlight a worked-through
# chapter (single-cell setup) as evidence of depth, not just reach.
sc_setup_sec <- ga |>
  filter(page == "/RBiocBook/single_cell/setup.html") |>
  pull(engagement)

summary_tbl <- tibble::tibble(
  Metric = c(
    "Total page views",
    "Distinct chapters/pages viewed",
    "Readers on the landing page (reach proxy)",
    "Avg. engagement on single-cell setup chapter (min)"
  ),
  Value = c(
    format(total_views, big.mark = ","),
    as.character(n_pages),
    format(landing_users, big.mark = ","),
    sprintf("%.1f", sc_setup_sec / 60)
  )
)

knitr::kable(summary_tbl, caption = "GA4 headline figures (2024-05-26 to 2026-06-22).")
```

```{r}
#| label: ga-top-pages
ga |>
  arrange(desc(views)) |>
  slice_head(n = 10) |>
  transmute(
    Page = sub("^/RBiocBook/", "", page),
    Views = views,
    Readers = users,
    `Avg. engagement (s)` = round(engagement)
  ) |>
  knitr::kable(caption = "Ten most-viewed pages.")
```

The combination of recurring classroom adoption (~100 trainees across four offerings)
and sustained, deep online engagement (7,500+ views across 60 chapters, ~876 landing-page
readers, multi-minute per-chapter engagement) is the basis for the *Evidence of use*
section of `paper.md`.
